---
title: "Exercises-pr-solutions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

Exercise 1.1 When tossing a fair coin a) what is the probability of heads? b) what is the probability of tails?
a) 0.5
b) 0.5

Exercise 1.2 When tossing a fair six-sided die a) what is the probability of getting 6? b) what is the probability of an even number? c) what is the probability of getting 3 or more? d) what is the expected value of dots on the die?
```{r}
# a)
1/6

# b) 
length(c(2,4,6))/length(c(1,2,3,4,5,6))

# c) 
length(c(3,4,5,6))/length(c(1,2,3,4,5,6))

# d) 
mean(c(1,2,3,4,5,6))

```
Exercise 1.3 In a single coin toss the probability of heads is 0.5.
In 20 coin tosses,
a) what is the probability of exactly 15 heads?
b) what is the probability of less than 7 heads?
c) What is the most probable number of heads?
d) what is the probability of 5 tails or less?
e) what is the probability of 2 heads or less?
```{r}
Nheads <- replicate(10000, {
  coins <- sample(c("H", "T"), size=20, replace=TRUE)
  sum(coins == "H")
})

# a) 
round(sum(Nheads == 15)/length(Nheads), 3)

# b) 
round(sum(Nheads < 7)/length(Nheads), 3)

# c) 
mean(Nheads)

# d) 
Nheads <- replicate(10000, {
  coins <- sample(c("H", "T"), size=20, replace=TRUE)
  sum(coins == "T")
})
round(sum(Nheads <= 5)/length(Nheads), 3)

# e) 
sum(Nheads <= 2)/length(Nheads)

```
Exercise 1.4 When rolling 10 six-sided dice, study the number of sixes.
a) Define the random variable of interest
b) What are the possible outcomes?
c) Using simulation, estimate the probability mass function
```{r}

Nheads <- replicate(10000, {
  dice <- sample(c(1,2,3,4,5,6), size=10, replace=TRUE)
  sum(dice == "6")
})



```


d) what is the probability to get at least 5 sixes?
```{r}

Nheads <- replicate(10000, {
  dice <- sample(c(1,2,3,4,5,6), size=10, replace=TRUE)
  sum(dice == "6")
})

sum(Nheads >= 5) / length(Nheads)

```

e) Which is the most likely number of sixes?
```{r}

# you can read it from histogram, here 1
hist(Nheads)

```

f) What is the probability to get exactly 2 sixes?
```{r}
sum(Nheads == 6) / length(Nheads)
```

g) On average how many sixes do you get when throwing ten dice?
```{r}
mean(Nheads)
```

Exercise 1.5 A deck of cards consists of 52 cards; 13 diamonds, 13 spades, 13 hearts and 13 clubs. When five cards are randomly selected (a poker hand), what is the probability of getting all hearts?
```{r}
Nhearts <- replicate(10000, {
  cards <- sample(c(rep("D", 13), rep("S", 13), rep("H", 13), rep("C", 13)), size=5, replace=FALSE)
  sum(cards == "D")
})
sum(Nhearts == 5)/length(Nheads)

```

Exercise 1.6 30% of a large population is allergic to pollen. If you randomly select 3 people to participate in your study, what is the probability than none of them will be allergic to pollen?
```{r}

Nhealthy<- replicate(10000, {
  people <- sample(c(rep("H", 7), rep("A", 3)), size=3, replace=TRUE)
  sum(people == "H")
})
sum(Nhealthy == 3)/length(Nhealthy)


# with binom()
# success: being allergic
# failure: being healthy 
# 0 successes in 3 attempts with probability success 3/10
pbinom(0, 3, prob  = 3/10)

```

Exercise 1.7 In a class of 20 students, 6 are allergic to pollen. If you randomly select 3 of the students to participate in your study, what is the probability than none of them will be allergic to pollen?

```{r}

Nhealthy<- replicate(10000, {
  people <- sample(c(rep("H", 14), rep("A", 6)), size=3, replace=FALSE)
  sum(people == "H")
})
sum(Nhealthy == 3)/length(Nhealthy)

# with binom()
# success: being allergic
# failure: being healthy 
# 0 successes in 3 attempts with probability success 3/10
pbinom(0, 3, prob  = 3/10)




```
Exercise 1.8 Of the 200 persons working at a company, 60 are allergic to pollen. If you randomly select 3 people to participate in your study, what is the probability that none of them are allergic to pollen?
```{r}
Nhealthy<- replicate(10000, {
  people <- sample(c(rep("H", 140), rep("A", 60)), size=3, replace=FALSE)
  sum(people == "H")
})
sum(Nhealthy == 3)/length(Nhealthy)

```

Exercise 1.9 Compare your results in 1.6, 1.7 and 1.8. Did you get the same results? Why/why not?
In 1.6 we sample with replacement, in 1.7 and 1.8 we sample without replacement (we cannot pick the same person twice from a class or a company)
```{r}

pbinom()

```


Exercise 1.10 Do exercises 1.6, 1.7 and 1.8 again, but using parametric distributions. Compare your results.
1.6 
```{r}



```


1.7 
```{r}
phyper(20, 200, 20000-200, 1000, lower.tail = FALSE)
#k = 20
# 
```
In a class of 20 students, 6 are allergic to pollen. If you randomly select 3 of the students to participate in your study, what is the probability than none of them will be allergic to pollen?
