---
title: "Daily challenge III"
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---


1. Given a linear model $y_i=\alpha + \beta x_i + \epsilon_i$ what terminology is used for $y$ and $x$?
- $y$ is called outcome, response and independent variable
- $x$ is called exposure, predictor or dependent variable
- $y$ is called outcome, response, dependent variable
- $x$ is called outcome, response, dependent variable


2. Given a linear model $Y_i=\beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \epsilon_i$ what is the design matrix ?

a) $\begin{bmatrix} 1 & x_1  \\ 1 & x_2  \\ \vdots & \vdots \\ 1 & x_{n} \end{bmatrix}$ 

b) $\begin{bmatrix} 1 & x_1 & x_2  \\ 1 & x_2 & x_2 \\ \vdots & \vdots & \vdots \\1 & x_n & x_{n}\end{bmatrix}$ 

b) $\begin{bmatrix} 1 & x_1 & x_2^2  \\ 1 & x_2 & x_2^2 \\ \vdots & \vdots & \vdots \\1 & x_n & x_{n}^2\end{bmatrix}$ 

d) $\begin{bmatrix} y_1  \\ y_2  \\ \vdots  \\ y_{n} \end{bmatrix}$ 


3. Given output of lm() regression  what is true based on the slope coefficient?
```{r, echo=F}
weight <- c(58, 70, 74, 63.5, 62.0, 70.5, 71.0, 66.0, 80) # body weight (kg)
plasma <- c(2.75, 2.86, 3.37, 2.76, 2.62, 3.49, 3.05, 3.12, 4) # plasma volume (liters)
gender <- c("Female", "Male", "Female", "Female", "Male", "Male", "Female", "Male", "Male")

#summary(lm(plasma ~ weight))
```

```{r}
# Call:
# lm(formula = plasma ~ weight)
# 
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -0.34908 -0.12950 -0.06887  0.21644  0.25220 
# 
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)   
# (Intercept) -0.81226    0.84364  -0.963  0.36772   
# weight       0.05745    0.01229   _____  ______ **
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.2328 on 7 degrees of freedom
# Multiple R-squared:  0.7573,	Adjusted R-squared:  0.7226 
# F-statistic: 21.84 on 1 and 7 DF,  p-value: 0.002279
```

a) for every unit increase in weight plasma increases by 0.05
b) for every unit increase in weight plasma decreases by 0.81
c) for every unit increase in weight plasma increases 0.012
d) for every 100 unit increase in weight plasma increases by 0.05

4. Given the output of the linear regression to model plasma volume given weight we would like to check whether we can reject the null hypothesis that $H_0: \beta = 0$ if favour of the alternative $H_0: \beta \ne 0$ . What is t-value and the associated p-value? Is there enough evidence to reject the null hypothesis?

a) t-value = 4.673, p-value = 0.01229: there is enough evidence reject the null hypothesis and conclude that there is a significant association between weight and plasma volume
b) t-value = 4.673, p-value = 0.00228: there is enough evidence reject the null hypothesis and conclude that there is a significant association between weight and plasma volume
c) t-value = 4.673, p-value = 0.00228: there is not enough evidence reject the null hypothesis and conclude that there is a significant association between weight and plasma volume
d) t-value = 0.05745, p-value = 0.01229: there is enough evidence reject the null hypothesis and conclude that there is a significant association between weight and plasma volume

5. Given the output of lm() with categorical gender variable what is the expected value of plasma for a man that weights 80 kg? 
```{r, echo=F}
weight <- c(58, 70, 74, 63.5, 62.0, 70.5, 71.0, 66.0, 80) # body weight (kg)
plasma <- c(2.75, 2.86, 3.37, 2.76, 2.62, 3.49, 3.05, 3.12, 4) # plasma volume (liters)
gender <- c("Female", "Male", "Female", "Female", "Male", "Male", "Female", "Male", "Male")

summary(lm(plasma ~ weight + gender))
```

a) 3.8
b) 3.74
c) 5.27
d) 5.33


6. In simple linear regression model, what is $R^2$ given the response and fitted response values
```{r}
plasma <- c(2.75, 2.86, 3.37, 2.76, 2.62)
plasma.hat <- c(2.60, 3.04, 3.19, 2.78, 2.74)

```

a) 0.69
b) 2.87
c) 0.10
d) 0.34

```{r, echo = F, include = F}
weight <- c(58, 70, 74, 63, 62) # body weight (kg)
plasma <- c(2.75, 2.86, 3.37, 2.76, 2.62) # plasma volume (liters)
m1 <- lm(plasma ~ weight)
summary(m1)

yhat <- round(m1$fitted.values, 2)

RSS <- sum((plasma - m1$fitted.values)^2)
TSS <- sum((plasma- mean(plasma))^2)
R2 = 1-RSS/TSS
round(R2, 2)

```


7. Which of the below residuals plot does not check any assumptions of the linear model?
a) histogram of residuals
b) residuals vs. fitted values
c) residuals vs. model coefficients
d) residuals vs. explanatory variable

8. Given the logistic regression output of Healthy/Unhealthy (1/0) vs. plasma and weight model by much are odds of being healthy multiplied with unit increase of plasma volume? 
```{r, echo=F}

weight <- c(58, 70, 74, 63.5, 62.0, 70.5, 71.0, 66.0, 80, 81) # body weight (kg)
plasma <- c(2.75, 2.86, 3.37, 2.76, 2.62, 3.49, 3.05, 3.12, 4, 4) # plasma volume (liters)
gender <- c("Female", "Male", "Female", "Female", "Male", "Male", "Female", "Male", "Male", "Male")
gender <- as.factor(gender)
healthy <- as.factor(c("H", "Male", "H", "H", "Male", "Male", "H", "Male", "Male", "Male"))

m1 <- glm(healthy ~ plasma + weight, family = binomial(link="logit"))
summary(m1)
```

a) 1.00
b) 6.79
c) 31.26
d) 1.75


----

## Solutions to selected exercies: 

Solutions to quiz: linear models (selected problems)

We calculate t-value as ￼ 
and we can obtain p-value in R as 
2*pt(0.05745/0.01229, df=7, lower=F).
We have 7 degrees of freedom, as we have 9 observations minus two model parameters, that is intercept and slope

round(-0.76557 + 80*0.05626 + 0.06251, 2)

> RSS <- sum((plasma - plasma.hat)^2) 
> TSS <- sum((plasma - mean(plasma))^2) 
> R2 = 1 - RSS/TSS 
> R2 
[1] 0.6988911
Double check that you cannot calculate R2 in R given the data 

exp(1.9158930) = 6.79
 
