---
title: "Survival Analysis"
format: 
  revealjs:
    slide-number: true
    view-distance: 10
    theme: [default]
    mermaid: 
      theme: forest
    chalkboard: 
      buttons: true
  html:
    code-fold: false
editor_options: 
  chunk_output_type: console
execute:
  cache: true
---

```{r}
#| message: false
#| warning: false

# load libraries
library(tidyverse)
library(magrittr)
library(kableExtra)
library(ggplot2)
library(rmarkdown)
library(ggbeeswarm)
library(gridExtra)
library(ggmosaic)
library(scales)
library(ggthemes)
```

## Survival analysis 🏕️ {.smaller} 

```{r}
#| include: false
#| message: false
#| warning: false

library(ISLR2)
library(janitor)
library(skimr)
library(tidyverse)
library(RColorBrewer)
library(survival)
library(survminer)

mycols <- brewer.pal(6, "Set1")
```

Statistical methods for **time-to-event data**

- Originally applied to time until death in mortality studies
- Now widely used in many fields:
  - Medicine: time to relapse
  - Marketing: time to next purchase
  - Engineering: time to failure
  - Criminology: time to reoffend

. . .

**Key feature:** the outcome is time until an event, often with **censoring**.

. . .

Survival analysis is not limited to calendar time, it can also be applied to for example:

- Number of treatment sessions until recovery
- Number of attempts until passing an exam

## Survival and censoring times {.smaller}

Let's imagine:

  - a mortality study of terminally ill patients receiving a medical treatment. 
  - a follow-up period of **24 months** 
- the outcome of interest is **time until death**
- for each patient, we record the **time until death or censoring**

. . .

For each patient, we suppose there is: 

- a true **survival time** $T$
  - representing the time at which the event of interest occurs (here death)
- a **censoring time** $C$
  - representing the time at which the patient is lost to follow-up or the study ends.

:::{.notes}
- a true **survival time** $T$
  - representing the time at which the event of interest occurs (here death)
- a **censoring time** $C$
  - representing the time at which the patient is lost to follow-up or the study ends.
:::


. . .

We observe the **random variable** $$Y = \min (T, C)$$


- If $T \leq C$, we observe the event time $T$
- If $T > C$, we observe the censoring time $C$

and an event indicator

$$
    \delta =
    \left\{
        \begin{array}{cc}
                1 & \mathrm{if\ } T \le C \\
                0 & \mathrm{if\ } T>C \\
        \end{array}
    \right.
$$

## Survival and censoring times {.smaller}

```{r}
#| label: fig-censoring
#| fig-cap: Example of right-censored data in a mortality study lasting 24 months. The horizontal lines represent the time to event or censoring for each patient. The points indicate whether the event occurred (event) or not (censored).
#| fig-cap-pos: margin

# Example data for 10 patients
data <- data.frame(
  patient = factor(1:10),     # Patient IDs
  time = c(23, 24, 9, 7, 14, 16, 19, 20, 24, 24),  # Time to event or censoring
  status = c(1, 0, 1, 0, 1, 1, 0, 1, 0, 1)       # Event status (1 = event, 0 = censored)
)

# Plot the data with horizontal lines
ggplot(data, aes(x = time, y = patient)) +
  geom_segment(aes(x = 0, xend = time, y = patient, yend = patient), color = mycols[2], size = 1.5) +  #
  geom_point(aes(shape = factor(status)), size = 3) +  
  geom_vline(xintercept = 24, linetype = "dashed") +
  scale_shape_manual(values = c(1, 16),               
                     labels = c("Censored", "Event")) +
  scale_y_discrete(limits = rev) +
  labs(x = "Time (months)", y = "Patient", shape = "Status") + 
  theme_minimal() +
  theme(legend.position = "top")
```

. . .

::: incremental
- For patient 1: $(y_1 = t_1 = 23, \delta_1 = 1)$
- For patient 2: $(y_2 = c_2 = 24, \delta_2 = 0)$
- For patient 4: $(y_4 = c_4 = 7, \delta_4 = 0)$, a patient lost to follow-up
:::

:::{.notes}

- for P1 we get $(y_1= t_1, \space \delta_1 = 1)$ since we observe event (death) as ca. 23 months, before the end of the follow-up at 24 months. 
- for P2 we get $(y_2= c_2, \space \delta_2 = 0)$ since the patient is still alive at the end of the study at 24 months, meaning that the patient has survived at least 24 months (censored)
- for P4, we get $(y_4= c_4, \space \delta_4 = 0)$, another censored data point. The study is not over yet, the patient has not died but is **lost to follow-up**, e.g. due to withdrawing from the study

:::


## Independent Censoring {.smaller}

To analyze survival data correctly, we assume **independent censoring**:
- the reason for censoring is unrelated to the risk of the event
- censored individuals are assumed to have the same event risk as those still under observation
  
. . .  

<br>

**Example**

- ✅ Patients leave a cancer study for unrelated reasons, e.g. moving away
- ❌ Patients drop out because they are very sick

## Left and interval censoring {.smaller}

```{r}
#| label: fig-left-interval
#| fig-cap: "Example of left and interval censoring in an age study. Left-censored patients were diagnosed before the study, and interval-censored patients were diagnosed between visits."

# Create a data frame for 8 patients
# Left-censored patients were diagnosed before the study, and interval-censored patients were diagnosed between visits
data <- data.frame(
  patient_id = 1:8,
  start_age = c(NA, 30, 40, 35, NA, 25, 45, 50),   # Start of observation or known age for interval censoring (NA for left-censored)
  end_age = c(25, 35, 45, 50, 20, 40, 55, 60),     # Age of diagnosis or end of observation
  type = c("Left Censored", "Fully Observed", "Fully Observed", "Fully Observed", 
                     "Left Censored", "Interval Censored", "Interval Censored", "Fully Observed")
)

# Plot the data
ggplot(data, aes(x = end_age, y = factor(patient_id), color = type)) +
  # Add points for observed ages (diagnosis age or censoring)
  geom_point(size = 4) +

  # For interval-censored patients, add a horizontal line to show the interval
  geom_segment(data = subset(data, type == "Interval Censored"), aes(x = start_age, xend = end_age, y = factor(patient_id), yend = factor(patient_id)),
               color = mycols[2], linetype = "dashed") +
  
  # For left-censored patients, add an arrow showing the uncertainty before diagnosis
  geom_segment(data = subset(data, type == "Left Censored"), aes(x = 0, xend = end_age, y = factor(patient_id), yend = factor(patient_id)),
               color = mycols[1], linetype = "dotted", arrow = arrow(length = unit(0.2, "cm"))) +
  
  # Customize the plot labels and appearance
  labs(x = "Age", y = "Patient ID") +
  scale_color_manual(values = c("Left Censored" = mycols[1], "Fully Observed" = "black", "Interval Censored" = mycols[2])) +
  theme_minimal() + 
  theme(legend.position = "top", legend.title = element_blank())

```

In this lecture, we focus on **right-censored** data.

## Kaplan-Meier estimator {.smaller}

The **survival function** is

$$S(t) = P(T > t)$$

the probability of surviving past time $t$.

. . .

The **Kaplan-Meier (KM) estimator** is
$$\hat{S}(t) = \prod_{t_j \le t} \left( 1 - \frac{d_j}{n_j} \right)$$ {#eq-KM}

where: 

- $d_j$: number of events at time $t_j$
- $n_j$: number at risk just before time $t_j$

. . .

The Kaplan-Meier estimator is also called the **product-limit estimator**.


## Kaplan-Meier estimator {.smaller}

*Tooth filling example: complete follow-up*

Suppose 10 patients receive a tooth filling, and we record the time until failure.

::::{.columns}

:::{.column width="40%"}

```{r}
#| label: tooth-filling
data_tooth <- data.frame(
  patientID = 1:10,
  time = c(10, 8, 11,  5, 4,  3,  7,  6, 12,  8), 
  status = 1
)

data_tooth |> gt() |> tab_options(table.font.size = px(16), data_row.padding = px(2))
```

:::

::: {.column width="10%"}
:::{.fragment}

<br><br><br><br><br>
$\Longrightarrow$

sort

:::
:::

::: {.column width="40%"}
:::{.fragment}

```{r}
#| label: tooth-filling-sorted
data_tooth |> arrange(time) |> gt() |> tab_options(table.font.size = px(16), data_row.padding = px(2))
```

:::
:::

Goal:

- estimate the probability that a filling remains intact over time
- first assume **complete follow-up**, i.e. no censoring

::::


## Kaplan-Meier estimator {.smaller}

*Tooth filling example: building the Kaplan-Meier table*

From the sorted data, we identify the event times and for each time compute:

:::: {.columns}

::: {.column width="48%"}

- the number at risk, $n_j$
- the number of events, $d_j$
- the conditional survival, $1-\frac{d_j}{n_j}$
- $$S(t) = \prod_{t_j \le t} \left( 1 - \frac{d_j}{n_j} \right)$$

```{r}
#| label: tbltooth1
#| echo: false
#| results: asis
#| tbl-align: "left"
#| tbl-cap: "Sorted data"
data_tooth |>
  arrange(time, patientID) |>
  gt() |>
  cols_label(
    patientID = "Patient",
    time = "Time",
    status = "Status"
  ) |>
  tab_options(
    table.font.size = px(15),
    data_row.padding = px(2),
    table.margin.left = px(0)
  )
```
:::

::: {.column width="48%"}

<div style="font-size: 70%; line-height: 1.1;">

:::{.r-stack}

::: {.fragment .current-visible}

| Time | At risk $n_j$ | Events $d_j$ | Conditional survival | Cumulative survival |
|---|---:|---:|---:|---:|
| 3  | 10 | 1 | $1-\frac{1}{10} = 0.9$ | $0.9$ |
| 4  |    |   |   |   |
| 5  |    |   |   |   |
| 6  |    |   |   |   |
| 7  |    |   |   |   |
| 8  |    |   |   |   |
| 10 |    |   |   |   |
| 11 |    |   |   |   |
| 12 |    |   |   |   |

:::

::: {.fragment .current-visible}

| Time | At risk $n_j$ | Events $d_j$ | Conditional survival | Cumulative survival |
|---|---:|---:|---:|---:|
| 3  | 10 | 1 | $0.9$ | $0.9$ |
| 4  | 9  | 1 | $1-\frac{1}{9} = 0.888$ | $0.9 \times 0.888$ |
| 5  |    |   |   |   |
| 6  |    |   |   |   |
| 7  |    |   |   |   |
| 8  |    |   |   |   |
| 10 |    |   |   |   |
| 11 |    |   |   |   |
| 12 |    |   |   |   |
:::

::: {.fragment .current-visible}


| Time | At risk $n_j$ | Events $d_j$ | Conditional survival | Cumulative survival |
|---|---:|---:|---:|---:|
| 3  | 10 | 1 | $0.9$ | $0.9$ |
| 4  | 9  | 1 | $0.888$ | $0.8$ |
| 5  | 8  | 1 | $1-\frac{1}{8 = 0.875}$ | $0.8 \times 0.875$ |
| 6  |    |   |   |   |
| 7  |    |   |   |   |
| 8  |    |   |   |   |
| 10 |    |   |   |   |
| 11 |    |   |   |   |
| 12 |    |   |   |   |
:::

::: {.fragment .current-visible}

| Time | At risk $n_j$ | Events $d_j$ | Conditional survival | Cumulative survival |
|---|---:|---:|---:|---:|
| 3  | 10 | 1 | $0.9$   | $0.9$ |
| 4  | 9  | 1 | $0.888$ | $0.8$ |
| 5  | 8  | 1 | $0.875$ | $0.7$ |
| 6  | 7  | 1 | $0.857$ | $0.6$ |
| 7  | 6  | 1 | $0.833$ | $0.5$ |
| 8  | 5  | 2 | $0.6$   | $0.3$ |
| 10 | 3  | 1 | $0.667$ | $0.2$ |
| 11 | 2  | 1 | $0.5$   | $0.1$ |
| 12 | 1  | 1 | $0$     | $0$ |

:::

:::

</div>

:::

::::


::: {.notes}
- At 3 months: 10 patients are at risk, and 1 failure occurs, so the survival probability for this period is 0.9. 
- For 4 months: 9 patients are at risk (since one patient failed at 3 months), and the survival probability is now cumulative (0.9 from the first interval multiplied by 0.888 from this interval).
- At time 8 months: two failures occur and the survival probability drops more sharply.
- At time 12 months: the survival probability reaches 0, indicating that all patients experienced the event (tooth filling failure) by the end of the study.
:::


## Kaplan-Meier estimator {.smaller}

```{r}
#| label: fig-tooth-filling-1
#| fig-height: 6
#| fig-cap: Kaplan-Meier curve for tooth filling data with complete follow-up.
fit.surv <- survfit(Surv(time, status) ~ 1, data = data_tooth)
plot(fit.surv, conf.int=FALSE,  xlab = "Months", ylab = "Estimated Probability of Survival", las = 1)
```

## Kaplan-Meier estimator {.smaller}

*Tooth filling example: what changes with censoring?*

Now suppose two patients are censored.

- censored patients remain in the **risk set** until censoring
- they do **not** count as events
- censoring affects future risk sets, but does not change survival at the censoring time

```{r}
#| label: tbltooth2
#| echo: false
#| results: asis
#| tbl-align: "left"
#| tbl-cap: "Sorted data with censoring"
data_tooth_cens <- data.frame(
  patientID = 1:10,
  time = c(10, 8, 11, 5, 4, 3, 7, 6, 12, 8),
  status = c(0, 1, 1, 0, 1, 1, 1, 1, 1, 1)
)

data_tooth_cens |>
  arrange(time, patientID) |>
  gt() |>
  cols_label(
    patientID = "Patient",
    time = "Time",
    status = "Status"
  ) |>
  tab_options(
    table.font.size = px(15),
    data_row.padding = px(2),
    table.margin.left = px(0)
  )
```

## Kaplan-Meier estimator {.smaller}

*Tooth filling example: what changes with censoring?*

Now suppose two patients are censored.

:::: {.columns}

::: {.column width="48%"}

```{r}
#| label: tbltooth2b
#| echo: false
#| results: asis
#| tbl-align: "left"
#| tbl-cap: "Sorted data with censoring"
data_tooth_cens |>
  arrange(time, patientID) |>
  gt() |>
  cols_label(
    patientID = "Patient",
    time = "Time",
    status = "Status"
  ) |>
  tab_options(
    table.font.size = px(15),
    data_row.padding = px(2),
    table.margin.left = px(0)
  )
```

:::

::: {.column width="48%"}

<div style="font-size: 70%; line-height: 1.1;">

:::{.r-stack}

::: {.fragment .current-visible}

| Time | At risk $n_j$ | Events $d_j$ | Censored | Conditional survival | Cumulative survival |
|---|---:|---:|---:|---:|---:|
| 3  | 10 | 1 | 0 | $1-\frac{1}{10} = 0.9$ | $0.9$ |
| 4  |    |   |   |   |   |
| 5  |    |   |   |   |   |
| 6  |    |   |   |   |   |
| 7  |    |   |   |   |   |
| 8  |    |   |   |   |   |
| 10 |    |   |   |   |   |
| 11 |    |   |   |   |   |
| 12 |    |   |   |   |   |

:::

::: {.fragment .current-visible}

| Time | At risk $n_j$ | Events $d_j$ | Censored | Conditional survival | Cumulative survival |
|---|---:|---:|---:|---:|---:|
| 3  | 10 | 1 | 0 | $0.9$ | $0.9$ |
| 4  | 9  | 1 | 0 | $1-\frac{1}{9} = 0.888$ | $0.9 \times 0.888$ |
| 5  |    |   |   |   |   |
| 6  |    |   |   |   |   |
| 7  |    |   |   |   |   |
| 8  |    |   |   |   |   |
| 10 |    |   |   |   |   |
| 11 |    |   |   |   |   |
| 12 |    |   |   |   |   |
:::

::: {.fragment .current-visible}

| Time | At risk $n_j$ | Events $d_j$ | Censored | Conditional survival | Cumulative survival |
|---|---:|---:|---:|---:|---:|
| 3  | 10 | 1 | 0 | $0.9$ | $0.9$ |
| 4  | 9  | 1 | 0 | $0.888$ | $0.8$ |
| 5  | 8  | 0 | 1 |  | $0.8$ |
| 6  |    |   |   |   |   |
| 7  |    |   |   |   |   |
| 8  |    |   |   |   |   |
| 10 |    |   |   |   |   |
| 11 |    |   |   |   |   |
| 12 |    |   |   |   |   |
:::

::: {.fragment .current-visible}

| Time | At risk $n_j$ | Events $d_j$ | Censored | Conditional survival | Cumulative survival |
|---|---:|---:|---:|---:|---:|
| 3  | 10 | 1 | 0 | $0.9$   | $0.9$ |
| 4  | 9  | 1 | 0 | $0.888$ | $0.8$ |
| 5  | 8  | 0 | 1 |  | $0.8$ |
| 6  | 7  | 1 | 0 | $0.857$ | $0.686$ |
| 7  | 6  | 1 | 0 | $0.833$ | $0.571$ |
| 8  | 5  | 2 | 0 | $0.6$   | $0.343$ |
| 10 | 3  | 0 | 1 |  | $0.343$ |
| 11 | 2  | 1 | 0 | $0.5$   | $0.171$ |
| 12 | 1  | 1 | 0 | $0$     | $0$ |

:::

:::

</div>

:::

::::


## Kaplan-Meier estimator {.smaller}

The Kaplan-Meier curve shows the cumulative survival probabilities. 

```{r}
#| label: fig-tooth-filling-2
#| fig-height: 6
#| fig.cap: Kaplan-Meier survival curve for tooth filling data. The curve shows the estimated probability of survival (filling intact) over time.

data_tooth[1, "status"] <- 0
data_tooth[4, "status"] <- 0
fit.surv <- survfit(Surv(time, status) ~ 1, data = data_tooth)
plot(fit.surv, mark.time=TRUE, conf.int=FALSE,  xlab = "Months", ylab = "Estimated Probabiliy of Survival", las = 1)

```

```{r KM2groups}
# Load required libraries
library(survival)
library(survminer)

# Simulate survival data for two groups
set.seed(123)

# Group 1: Worse prognosis (shorter survival times)
group1 <- data.frame(
  time = rexp(50, rate = 0.2),   # higher event rate → shorter survival
  status = rep(1, 50),
  group = "High Event Rate"
)

# Group 2: Better prognosis (longer survival times)
group2 <- data.frame(
  time = rexp(50, rate = 0.05),  # lower event rate → longer survival
  status = rep(1, 50),
  group = "Low Event Rate"
)

# Combine the data
data_km <- rbind(group1, group2)

# Fit KM survival curves
fit2groups <- survfit(Surv(time, status) ~ group, data = data_km)

# Plot the curves
pl2groups <- ggsurvplot(fit2groups,
           data = data_km,
           xlab = "Time",
           ylab = "Survival Probability",
           risk.table = FALSE,
           palette = mycols[1:2],
           font.main = c(20, "bold"),
           font.x = c(18),
           font.y = c(18),
           font.tickslab = c(18),
           font.legend = c(18),
           font.risk.table = c(22),
           legend.title = "Group",
           legend.labs = c("High Event Rate", "Low Event Rate"))
```


## Kaplan-Meier estimator {.smaller}

*Reading a survival probability from the curve*


:::: {.columns}

::: {.column width="50%"}

At a specific time point, we can estimate the survival probability by reading the value of the curve.

Example: what is the probability that a tooth filling lasts at least **10 months**?

:::

::: {.column width="50%"}

:::{.r-stack}

:::{.fragment}

```{r}
#| label: KM-01
#| fig-width: 5
#| fig.cap: Kaplan-Meier survival curve for tooth filling data.

plot(fit.surv, conf.int=FALSE,  xlab = "Months", ylab = "Estimated Probabiliy of Survival", las = 1)
#abline(v = 10, h = 0.343, col = "red", lty = 2)

```

:::

:::{.fragment}

```{r}
#| label: KM-02
#| fig-width: 5
#| fig.cap: Kaplan-Meier survival curve for tooth filling data.

plot(fit.surv, conf.int=FALSE,  xlab = "Months", ylab = "Estimated Probabiliy of Survival", las = 1)
abline(v = 10, h = 0.343, col = "red", lty = 2)

```

:::
:::

:::

::::

```{r}
#| label: figtooth-filling-20-part
##| fig.cap: Kaplan-Meier survival curves for a new extended study of tooth filling data, including 20 participants (10 men and 10 women).
##| fig-height: 7

# Create data frame for survival data
data_tooth_ext <- data.frame(
  time = c(3, 5, 6, 7, 9, 10, 12, 15, 18, 20, 
           4, 6, 7, 9, 11, 13, 14, 16, 18, 22),
  status = c(1, 1, 1, 1, 1, 0, 1, 1, 0, 1, 
               1, 1, 0, 1, 1, 1, 1, 0, 1, 1),
  gender = c(rep("M", 10), rep("F", 10))
)

# Create a survival object and
surv_object <- Surv(time = data_tooth_ext$time, event = data_tooth_ext$status)

# Fit Kaplan-Meier curves for gender
fitg <- survfit(surv_object ~ gender, data = data_tooth_ext)

# Plot Kaplan-Meier survival curves
plgender <- ggsurvplot(fitg, 
           data = data_tooth_ext,
           pval = FALSE,             # Show p-value for difference in survival
           conf.int = FALSE,         # Add confidence interval
           risk.table = FALSE,       # Show risk table at the bottom
           ggtheme = theme_minimal(), # Use a minimal theme
           xlab = "Time (Months)",
           ylab = "Survival Probability", 
           palette = mycols[1:2], 
           font.main = c(20, "bold"),
           font.x = c(18),
           font.y = c(18),
           font.tickslab = c(18),
           font.legend = c(18),
           font.risk.table = c(22),
           legend.title = "Gender",
           legend.labs=c("Female", "Male"))
```


## Comparison of survival {.smaller}

<br>

When comparing survival between groups, we often start by plotting **Kaplan-Meier curves**.

- These curves show how the **survival probability** changes over time in each group.
- If the curves are close together, the groups may have similar survival experiences.
- If the curves are clearly separated, this suggests a difference in survival.
- If the curves cross, the difference may change over time and interpretation becomes more difficult.

```{r KMcompare}
#| fig-width: 20
#| fig-height: 5
#| fig-align: "center"
arrange_ggsurvplots(list(plgender, pl2groups), ncol=2, nrow=1, plot=TRUE)
```

## Comparisons of hazards {.smaller}


To compare groups more formally, we often focus on the **hazard**.

- The **hazard** is the instantaneous event rate at time $t$ among individuals still at risk.
- Instead of comparing survival probabilities directly, we compare how quickly events occur in the groups over follow-up.

. . .

<br>

A common assumption is the **proportional hazards assumption**:

$$
\frac{h_1(t)}{h_0(t)} = \text{constant}
$$

where $h_1(t)$ and $h_0(t)$ are the hazards in the two groups at time $t$.

If this assumption is reasonable, the groups differ by a constant **hazard ratio** over time.

## Mantel-Cox (log-rank test) {.smaller}

The **log-rank test** compares survival between two or more groups.

- Null hypothesis: no difference in survival between groups
- At each event time, compare the **observed** and **expected** number of events
- Combine these differences over all event times
- Test statistic follows a **chi-square distribution**

## Mantel-Cox (log-rank test) {.smaller}

:::: {.columns}

::: {.column width="55%"}

At a given event time:

|            | Died | Survived | Total at risk |
|------------|-----:|---------:|--------------:|
| Group A    | $d_{jA}$ | $r_{jA}$ | $n_{jA}$ |
| Group B    | $d_{jB}$ | $r_{jB}$ | $n_{jB}$ |
| Total      | $d_j$    | $r_j$    | $n_j$    |

:::

::: {.column width="45%"}

:::{.fragment}
Observed number of events at time $t_j$:

$$
O_{jA} = d_{jA}, \qquad O_{jB} = d_{jB}
$$
:::

:::{.fragment}
Expected number of events at time $t_j$:

$$
E_{jA} = \frac{n_{jA} d_j}{n_j}, \qquad
E_{jB} = \frac{n_{jB} d_j}{n_j}
$$
:::

:::{.fragment}
Large differences between **observed** and **expected** numbers of events give evidence against the null hypothesis. 
:::

:::
::::

## Mantel-Cox (log-rank test) {.smaller}

The Mantel-Cox (log-rank) statistic combines these observed-minus-expected differences over all event times:

$$
\chi^2_{MC} = \sum_{j} \left( \frac{(O_{jA} - E_{jA})^2}{E_{jA}} + \frac{(O_{jB} - E_{jB})^2}{E_{jB}} \right)
$$

. . .

For two groups, the statistic is approximately $\chi^2$ distributed with 1 degree of freedom.

# Regression with survival data


## Cox proportional hazards model {.smaller}

*Studying the relationship between survival time and predictors*

The **Cox proportional hazards model** is used to study how predictor variables affect the hazard of an event.

- allows adjustment for several covariates at the same time
- estimates **hazard ratios**
- does not require specifying the baseline hazard function

. . .

The model is

$$
h(t|\boldsymbol{x}) = h_0(t)\exp(\boldsymbol{\beta}^T\boldsymbol{x})
$$

where: 

- $h(t|\boldsymbol{x})$: hazard at time $t$ for a subject with covariates $\boldsymbol{x}$
- $h_0(t)$: baseline hazard
- $\boldsymbol{\beta}^T\boldsymbol{x} = \beta_1x_1 + \beta_2x_2 + \cdots + \beta_px_p$: effect of covariates

## Cox proportional hazards model {.smaller}

*Brain cancer example*

The `BrainCancer` data set from the `ISLR2` package contains survival times for patients with primary brain tumors.

Variables included in the study are:

- `sex`: male or female
- `diagnosis`: meningioma, LG glioma, HG glioma, or other
- `loc`: tumor location, infratentorial or supratentorial
- `gtv`: gross tumor volume $cm^3$
- `ki`: Karnofsky index
- `stereo`: stereotactic radiosurgery (SRS) or fractionated stereotactic (SRT) radiotherapy

## Cox proportional hazards model {.smaller}

```{r}
#| echo: true
# Cox model with multiple predictors
fit.cox_multi <- coxph(Surv(time, status) ~ sex + diagnosis + ki, data = BrainCancer)
summary(fit.cox_multi)
```



## Cox proportional hazards model {.smaller}

*Interpreting coefficients*

Based on the fitted Cox model:

- HR for a male patient is about $e^{0.11} = 1.11$
- This suggests a slightly higher hazard for men than women, but the p-value is large
- A one-unit increase in Karnofsky index gives HR $e^{-0.06} = 0.94$
- Higher Karnofsky index is associated with **lower hazard**
- For a 10-unit increase in Karnofsky index:
  $$
  e^{-0.06 \times 10} = 0.54
  $$

## Cox proportional hazards model {.smaller}

*Forest plot*

Hazard ratios and their 95% confidence intervals can be visualized in a forest plot.

```{r}
#| label: forest-plot

# make forest plot with ggforest()
ggforest(fit.cox_multi, data = data_brain)
```

## Cox proportional hazards model {.smaller}

*Model output*

<br>

Cox regression output includes several useful tests.

. . . 

<br>

**Likelihood Ratio Test (LRT)**

- Compares the log-likelihood of:
  - A null model (no covariates)
  - A full model (with covariates)
- Tests whether the covariates improves model fit

. . .

<br>

**Wald test**

- Individual Wald tests assess whether each coefficient is significantly different from 0
- A global Wald test assesses whether all coefficients are 0 simultaneously

. . .

<br>

**Score test**

- Based on **expected vs. observed** events
- In simple group-comparison settings, it corresponds to the log-rank test

## Cox proportional hazards model {.smaller}

*Concordance*

The **concordance index** measures how well the model predicts the order of events.

- **C-index = 0.5** → No better than random guessing
- **C-index = 1** → Perfect prediction of event order

. . .

**Example:**

- Concordance = **0.782**
- The model correctly orders event times in 78.2% of comparable pairs

## Cox proportional hazards model {.smaller}

*Time-dependent covariates*

So far, predictors have been fixed over time.

But some predictors may **change during follow-up**, for example:

- blood pressure
- treatment status
- disease severity

. . .

The Cox model can be extended to time-dependent covariates:

$$
h(t|\boldsymbol{x}(t)) = h_0(t)\exp(\boldsymbol{\beta}^T\boldsymbol{x}(t))
$$

In practice, one individual may contribute several rows, one for each interval where the covariates are constant.

## Multiple events: competing risks {.smaller}

A **competing risk** is an event that prevents the event of interest from occurring.

<br>

**Examples:**

- In a study of heart disease death, death from cancer is a competing risk
- In a study of relapse, death from other causes is a competing risk

. . .

**Why it matters:**

- standard Cox models typically focus on one event type at a time
- ignoring competing risks can overestimate the cumulative probability of the event of interest

. . .

Useful tools:

- **Cumulative incidence function (CIF)**  
  estimates the probability of a specific event by time $t$, accounting for competing risks

- **Gray’s test**  
  tests whether CIFs differ between groups

- **Fine and Gray model**  
  models the effect of covariates on the **subdistribution hazard**, and is used when interest lies in prediction or comparison of cumulative incidence


## 💻 Lab {.smaller}


In the lab there are examples of: 

- constructing Kaplan-Meier curves
- comparing survival curves with the log-rank test
- fitting Cox proportional hazards models
- interpreting model output
- handling time-dependent covariates
- dealing with competing risks



