---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Wilcoxon signed rank test I

*for a median*

```{r}
#| message: false
#| warning: false

# load libraries
library(tidyverse)
library(magrittr)
library(kableExtra)
```

Named after Frank Wilcoxon (1892–1945), Wilcoxon signed rank test was one of the first "non-parametric" methods developed. It can be used to:

i) compare the sample median against a hypothetical median (equivalent to one sample *t*-test) 
ii) examine the difference between paired observations (equivalent to paired *t*-test). It does make some assumptions about the data, namely: 

- the random variable X is continuous
- the probability density function of X is symmetric


::: {#exm-wilcoxon-signed}
## Wilcoxon signed rank test (for a median)

Let's imagine we are a part of team analyzing results of a placebo-controlled clinical trial to test the effectiveness of a sleeping drug. We have collected data on 10 patients when they took a sleeping drug and when they took a placebo.

The hours of sleep recorded for each study participant: 
```{r}
# input data
data.sleep <- data.frame(id = 1:10, 
                         drug = c(6.1, 6.0, 8.2, 7.6, 6.5, 5.4, 6.9, 6.7, 7.4, 5.8), 
                         placebo = c(5.2, 7.9, 3.9, 4.7, 5.3, 7.4, 4.2, 6.1, 3.8, 7.3)) 

print(data.sleep)

```

Before we investigate the effect of drug, a senior statistician ask us: 

**"Is the median sleeping time without taking the drug significantly less than the recommended 7 h of sleep?"**

:::


## Define the null and alternative hypothesis under study

$H_0: m = m_0$  the median sleeping time is equal to $m_0$, $m_0 = 7$ h

$H_1 < m_0$  the median sleeping time is less than $m_0$, $m_0 = 7$ h

## Calculate the value of the test statistics

1. we subtract the median from each measurement, $X_i - m_0$
2. we find absolute value of the difference, $|X_i - m_0|$
3. we rank the absolute value of the difference
4. we find the value of $W$, the Wilcoxon signed-rank test statistics as $$W =\displaystyle \sum_{i=1}^{n}Z_iR_i$${#eq-w} where $Z_i$ is an indicator variable such as: 

\begin{equation}
    Z_i =
    \left\{
        \begin{array}{cc}
                0  & \mathrm{if\ } X_i - m_0 < 0 \\
                1  &  otherwise \\
        \end{array}
    \right.
\end{equation}


```{r}
#| label: tbl-wilcoxon
#| tbl-cap: "Demonstrating steps in the calculating W, Wilcoxon signed-rank test statistics on the placebo column: x stands for placebo sleeping hours"
#| tbl-cap-location: margin

m0 <- 7
data.wilcoxon <- data.sleep %>% 
  select(!drug) %>% # remove drug data for now
  rename(x = placebo) %>% # rename placebo column to x so it is easier to type and follow eq
  mutate(`x-m0` = x - 7) %>% # subtract m0
  mutate(`abs(x-m0)` = abs(`x-m0`)) %>% # take absolute value
  mutate(R = rank(`abs(x-m0)`)) %>% # rank
  mutate(Z = ifelse(`x-m0` < 0, 0, 1)) %>% # define indicator variable Z
  mutate(ZR = R*Z) # calculate ranks R times Z

# print the table
data.wilcoxon %>%
  kable() %>% kable_styling(full_width = FALSE)

```

We can now calculate $W$ following equation @eq-w and we get:

```{r}
# sum up the ranks multiplied by Z indicator value
W <- data.wilcoxon$ZR %>% sum()
print(W)
```

## Compare the value to the test statistics to values from known probability distribution

We got $W = 6.5$ and now we need to calculate the *P*-value associated with $W$ to be able to make decision about rejecting the null hypothesis. We refer to a statistical table "Upper and Lower Percentiles of the Wilcoxon Signed Rank Test, W" that can be found online or [here](statstable/Wilcoxon-signed-rank-test.pdf).

For sample size $n=10$ we can see that probability of observing $W \le 3$ or $W \ge 52$ is small, 0.005. Probability of observing $W \le 4$ or $W \ge 51$ is 0.007, still small but slightly larger. While we are getting towards the middle of the distribution the probability of observing $W$ is getting larger and the probability of observing $W \le11$ or $W \ge 44$ is 0.053.  

The *P*-value associated with observing $W=6.5$ is just under $0.019$. Assuming 5% significance level, we have enough evidence to reject the null hypothesis and conclude that the median is significantly less than 7 hours. 

## Obtaining probablity mass function

Where is the statistical table coming from? 

Briefly, Wilcoxon described and showed examples how to calculate both the test statistics $W$ for an example data as well as the distribution of $W$ under the null hypothesis @Wilcoxon1945. We can try to find the distribution of W ourselves for a simple scenario with less, four observation ($n=4$)

```{r}
#| label: tbl-W-space
#| tbl-cap: "Enumeration of possible of ranks given 4 observations together with calculated W statistics"
#| tbl-cap-location: margin

# enumerate all rank possibilties (by hand)
r1 <- c(1, -1, 1, 1, 1, -1, -1, -1, 1, 1, 1, -1, -1, -1, 1, -1)
r2 <- c(2, 2, -2, 2, 2, -2, 2, 2, -2, 2, -2, -2, -2, 2, -2, -2)
r3 <- c(3, 3, 3, -3, 3, 3, -3, 3, -3, -3, 3, -3, 3, -3, -3, -3)
r4 <- c(4, 4, 4, 4, -4, 4, 4, -4, 4, -4, -4, 4, -4, -4, -4, -4)

data.w <- rbind(r1, r2, r3, r4)
data.w.ind <- data.w
data.w.ind[data.w < 0] <- 0
r.sum <- apply(data.w.ind, 2, sum)

data.w <- rbind(data.w, r.sum)
rownames(data.w) <- c("id1", "id4", "id3", "id4", "W")
colnames(data.w) <- paste("c", 1:16, sep="")

data.w %>% kable() %>% kable_styling(full_width = TRUE) %>%
  row_spec(5, bold = T, color = "black", background = "#deebf7")

```

- Given 4 observations, we could get ranks $R_i$ of 1, 2, 3 or 4 only
- Further, depending where the observation would be with respect to $m_0$, the rank $R_i$ could be positive or negative. 
- For example, the first column $c1$ corresponds to all 4 observations having positive ranks, so all $x_i - m_0 > 0$, whereas column $c16$ corresponds to all observations having negative ranks, so $x_i - m_0 < 0$. 
- As $W$ test statistics is derived by summing up the positive ranks, we can see by listing all the combinations in the table, that $0 \le W \le10$. 

We can also now write down the probability mass function given the table. 


```{r}
#| label: tbl-W-pmf
#| tbl-cap: "Probability mass function for 4 observations"
#| tbl-cap-location: margin

# calculate probability mass function
W <- data.w[5,]

df.w <- data.frame(W = W) %>%
  group_by(W) %>%
  summarize(n = n()) %>%
  mutate(per = n / 16) 

dist.W <- rbind(W = formatC(df.w$W), `p(W)`=df.w$per)

dist.W %>% t() %>%
  kable(digits = 2) %>%
  kable_styling(full_width = T) %>%
  row_spec(1, )

```


```{r}
#| label: fig-pmf
#| fig-cap: Probablity mass function for observations
#| fig-cap-location: margin

# plot pmf
barplot(df.w$per, names.arg = 0:10, ylab = "p(W)", xlab="W")
```

Now we can use our knowledge from the Probability session on discrete distributions to calculate the probability associated with observing test statistics $W$ given the known probability mass function.

For more examples on how to manually obtain the distribution $W$ under the null hypothesis visit [PennState Elbery collage website](https://online.stat.psu.edu/stat415/lesson/20/20.2).

## In `R` we use `wilcox.test()` function:

```{r}
#| code-fold: false

# run Wilcoxon signed rank test for a median
wilcox.test(x = data.sleep$placebo, 
            y = NULL,
            alternative = "less",
            mu = 7, 
            exact = F
            )
```




