---
title: "Exercises: Principal component analysis"
author: "Mun-Gwan Hong"
date: "2022-09-15"
output:
  html_document:
    theme: null
    self_contained: no
params:
  show_the_output: false
knit: (function(input, ...) rmarkdown::render(input, ..., output_dir= "html"))
---

```{r, setup, echo=FALSE, include=FALSE}
library(ggplot2)

knitr::opts_chunk$set(
  include = params$show_the_output, # control the output
  cache.path = "cache/",   # auto. created unless exists
  results = "markup",
  message = FALSE,    # no message from the code
  warning = FALSE,    # no warning from the code
  fig.dim = c(8, 8)
)
options(digits=2)
```


#### Exercise 1

Let's start principal component analysis (PCA) with Edgar Anderson's iris data,
which is already available as `iris` in R.

a) Examine the contents and data types of the data set *iris*.
   R functions `str`, `summary` and `pairs` can be useful for this.

```{r, fig.dim=c(8, 8)}
str(iris)
summary(iris)
pairs(iris)
```

b) As Setosa is somewhat different from the other two species, we will limit our analysis to these two. 
   Leave the data of Versicolor and Virginica only, and examine the contents of the data.

```{r}
vv <- iris[iris$Species %in% c("versicolor", "virginica"), ]
pairs(vv)
```

c) The first 4 columns contains 4 different measures and the last column has species.
   Split the data into a numeric matrix `m` and a character vector `species`. 
   
```{r}
m <- as.matrix(vv[, 1:4])
species <- vv[, 5]
```

d) Transform each variable to have the same mean 0 and the same standard deviation using `scale`.

```{r}
csm <- scale(m)
```

e) Compute eigenvectors and eigenvalues of the variance-covariance matrix of the transformed (or scaled) `m` using `cov` and `eigen`.

```{r}
s <- cov(csm)
e <- eigen(s)
print(e)
```

f) Compute the scores of the first principal component (PC). 
   Matrix dot product in R is `%*%` as `A %*% B` for $\mathbf{A} \mathbf{B}$.

```{r}
pc1 <- csm %*% e$vectors[, 1]
head(pc1)
```

g) Compute the proportion of variance explained by each PC and find the rotation matrix or loading matrix.

```{r}
e$values / sum(e$values)
e$vectors   # rotation matrix
e$vectors %*% diag(sqrt(e$values))  # loading matrix
```

h) Using `prcomp`, perform PCA.
   Don't forget to add `scale. = TRUE` in the arguments of the function. 
   The default is `scale. = FALSE` though the help of the function states scaling is recommended.

```{r}
pca <- prcomp(m, scale. = TRUE)
```

i) Compare the scores from our manual computation using eigenvector and from `prcomp`.

```{r}
all.equal(pca$x[, 1], pc1[, 1])
```

j) Compute the proportion of explained variance using the output from `prcomp`.
   Compare the results with that of g).

```{r}
pca$sdev^2 / sum(pca$sdev^2)
```

k) Find the rotation matrix or loading matrix from the output of `prcomp` and compare it with that of g).

```{r}
pca$rotation   # rotation matrix
pca$rotation %*% diag(pca$sdev)    # loadings
```

l) Draw a PCA score plot for PC1 and PC2.
   Add colors by `species`.
   What does the plot show?

```{r}
plot(pca$x[, 1:2], pch = 16, col = species)
```
```{r, results='asis', echo=FALSE}
cat("\nTwo species were separated by the first PC.\n\n")
```

m) Draw a PCA score plot for PC2 and PC3.

```{r}
plot(pca$x[, 2:3], pch = 16, col = species)
```

n) Draw a scree plot.

```{r}
plot(pca)
```

o) Draw a bi-plot.
   Can you tell 
   1) which variables are the most important for PC1?
   2) which variables are highly correlated?

```{r}
biplot(pca)
```
```{r, results='asis', echo=FALSE}
cat("\n1) `Sepal.Length`, `Petal.Length` and `Petal.Width` are the most important variables",
    "for PC1, as their loadings are far from zero.\n")
cat("2) They are strongly correlated as they are next to each other in loading plot.\n\n")
```

You may find the loadings displayed in the `biplot` are in different scale from those you computed. 
It is primarily due to the scaling by function `biplot`. 
Please refer to https://stats.stackexchange.com/questions/276645/arrows-of-underlying-variables-in-pca-biplot-in-r for the question. 


&nbsp;

#### Exercise 2

NCI60 data set consists of gene expression values for 6830 genes for 64 cell lines.
Using this data set, repeat the analysis using `prcomp` of Exercise 1.
You can download the data using the command below.

```
m <- t(as.matrix(read.table(
  url("https://web.stanford.edu/~hastie/ElemStatLearn/datasets/nci.data.csv"),
  sep = ",", row.names = 1, header = TRUE
)))
```

```{r nci_data_loading, echo=FALSE, cache=TRUE}
m <- t(as.matrix(read.table(
  url("https://web.stanford.edu/~hastie/ElemStatLearn/datasets/nci.data.csv"),
  sep = ",", row.names = 1, header = TRUE
)))
```

a) Examine the contents and data types of the data set `nci`.
   Check the dimension of the data first. 
   This is much larger than `iris`.

```{r}
str(m)
summary(m[, 1:10])
```

b) Using `prcomp`, perform PCA.

```{r}
pca <- prcomp(m, scale. = TRUE)
```

c) Compute the proportion of variance explained by each PC.

```{r}
pca$sdev^2 / sum(pca$sdev^2)
```

d) Find the rotation matrix.
   Show the first 5 columns and rows only.

```{r}
pca$rotation[1:5, 1:5]
```

e) Draw a PCA score plot for PC1 and PC2.

```{r}
plot(pca$x[, 1:2])
```

f) Draw a PCA score plot for PC2 and PC3.

```{r}
plot(pca$x[, 2:3])
```

g) Draw a scree plot.

```{r}
plot(pca)
```

h) Draw a bi-plot.
   Note it can take long time. 

```{r}
biplot(pca)
```


