---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Generalized linear models

**Aims**

- to briefly introduce GLMs via examples of modeling binary and count response

**Learning outcomes**

- to understand the limits of linear regression and the application of GLMs
- to be able to use `glm()` function to fit and interpret logistic and Poisson regression

```{r, echo=F, message=FALSE, include=FALSE}

library(ggplot2)  
library(kableExtra)
library(ggiraphExtra)
library(magrittr)
```

## Why Generalized Linear Models (GLMs)
- GLMs extend linear model framework to outcome variables that do not follow normal distribution
- They are most frequently used to model binary, categorical or count data
- In the Galapagos Island example we have tried to model Species using linear model
- It kind of worked but the predicted counts were not counts (natural numbers) but rational numbers instead that make no sense when taking about count data
- Similarly, fitting a regression line to binary data yields predicted values that could take any value, including $<0$
- not to mention that it is hard to argue that the values of 0 and 1s are normally distributed

```{r, echo=F, message=F, fig.align="center", fig.width=4, fig.height=3, fig.align="center", fig.cap="Example of fitting linear model to binary data, to model the acceptance to medical school, coded as 1 (Yes) and 0 (No) using GPA school scores. Linear model does not fit the data well in this case"}

medgpa <- read.csv("data/lm/medgpa.csv")

ggplot(data = medgpa, aes(y=Acceptance, x=GPA)) +
  geom_jitter(width=0, height = 0) +
  geom_smooth(method="lm", se=FALSE, color="red")

```

## Logistic regression
- [Yanny or Laurel auditory illusion](https://www.theguardian.com/global/video/2018/may/16/what-do-you-hear-in-this-audio-clip-yanny-or-laurel-takes-internet-by-storm-video) appeared online in May 2018. You could find lots of information about it, together with some plausible explanations why some people hear Yanny and some year Laurel
- One of the explanation is that with age we lose the ability to hear certain sounds
- To see if there is evidence for that, someone has already collected some data for 198 people including their age and gender


```{r, echo=T, collapse=TRUE, fig.align="center", fig.width=8, fig.height=4, fig.cap="Yanny and Laurel auditory illusion data, Yanny (1), Laurel (0)"}
# Read in and preview data
yl <- read.csv("data/lm/YannyLaurel.csv")
head(yl)

# Recode Laurel to 0 and Yanny as 1 in new variable
yl$word <- 0
yl$word[yl$hear=="Yanny"] <- 1

# Make some exploratory plots
par(mfrow=c(1,2))
plot(yl$age, yl$word, pch=19, xlab="age", ylab="", las=1)
boxplot(yl$age~yl$hear, xlab="", ylab="age", col="lightblue")

```

- Since the response variable takes only two values (Yanny or Laurel) we use GLM model 
- to fit **logistic regression** model for the **probability of hearing Yanny**
- we let $p_i=P(Y_i=1)$ denote the probability of hearing Yanny (success)
- and we assume that the response follows binomial distribution: $Y_i \sim Bi(1, p_i)$ distribution 
- We can write the regression model now as: 
$$log(\frac{p_i}{1-p_i})=\beta_0 + \beta_1x_i$$
and given the properties of logarithms this is also equivalent to:
$$p_i = \frac{exp(\beta_0 + \beta_1x_i)}{1 + exp(\beta_0 + \beta_1x_i)}$$
- In essence, the GLM generalizes linear regression by allowing the linear model to be related to the response variable via a **link function**.
- Here, the **link function** $log(\frac{p_i}{1-p_i})$ provides the link between the binomial distribution of $Y_i$ (hearing Yanny) and the linear predictor (age) 
- Thus the **GLM model** can be written as $$g(\mu_i)=\mathbf{X}\boldsymbol\beta$$ where `g()` is the link function.
- We use `glm()` function in R to fit GLM models

```{r, fig.align="center", fig.width=4, fig.height=3, fig.cap="Fitted logistic model to the Yanny and Laurel data"}
# fit logistic regression model
logmodel.1 <- glm(word ~ age, family = binomial(link="logit"), data = yl)

# print model summary
print(summary(logmodel.1))

# plot
ggPredict(logmodel.1)

# to get predictions use predict() functions
# if no new observations is specified predictions are returned for the values of exploratory variables used
# we specify response to return prediction on the probability scale
predict(logmodel.1, type="response")


```

- The regression equation for the fitted model is:
$$log(\frac{\hat{p_i}}{1-\hat{p_i}})=0.97  -  0.02x_i$$
- we see from the output that $\hat{\beta_0} = 0.97$ and $\hat{\beta_1} = -0.02$
- these estimates are arrived at via maximum likelihood estimation, something that is out of scope here
- but similarly to linear models, we can test the null hypothesis $H_0:\beta_1=0$ by comparing,  $z = \frac{\hat{\beta_1}}{e.s.e(\hat{\beta_1)}} = -2.33$ with a standard normal distribution, and the associated value is small so there is enough evidence to reject the null, meaning that age is significantly associated with the probability with hearing Laurel and Yanny, **Wald test**
- the same conclusion can be reached if we compare the **residual deviance**

**Deviance**

- deviance is the number that measures the goodness of fit of a logistic regression model
- we use saturated and residual deviance to assess model, instead of $R^2$ or $R^2(adj)$
- for a GLM model that fits the data well the approximate deviance $D$ is
$$\chi^2(m-p)$$ where $m$ is the number of parameters in the saturated model (full model) and $p$ is the number of parameters in the model of interest
- for our above model we have $274.41  - 268.73 = 5.68$ which is larger than 95th percentile of $\chi^2(197-196)$
```{r, collapse=TRUE}
qchisq(df=1, p=0.95)
```
- i.e. $5.68 > 3.84$ and again we can conclude that age is a significant term in the model

**Odds ratios**

- In logistic regression we often interpret the model coefficients by taking $e^{\hat{\beta}}$
- and we talk about **odd ratios**
- e.g. we can say, given our above model, $e^{-0.02444} = 0.9758562$ that for each unit increase in age the odds of hearing Yanny get multiplied by 0.98

**Other covariates**

- Finally, we can use the same logic as in multiple regression to expand by models by additional variables, numerical, binary or categorical
- E.g. we can test whether there is a gender effect when hearing Yanny or Laurel

```{r, collapse=TRUE, fig.align="center", fig.width=4, fig.height=3, fig.cap="Yanny Laurel data modelled with logistic regression given age and gender. Regression lines in males and females are very alike and the model suggest no gender effect", warning=FALSE}
# fit logistic regression including age and gender
logmodel.2 <- glm(word ~ age + gender, family = binomial(link="logit"), data = yl)

# print model summary
print(summary(logmodel.2))

# plot model
ggPredict(logmodel.2)
```

**Simulated data**

This is beyond the scope of this course but a more advanced model might be needed to better explain these specific data. As an exercise, let us simulate a dataset where the logistic regression would be a better fit (it would probably be the case if the age effect had been larger than the one observed in the Yanny/Laurel example above). 

```{r, echo=T, collapse=TRUE, fig.align="center", fig.width=8, fig.height=4}
# In a similar way as for the first Yanny/Laurel model above (logmodel.1)
# where a binary variable (hearing Yanny/Laurel) was explained by one 
# continuous variable (age), let us simulate the data below:

# - we will simulate a sample of 60 individuals where the binary variable 
# (e.g. hearing Yanny/Laurel) is equal to zero 30 times and to one 30 times

set.seed(1)
n <- 30
binaryVar <- c(rep(0, n), rep(1, n))

# - we would like to simulate a strong effect of the continuous variable 
# so we can simulate people with binaryVar 0 and binaryVar 1 from 
# different distributions. 

distr0 <- rnorm(n, mean=65, sd=15) %>% round()
distr1 <- rnorm(n, mean=25, sd=12) %>% round()

dat <- data.frame(binaryVar=c(rep(0, n), rep(1, n)),
                  continuousVar = c(distr0, distr1))
idx <- sample(1:(2*n), 2*n) 
dat <- dat[idx,] #reorder samples randomly

head(dat)

# Make some exploratory plots
par(mfrow=c(1,2))
plot(dat$continuousVar, dat$binaryVar, pch=19, 
     xlab="Continuous variable (for instance age)", 
     ylab="Binary variable (for instance hearing Yanny/Laurel)", las=1)

boxplot(dat$continuousVar~dat$binaryVar, 
        xlab="Binary variable", 
        ylab="Continuous variable", 
        col="lightblue")
```

```{r, collapse=TRUE, fig.align="center", fig.width=4, fig.height=3, warning=FALSE}

# fit logistic regression model
logmodel.3 <- glm(binaryVar ~ continuousVar, 
                  family = binomial(link="logit"), data = dat)

# print model summary
print(summary(logmodel.3))

# plot
ggPredict(logmodel.3)
```

## Poisson regression
- GLMs can be also applied to count data
- e.g. hospital admissions due to respiratory disease or number of bird nests in a certain habitat
- here, we commonly assume that data follow the Poisson distribution $Y_i \sim Pois(\mu_i)$
- and the corresponding model is 
$$E(Y_i)=\mu_i = \eta_ie^{\mathbf{x_i}^T\boldsymbol\beta}$$ with a log link $\ln\mu_i = \ln \eta_i + \mathbf{x_i}^T\boldsymbol\beta$

**Data set**
Suppose we wish to model $Y_i$ the number of cancer cases in the i-th intermediate geographical location (IG) in Glasgow. We have collected data for 271 regions, a small areas that contain between 2500 and 6000 people. Together with cancer occurrence with have data:

- Y\_all: number of cases of all types of cancer in the IG in 2013
- E\_all: expected number of cases of all types of cancer for the IG based on the population size and demographics of the IG in 2013
- pm10: air pollution
- smoke: percentage of people in an area that smoke
- ethnic: percentage of people who are non-white
- log.price: natural log of average house price
- easting and northing: co-ordinates of the central point of the IG divided by 10000

We can model the **rate of occurrence of cancer** using the very same `glm` function:¨
- now we use **poisson family distribution** to model counts
- and we will include an **offset term** to the model as we are modeling the rate of occurrence of the cancer that has to be adjusted by different number of people living in different regions

```{r, collapse=TRUE}
# Read in and preview data
cancer <- read.csv("data/lm/cancer.csv")
head(cancer)

# fit Poisson regression
epid1 <- glm(Y_all ~ pm10 + smoke + ethnic + log.price + easting + northing + offset(log(E_all)), 
             family = poisson, 
             data = cancer)

print(summary(epid1))
```

**Hypothesis testing, model fit and predictions**

- follows stay the same as for logistic regression

**Rate ratio**

- similarly to logistic regression, it is common to look at the $e^\beta$
- for instance we are interested in the effect of air pollution on health, we could look at the pm10 coefficient
- coefficient is positive, 0.0500269, indicating that cancer incidence rate increase with increased air pollution
- the rate ratio allows us to quantify by how much, here by a factor of $e^{0.0500269} = 1.05$ 

-----

## Exercises (GLMs)

```{exercise, "glm-rerun"}

Make sure you can run and understand the above code for logistic and Poisson regression
The data files can be downloaded in Canvas from Files/data-exercises/linear-models. An Rmd file with the code from the book chapter is provided in Files/exercises/GLM

```


```{exercise, "glm-wcgs"}

Additional practice with a bigger more realistic data set. 

What might affect the chance of getting a heart disease? One of the earliest studies addressing this issue started in 1960 in 3154 healthy men in the San Francisco area. At the start of the study all were free of heart disease. Eight years later the study recorded whether these men now suffered from heart disease (chd), along with many other variables that might be related. 

The data is available from the faraway package and includes variables:

- age: age in years
- height: height in inches
- weight: weight in pounds
- sdp: systolic blood pressure in mm Hg
- dbp: diastolic blood pressure in mm Hg
- chol: Fasting serum cholesterol in mm %
- behave: behavior type which is a factor with levels A1 A2 B3 B4
- cigs: number of cigarettes smoked per day
- dibep: behavior type a factor with levels A (Agressive) B (Passive)
- chd: coronary heat disease developed is a factor with levels no yes
- typechd: type of coronary heart disease is a factor with levels angina infdeath none silent 
- timechd: Time of CHD event or end of follow-up
- arcus: arcus senilis is a factor with levels absent present

a) using logistic regression, can you discover anything interesting about the probability of developing heart disease (chd)?
b) using Poisson regression, can you comment about the numbers of cigarettes smoked (cigs)?

```


```{r, collapse=TRUE}

library(faraway)
data(wcgs, package="faraway")

head(wcgs)

```

----

Answers to selected exercises

Exr. \@ref(exr:glm-wcgs) possible solution

a) probability of developing heart disease

We first check the relationship between variables to gain more understanding of the data. We discover that a couple of variables are exactly collinear with other variables, including `typechd`, `timechd` and `dibep`. We do not include these in the model. 
```{r, collapse=TRUE}

# `chd` and `typechd` were correlated.
with(wcgs, table(chd, typechd))

# `timechd` is an outcome variable affected by `chd`.
by(wcgs$timechd, wcgs$chd, summary)

# `behave` has more detailed info of `dibep` -> exact collinearity
with(wcgs, table(behave, dibep))

```

We fit logistic regression model to explain the probability of developing cardiac disease (`chd`) given the remaining variables
```{r}
model1 <- glm(chd ~ . - typechd - timechd - dibep, data = wcgs, family = binomial)
summary(model1)
```

And we notice that many variables including `age`, `chol`, and `cigs`, were significantly associated with heart disease development. For example, increment of one mm % of Fasting serum cholesterol (`chol`) elevated the odds of the disease by a factor of $e^{0.010726} = 1.010784$ after adjustment for the effects of the other variables.

b) numbers of cigarettes smoked

Many variables were correlated with the number of cigarettes.
For example, one mm Hg increase of systolic blood pressure was correlated with the increase of average number of cigarettes smoked by a factor of $e^{0.0024264} = 1.002429$. 
```{r}
# check distribution
hist(wcgs$cigs, breaks = 25)

# Poisson regression for age
model2 <- glm(cigs ~ age, data = wcgs, family = poisson)
summary(model2)

# Poisson regression for weight
model3 <- glm(cigs ~ weight, data = wcgs, family = poisson)
summary(model3)

# Poisson regression for systolic blood pressure
model4 <- glm(cigs ~ sdp, data = wcgs, family = poisson)
summary(model4)

```




