---
title: "Mixed Effects Models"
author: "Eva Freyhult"
date: "2026-09-09"
date-format: long
institute: NBIS, SciLifeLab
embed-resources: true
format: 
  revealjs:
    slide-number: true
    theme: [default]
editor: source
execute: 
  cache: true
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(lme4)
library(nlme)
```

## Introduction to mixed models {.smaller}

Grouped observations are very common in the life sciences:

- **Repeated measures**: multiple measurements from the same subject
- **Longitudinal studies**: repeated observations over time
- **Nested designs**: e.g. mice within cages within labs
- **Crossed designs**: e.g. measurements involving both lab and technician
- **Experimental batches**: measurements from different labs, hospitals, plates, or regions

Ordinary linear regression assumes that observations are **independent**.
When observations are grouped, this assumption is often violated.

---

## Simple linear regression {.smaller}

:::{.columns}
:::{.column width="50%"}
```{r figlm}
#| fig-height: 7
#| fig-width: 7
sleep <- sleepstudy |> filter(Subject %in% c(333, 334, 369, 349, 372), Days>=2) |> mutate(Days=Days-2)
sleep2 <- sleepstudy |> filter(Days>=2) |> mutate(Days=Days-2)
sleep |> ggplot(aes(x=Days, y=Reaction)) + geom_point() + geom_smooth(method="lm", se=FALSE) + theme_bw() +
  labs(title = "Reaction time vs Days of sleep deprivation",
       x = "Days of sleep deprivation",
       y = "Reaction time (ms)")
```

$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i
$$

:::
:::{.column width="50%"}

#### Assumptions

- **Linearity**: The relationship between x and y is linear
- **Independence**: Observations are independent
- **Homoscedasticity**: Constant variance of residuals
- **Normality**: Residuals are normally distributed

The key assumption for grouped data is often **independence**.
:::
:::

## [Example: Age and plasma concentration]{style="font-size: 0.8em;"}

Is there an association between age and plasma concentration?

```{r simgene}
simulate_gene_expression <- function(Nsubj = 4, n=10, Gslope = 2, slope=-4, Gint = 100, xm=30, xmsd=5, xsd=2, esd=2) {
  xmean <- rnorm(Nsubj, mean = xm, sd = xmsd)
  
data.frame(subject= rep(as.character(1:Nsubj), each = n),
           xmean = rep(xmean, each = n)) |>
           mutate(xd = rnorm(Nsubj * n, mean = 0, sd = xsd), x=xmean+xd, y = Gint + Gslope * xmean + slope * xd + rnorm(Nsubj * n, sd = esd))
  
}
set.seed(3431)
df <- simulate_gene_expression(xmsd=10, Gslope=2, slope=-2, xsd=5, esd=8) |> rename(group=subject, age=x, conc=y)

df |> ggplot(aes(x=age, y=conc)) + geom_point() + geom_smooth(method="lm", se=FALSE) + theme_bw()
```

## [Example: Age and plasma concentration]{style="font-size: 0.8em;"} {.smaller}

```{r m0, echo=TRUE}
m0 <- lm(conc ~ age, data = df)
summary(m0)
```

## [Example: Age and plasma concentration]{style="font-size: 0.8em;"}

Are the assumptions fulfilled?

```{r figgeneprot0}
df |> ggplot(aes(x=age, y=conc, color=group)) + geom_point() + theme_bw()
```

:::{.notes}
The points form clusters. Observations within a group are likely to be more similar to each other than observations from different groups.
:::

## Group as fixed effect

```{r m1, echo=TRUE}
m1 <- lm(conc ~ age + group, data = df)
summary(m1)
```

## [Example: Age and plasma concentration]{style="font-size: 0.8em;"}


```{r}
df$pred <- predict(m1)
df |> ggplot(aes(x=age, y=conc, color=group)) + geom_point() + geom_line(aes(y=pred)) + theme_bw() + theme(legend.position = "none")
```


## Linear mixed models

:::{.notes}
In our example there were many parameters to fit and what if there were even more batches/groups to consider? We can use linear mixed models to account for the grouping structure of the data.
:::

A linear mixed model accounts for the grouping structure by including **random effects**.

Ignoring the grouping structure can lead to:

  - Misleading conclusions
  - Incorrect standard errors
  - Inflated type I error rates

## Fixed effects and random effects

**Fixed effects**

- population-level effects
- estimated explicitly
- typically the main effects of interest

**Random effects**

- account for variation between groups
- assume group effects come from a distribution
- usually summarized by variance components rather than individual parameters

For grouped data:

- **Fixed group effect**: estimate each group separately
- **Random group effect**: model between-group variability

## Grouping structure

The grouping structure can be cohorts, batches, subjects, schools, hospitals, doctors, cages, or labs.

It should be determined from the **study design** and not inferred only from the data.

Note: the grouping variable is categorical.

Note 2: Very few groups makes random-effects estimation unstable.


## Model with fixed and random effects

Once the grouping is decided, we can decide what effects are fixed and what effects are random.

The effects can be:

* Intercept
* Slope
* Interaction


## Model with fixed and random effects

```{mermaid}
%%| mermaid-format: png
flowchart TD
  Q1["Does the grouping structure affect the model?"] -->|No| A1["Model as Fixed Effect"]
  Q1 -->|Yes| Q2["Do we want to estimate the effect for each group?"]
  Q2 -->|Yes| A2["Model group as fixed effect,<br>i.e. include group as covariate"]
  Q2 -->|No| Q3["Are there many groups?"]
  Q3 -->|Yes| A3["Use Random Effects"]
  Q3 -->|No| A4["Random Effects may be unstable,<br>consider Fixed Effects"]
```

## [Example: Age and plasma concentration]{style="font-size: 0.8em;"}

We believe that:

- the effect of age on concentration is the same across groups, hence **fixed**
- the intercept (baseline concentration) varies between groups

```{r figgeneprotgroup2}
df |> ggplot(aes(x=age, y=conc, color=group)) + geom_point() + theme_bw() + theme(legend.position = "none")
```

::: {.notes}
Clinics may differ in:

- Sampling methods
- Storage procedures
- Measurement equipment
:::

## Random intercept model

```{r figgeneprotgroup4}
mm <- lmer(conc ~ age + (1 | group), data = df)
#Show the predicted lines
df$pred <- predict(mm)

pl <- df |> ggplot(aes(x=age, y=conc, color=group)) + geom_point() + theme_bw() + theme(legend.position = "none")
pl + geom_line(aes(y=pred)) + labs(title = "Predicted concentration by age and group")
```


## Random intercept model

$$y_{ij} = \beta_0 + \beta_1 x_{ij} + b_{0i} + \varepsilon_{ij}$$
  
- $y_{ij}$: outcome, dependent variable (concentration)
- $\beta_0, \beta_1$: fixed effects (intercept and slope)
- $b_{0i}$: random intercept for group $i$
- $b_{0i} \sim N(0, \sigma_{b_0}^2)$, where $\sigma_{b_0}$ is the standard deviation of the random intercept
- $\varepsilon_{ij}$: residual error

## Mixed model in R

```{r lmer0, echo=TRUE}
mm <- lmer(conc ~ age + (1 | group), data = df)
summary(mm)
```

- `age`: fixed effect
- `(1 | group)`: random intercept for group

## Interpreting the model

The output contains:

- **Fixed effects**: the overall association between age and concentration
- **Random effects**: variability in intercepts across groups
- **Residual variance**: remaining within-group variation

```{r varcomp, echo=TRUE}
var_components <- as.data.frame(VarCorr(mm))
icc <- var_components$vcov[1] / sum(var_components$vcov)
icc
```

The intra-class correlation coefficient (ICC) is the proportion of total variance attributable to the grouping structure.

## Predicting fixed and random effects

Fixed effects can be extracted using `fixef()`.

```{r fixef, echo=TRUE}
fixef(mm)
```

The random effects can be predicted using `ranef()`.

```{r ranef, echo=TRUE}
ranef(mm)
```

The fixed effects give the overall mean structure, and the random effects are deviations for each group.

## Random slope 

```{r}
mm2 <- lmer(conc ~ age + (1 + age | group), data = df)
#Show the predicted lines
df$pred2 <- predict(mm2)
pl <- df |> ggplot(aes(x=age, y=conc, color=group)) + geom_point() + theme_bw() + theme(legend.position = "none")
pl + geom_line(aes(y=pred2)) + labs(title = "Predicted concentration by age and group")
```

## Random slope

If you believe that the slope also varies between groups, include random slope for age in addition to random intercept.

  $$y_{ij} = \beta_0 + \beta_1 x_{ij} + b_{0i} + b_{1i} x_{ij} + \varepsilon_{ij} = \beta_0 + b_{0i} + (\beta_1 + b_{1i}) x_{ij} + \varepsilon_{ij}$$
  
- $y_{ij}$: outcome, dependent variable (concentration)
- $\beta_0, \beta_1$: fixed effects (intercept and slope)
- $b_{0i}$: random intercept per group (for group $i$)
- $b_{0i} \sim N(0, \sigma_{b_0}^2)$, where $\sigma_{b_0}$ is the standard deviation of the random intercept
- $b_{1i} \sim N(0, \sigma_{b_1}^2)$, where $\sigma_{b_1}$ is the standard deviation of the random slope
- $\varepsilon_{ij}$: residual error


## Random slope in R

```{r rslope, echo=TRUE}
mm2 <- lmer(conc ~ age + (1 + age | group), data = df)
summary(mm2)
```

```{r eval=FALSE, echo=FALSE}
sleepstudy |> ggplot(aes(x=Days, y=Reaction, color=Subject)) + geom_point() + geom_line() + theme_bw() +
  labs(title = "Reaction time vs Days of sleep deprivation",
       x = "Days of sleep deprivation",
       y = "Reaction time (ms)") + theme_bw() + theme(legend.position = "none")
mm2 <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepall)
summary(mm2)
```

## More complex grouping structures

So far we have considered one grouping factor.

In real studies, grouping structures can be more complex and involve more than one level.

## Nested and crossed designs {.incremental}

**Nested**: lower-level units are associated with only **one** higher-level unit

**Crossed**: lower-level units are associated with **multiple** higher-level units

---

## Nested designs

In a nested design, observations are organized hierarchically. For example, patients within clinics and clinics within regions.

```{mermaid}
graph TD
    A[Region 1] 
    A --> B[Clinic 1]
    A --> C[Clinic 2]
    A --> D[Clinic 3]
    
    B -.-> M1[Patient 1]
    B -.-> N2[Patient 2]
    B -.-> O3[Patient 3]

    C -.-> P4[Patient 4]
    C -.-> Q5[Patient 5]
    C -.-> R6[Patient 6]
    
    D -.-> S7[Patient 7]
    D -.-> T8[Patient 8]
    D -.-> U9[Patient 9]
    
    classDef region fill:#f9f,stroke:#333,stroke-width:2px;
    classDef clinic fill:#ccf,stroke:#333,stroke-width:2px;
    classDef patient fill:#cff,stroke:#333,stroke-width:1px;
    
    class A region;
    class B,C,D clinic;
    class M1,N2,O3,P4,Q5,R6,S7,T8,U9 patient;
    
    linkStyle default stroke-width:4px;
```


```{mermaid}
graph TD
    E[Region 2]
    E --> F[Clinic 4]
    E --> G[Clinic 5]
    E --> H[Clinic 6]
    
    F -.-> V10[Patient 10]
    F -.-> W11[Patient 11]
    F -.-> X12[Patient 12]

    G -.-> Y13[Patient 13]
    G -.-> Z14[Patient 14]
    G -.-> A15[Patient 15]
    
    H -.-> B16[Patient 16]
    H -.-> C17[Patient 17]
    H -.-> D18[Patient 18]
    
    I[Region 3]
    I --> J[Clinic 7]
    I --> K[Clinic 8]
    I --> L[Clinic 9]
    
    J -.-> E19[Patient 19]
    J -.-> F20[Patient 20]
    J -.-> G21[Patient 21]

    K -.-> H22[Patient 22]
    K -.-> I23[Patient 23]
    K -.-> J24[Patient 24]
    
    L -.-> K25[Patient 25]
    L -.-> L26[Patient 26]
    L -.-> M27[Patient 27]

    classDef region fill:#f9f,stroke:#333,stroke-width:2px;
    classDef clinic fill:#ccf,stroke:#333,stroke-width:2px;
    classDef patient fill:#cff,stroke:#333,stroke-width:1px;
    
    class E,I region;
    class F,G,H,J,K,L clinic;
    class V10,W11,X12,Y13,Z14,A15,B16,C17,D18,E19,F20,G21,H22,I23,J24,K25,L26,M27 patient;
    
    linkStyle default stroke-width:4px;
```

## Nested model


```{r nesteddata}
simulate_grouped_trend_nested <- function(region_count = 3, clinics_per_region = 5, measurements_per_clinic = 10,
                                                         global_slope = -10, global_intercept = 30, 
                                                         region_effects = c(), region_slopes = c(),
                                                         clinic_effects = list(), clinic_slopes = list()) {
  set.seed(123) # Setting a seed for reproducibility
  
  # Initialize an empty data frame to store the simulated data
  data <- data.frame(x = numeric(), y = numeric(), clinic = integer(), region = integer())

  clinic_id_counter = 1 # Initialize a counter for clinic IDs across regions
  
  # Loop to create data for each region
  for (region in 1:region_count) {
    # Use the specific region effect and slope
    region_effect_adj = region_effects[region]
    region_slope_adj = region_slopes[region]
    
    # Loop to create data for each clinic within a region
    for (clinic in 1:clinics_per_region) {
      x_start = runif(1, min = 0, max = 10) + 10 * (clinic - 1) # More continuous x values across clinics
      x = runif(measurements_per_clinic, min = x_start, max = x_start + 10) # Continuous x for measurements
      
      # Use the specific clinic effect and slope
      clinic_effect = clinic_effects[[region]][clinic]
      clinic_slope = clinic_slopes[[region]][clinic]
      
      # Simulate measurements for each clinic
      for (i in 1:measurements_per_clinic) {
        # Model y incorporating both global and specific slopes and effects
        y = (global_intercept + region_effect_adj + clinic_effect) + 
            (global_slope + region_slope_adj + clinic_slope) * x[i] + 
            rnorm(1, mean = 0, sd = 1) # Assuming measurement_noise_sd is constant for simplicity
        
        # Combine this measurement with the overall dataset
        data = rbind(data, data.frame(x = x[i], y = y, clinic = clinic_id_counter, region = region))
      }
      
      clinic_id_counter <- clinic_id_counter + 1 # Increment clinic ID for unique identification across regions
    }
  }
  
  return(data)
}




set.seed(10)
data_int_nested<-simulate_grouped_trend_nested(region_count = 4,
                                               clinics_per_region = 3,
                                               measurements_per_clinic = 20,
                                               global_slope = 0,global_intercept = 100,
                                               region_effects = rnorm(4,mean = 0,sd = 20),
                                               region_slopes=rnorm(4,mean = 0.1,sd = 2),
                                               clinic_effects = lapply(1:4,function(i){rnorm(3,mean = 0,sd = 10)}),
                                               clinic_slopes = lapply(1:4,function(i){rnorm(3,mean = 0.1,sd = 1)}))

data_int_nested |> 
  ggplot(aes(x = x, y = y, color = factor(region), shape = factor(clinic))) +
  geom_point() +
  labs(title = "Simulated Data with Nested Structure",
       x = "Age",
       y = "Protein Expression",
       color = "Region",
       shape = "Clinic") +
  scale_shape_manual(values=1:12) + theme_bw()
```

## Nested models in R {.smaller .incremental}

In R and `lme4`, random effects for nested groupings can be included. Random intercepts for both regions and clinics nested within regions can be included as follows;

```{r nestedmodel, echo=TRUE}
nested_model <- lmer(y ~ x + (1 | region/clinic), 
                     data = data_int_nested)
```

`(1 | region/clinic)` specifies random intercepts for clinics nested within regions.

. . .

Note that

`(1 | region/clinic)` is identical to

`(1 | region) + (1 | region:clinic)`

. . .

A third way to implement the same model is to make the nesting implicit by coding the lower levels (here clinics) uniquely, so that a clinic name only exists in one region. With such a naming scheme, we can use the notation `(1|region) + (1|clinic)`.


## Nested models in R {.smaller}

```{r nestedmodel1, echo=TRUE}
nested_model <- lmer(y ~ x + (1 | region/clinic), data = data_int_nested)
summary(nested_model)
```

## Crossed designs

In a crossed design, an observation is associated with multiple groupings. For example, patients may be treated by different doctors at several clinics.

```{mermaid}
graph TD

    A[Clinic 1]
    E[Clinic 2]
    I[Clinic 3]

    B[Doctor 1]
    C[Doctor 2]
    D[Doctor 3]

    M1[Patient 1]
    M2[Patient 2]
    M3[Patient 3]
    M4[Patient 4]
    M5[Patient 5]
 

    A -.-> B
    A -.-> C
    A -.-> D
    E -.-> B
    E -.-> C
    E -.-> D
    I -.-> B
    I -.-> C
    I -.-> D
   
    %% Patients connected to clinics to show they can come from any clinic, implying crossed with regions
    B --> M1
    B --> M2
    B --> M3
    B --> M4
    B --> M5
    
    C --> M1
    C --> M2
    C --> M3
    C --> M4
    C --> M5
    
    D --> M1
    D --> M2
    D --> M3
    D --> M4
    D --> M5
    
    %% Additional patient connections omitted for brevity

    classDef region fill:#f9f,stroke:#333,stroke-width:2px;
    classDef clinic fill:#ccf,stroke:#333,stroke-width:2px;
    classDef patient fill:#cff,stroke:#333,stroke-width:1px;
    
    class A,E,I region;
    class B,C,D,F,G,H,J,K,L clinic;
    class M1,M2,M3,M4,M5 patient;
    
    linkStyle default stroke-width:4px;
```


## Crossed model in R

A crossed design can be coded as:

`y ~ x + (1 | clinic) + (1 | doctor)`

## Model selection in mixed models 

:::{.incremental}

Mixed models can become complex and model selection can help us choose a suitable model.

Likelihood ratio tests (LRT) can be used to compare nested models

We can compare:

- different **random-effects** structures
- different **fixed-effects** structures
:::

## What is LRT? {.smaller .incremental}

The likelihood is a measure of how well a model explains the data.

. . .

The likelihood ratio test compares the likelihoods of two models: the **null** model, which is a simpler model, and the **alternative**, more complex, model.

. . . 

$$\lambda = - 2\ln{ \frac{L(restricted\,model)}{L(full\, model)}} $$

The test statistic is approximately $\chi^2$-distributed with degrees of freedom equal to the difference in the number of parameters between the two models.

. . .

A significant difference in likelihoods suggests that the more complex model explains the data better than the simpler model.

. . .

For mixed models, we can use the `anova()` function to perform LRT. We will start by testing the random effects structure, without change of fixed effects. Then we will test the fixed effects structure, keeping the random effects fixed.

## ML and REML {.smaller}

Mixed models can be fitted using:

- **Maximum Likelihood (ML)**
- **Restricted Maximum Likelihood (REML)**

Use:

- **REML** when comparing random-effects structures with the same fixed effects (`refit = FALSE`)
- **ML** when comparing fixed-effects structures (`refit = TRUE`)
- **REML** for the final model

## Random effects comparison

Sleep study data: random intercept

```{r sleepstudy1, echo=TRUE}
sleepall <- sleepstudy |> filter(Days>=2) |> mutate(Days=Days-2)
m1 <- lmer(Reaction ~ Days + (1 | Subject), data = sleepall)
summary(m1)
```

## Random effects comparison

Sleep study data: random intercept and random slope for Days.


```{r sleepstudy2, echo=TRUE}
m2 <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepall)
summary(m2)
```
## Random effects eomparison

Compare the two models using a likelihood ratio test.

```{r sleepstudyLRT, echo=TRUE}
anova(m1, m2, refit = FALSE)
```

## Fixed effects comparison

We can also compare models with different fixed effects structures. For example, we can compare a model with a fixed effect of `Days` to a model with a fixed effect of `Days` and `Sex`.

```{r}
sleepall <- sleepall |> mutate(sex=sample(c("M", "F"), nrow(sleepall), replace=TRUE))
```


```{r sleepstudy3, echo=TRUE}
m2 <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepall)
m3 <- lmer(Reaction ~ Days + sex + (1 + Days | Subject), data = sleepall)
anova(m2, m3, refit=TRUE)
```

## P-value calculations {.smaller}

As you have noticed, `lme4` does not report p-values for the fixed effects.

In linear regression, p-values are calculated based on the t-distribution of the estimated coefficients. The calculation of denominator degrees of freedom is straightforward since all observations are independent.

In mixed models, p-values are more complicated because the denominator degrees of freedom are not straightforward.

Approximation methods such as **Satterthwaite** and **Kenward-Roger** can be used.

---

## `lmerTest`

```{r lmerTest, echo=TRUE}
library(lmerTest)
mm <- lmer(conc ~ age + (1 | group), data = df)
summary(mm)
```

## `lmerTest`

```{r lmerTest2, echo=TRUE}
mmsleep <- lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepall)
summary(mmsleep)
```

---

# Summary

- Mixed models are used when observations are grouped and therefore not independent
- Fixed effects describe population-level effects
- Random effects account for variation between groups
- Random intercepts allow different baselines
- Random slopes allow different responses to a predictor
- Grouping structures can be nested or crossed
- Model comparison should distinguish between fixed and random parts
- Use REML for the final model, and ML when comparing fixed effects
- Approximate p-values can be obtained using `lmerTest`
```