---
editor: source
---

# Multiple testing

Multiple testing is especially important in life-science applications such as genomics, transcriptomics, proteomics, and metabolomics, where thousands of hypotheses may be tested simultaneously.

```{r setup, include=FALSE}
library(tidyverse)
library(gt)
library(kableExtra)
options(digits=4)
```


## Error types

```{r}
#| label: tbl-errortypes2
#| tbl-cap: "A hypothesis test either rejects $H_0$ or fails to reject $H_0$. The decision may be correct or incorrect depending on whether $H_0$ is true or false. TN = true negative, TP = true positive, FN = false negative, FP = false positive."
tibble(
  Decision = c("Fail to reject H₀", "Reject H₀"),
  `H₀ is true` = c("TN", "Type I error, false alarm, FP"),
  `H₀ is false` = c("Type II error, miss, FN", "TP")
) |>
  gt() |>
  cols_align(align = "center", columns = everything())
```


Remember from @sec-errortype that the probability of type I and II errors are denoted $\alpha$ and $\beta$, respectively;


$$\alpha = P(\textrm{type I error}) = P(\textrm{false alarm}) = P(\textrm{Reject }H_0 \mid H_0 \text{ is true})$$
$$\beta = P(\textrm{type II error}) = P(\textrm{miss}) = P(\textrm{Fail to reject }H_0 \mid H_1 \text{ is true})$$
and the statistical power

$$\textrm{power} = 1 - \beta = P(\textrm{Reject }H_0 \mid H_1\textrm{ is true}).$$
```{r}
#| label: fig-alphabeta
#| fig-cap: "The probability density functions under H0 and H1, respectively. The probability  of type I error ($\\alpha$) and type II error ($\\beta$) are indicated."
#| fig-width: 5
x<-seq(-3,6,0.01)
d <- 2.5
q=qnorm(0.975)
df <- data.frame(x=x, h=rep(c("H0", "H1"), each=length(x))) |> mutate(m=c(H0=0, H1=d)[h], f=dnorm(x, m, 1))
ggplot(df, aes(x=x, y=f, color=h, fill=h)) + geom_line()  + geom_area(data=df %>% filter(x>q, h=="H0"), alpha=0.5)  + geom_area(data=df %>% filter(x<q, h=="H1"), alpha=0.5) + xlab("x") + ylab("f(x)") + theme_bw() + scale_color_manual("", guide="none", values=c("black", "orange")) + scale_fill_manual("", guide="none", values=c("black", "orange")) + annotate("label", x=d+1, y=dnorm(1,0,1), label="H1", color="orange") + annotate("label", x=-1, y=dnorm(1,0,1), label="H0", color="black") + annotate("label", x=2.2, y=0.02, label="alpha", parse=TRUE) + annotate("label", x=1.2, y=0.08, label="beta", parse=TRUE, color="orange")
```

  
## Multiple testing
  
  If a single test is performed, we know that
  
  -   P(One type I error) = $\alpha$
  -   P(No type I error) = $1 - \alpha$
  
If $m$ independent tests are performed (e.g. investigate many genes or proteins), the risk of at false alarm (type I error) increases;
  
  -   P(No type I errors in $m$ tests) = $(1 - \alpha)^m$
  -   P(At least one type I error in $m$ tests) = $1 - (1 - \alpha)^m$
  
```{r multiple, echo=FALSE, out.width="50%", fig.align="center", fig.width=4, fig.height=4}
a=0.05
k <- 1:100
ggplot(data.frame(k=k, p = 1-(1-a)^k), aes(x=k, y=p)) + geom_line() + xlab("Number of tests") + ylab("P(At least one type I error)") + theme_bw() + annotate("label", x=75, y=0.2, label="alpha == 0.05", parse=TRUE)
```

Two common principles for dealing with multiple testing are control of family-wise error rate or false discovery rate.

-   FWER: family-wise error rate, control the probability of one or more false positive $P(N_{FP}>0)$, e.g. Bonferroni, Holm
-   FDR: false discovery rate, control the expected value of the proportion of false positives among hits (rejected hypotheses), $E[N_{FP}/(N_{FP}+N_{TP})]$ (note: for 0 positives the ratio is defined to be 0), e.g. Benjamini-Hochberg, Storey

## Bonferroni correction

To achieve a family-wise error rate of $FWER \leq \gamma$ when performing $m$ tests, declare significance and reject the null hypothesis for any test with $p \leq \gamma/m$.

A disadvantage of the Bonferroni correction is that it can be overly conservative, especially when many tests are performed.

## Benjamini-Hochbergs FDR procedure

```{r}
tibble(
  Decision = c("Fail to reject H₀", "Reject H₀"),
  `H₀ is true` = c("TN", "FP"),
  `H₀ is false` = c("FN", "TP")
) |>
  gt() |>
  cols_align(align = "center", columns = everything())
```

The false discovery rate is the expected proportion of false positives among 'hits', i.e. $\frac{FP}{TP+FP}$.

The Benjamini-Hochberg procedure controls the FDR level, $\gamma$, underthe assumption of independence (and also under certain forms of positive dependence) when performing $m$ tests, as follows:

1.  Sort the p-values $p_1 \leq p_2 \leq \dots \leq p_m$.
2.  Find the maximum $j$ such that $p_j \leq \gamma \frac{j}{m}$.
3.  Declare significance for all tests $1, 2, \dots, j$.

## 'Adjusted' p-values

Sometimes an adjusted significance threshold is not reported, but instead 'adjusted' p-values are reported.

-   Using Bonferroni's method the 'adjusted' p-values are:
  
  $\tilde p_i = \min(m p_i, 1)$.

A feature's Bonferroni-adjusted p-value represents the smallest family-wise error rate at which that feature would be deemed significant.

-   Benjamini-Hochberg's 'adjusted' p-values are often referred to as $q$-values and can be compted after sorting the p-values as follows:
  
  $q_i = \min(\frac{m}{i} p_i, 1)$
  
  A feature's $q$-value can be interpreted as the lowest FDR at which the corresponding null hypothesis will be rejected, i.e. the feature will be deemed significant.

In practice, adjusted p-values are usually computed using software such as the R function `p.adjust`.  

::: {#exm-10000}
# 10000 independent tests

```{r padjust, results="asis"}
p <- sort(c(1.7e-8, 5.8e-8, 3.4e-7, 9.1e-7, 1.0e-6, 2.4e-6, 3.6e-5, 2.3e-5, 2.3e-4, 2.2e-4, 8.9e-3,7.3e-4, 0.0045, 0.0032, 0.0087, 0.012, 0.014, 0.045, 0.08, 0.23))
## Here, only 20 p-values are shown, but the adjustments are computed as if they came from a total of 10000 tests.
kable(data.frame(`p-value`=sprintf("%.3g", p), `adj p (Bonferroni)` = p.adjust(p, "bonferroni", n = 10000), `q-value (B-H)`=p.adjust(p, "BH", n = 10000), check.names=FALSE))
```
:::