---
editor: source
---

# Exercises: Hypothesis tests, parametric {.unnumbered}

::: {#exr-Hb1}
The hemoglobin value (Hb) in women is on average 140 g/L. You observe the following Hb values in a set of five male blood donors: 154, 140, 147, 162, 172. Assume that Hb is normally distributed. Is there a reason to believe that the mean Hb value in men differs from that in women?

:::{.callout-note collapse="true"}

# Solution
Let $X$ denote the Hb value in g/L for male blood donors.

$$H_0: \mu = 140$$
$$H_1: \mu \neq 140$$

Use the significance level $\alpha=0.05$.

Use the test statistic
$T = \frac{\bar X - \mu_0}{SE},$
where $\mu_0$ is the population mean under $H_0$, i.e. 140, and the standard error of the sample mean is estimated by
$SE \approx \frac{s}{\sqrt{n}}.$

Compute $t_{obs} = \frac{\bar x-140}{s/\sqrt{n}}.$

```{r Hb1, echo=TRUE}
x <- c(154, 140, 147, 162, 172)
## sample mean
m <- mean(x)
## estimated standard error of the sample mean
SE <- sd(x)/sqrt(5)
## Observed value of the test statistic
(tobs <- (m - 140)/SE)
```

Compute the p-value. For a two-sided test $P(|T|>|t_{obs}|) = P(T>|t_{obs}|) + P(T<-|t_{obs}|)$.

```{r, echo=TRUE}
## P(T>tobs)
pt(tobs, df=4, lower.tail=FALSE)
## p = P(t>tobs) + P(t<-tobs) = 2 * P(t>tobs)
(p <- 2*pt(tobs, df=4, lower.tail=FALSE))
```

As $p > \alpha$, we fail to reject the null hypothesis. The data do not provide evidence that the Hb value in men differs from that of women (140 g/L).

Note that the same can be achieved using the function `t.test`;

```{r}
t.test(x, mu=140)
```

:::
:::

::: {#exr-Hb2}

The hemoglobin value (Hb) in men is on average 188 g/L. The Hb values in @exr-Hb1 were actually measured after the men had donated blood. Is there a reason to believe that the mean Hb level for men after blood donation is less than 188 g/L?

:::{.callout-note collapse="true"}

# Solution

$H_0: \mu = 188$ $H_1: \mu < 188$

Use the significance level $\alpha=0.05$.

Use the test statistic $T = \frac{\bar X - \mu_0}{SE},$ where $\mu_0=188$ and the standard error is estimated by $SE \approx \frac{s}{\sqrt{n}}.$

Compute $t_{obs} = \frac{\bar x - 188}{s/\sqrt{n}}$

```{r, echo=TRUE}
## Observed value of test statistic
(tobs <- (m-188)/SE)
```

Compute the p-value, $P(T<t_{obs})$

```{r, echo=TRUE}
## p=P(T<tobs)
(p <- pt(tobs, df=4, lower.tail=TRUE))
```

As $p < \alpha$ the null hypothesis is rejected and we can conclude that the data provide evidence that the mean Hb value after blood donation is less than 188 g/L.

Note that the same can be achieved using the function `t.test`;

```{r echo=TRUE}
t.test(x, mu=188, alternative="less")
```

:::
:::


::: {#exr-Hb3}

By observing the Hb values in 5 male blood donors: 154, 140, 147, 162, 172 g/L, and 5 female blood donors: 123, 140, 137, 132, 127 g/L, is there a reason to believe that the Hb level is higher in men than in women?

:::{.callout-note collapse="true"}

# Solution

Let $X_m$ denote the Hb value in men and $X_w$ the Hb value in women.

$$H_0: \mu_m = \mu_w$$
$$H_1: \mu_m > \mu_w$$

Use the significance level $\alpha=0.05$.

Use the test statistic $T = \frac{\bar X_m - \bar X_w}{SE}$. If we assume equal variances we can use Student's t-test and compute SE using the pooled standard deviation. An alternative is to use Welch's t-test (unequal variances t-test) and compute $SE=\sqrt{\frac{s_m^2}{n_m} + \frac{s_w^2}{n_w}}$. The test statistic is t-distributed and the degrees of freedom are approximated using the Welch-Satterthwaite's equation, as implemented in `t.test`.

```{r, echo=TRUE}
x <- c(154, 140, 147, 162, 172)
y <- c(123, 140, 137, 132, 127)
## Perform Welch's t-test with unequal variances
tt <- t.test(x, y, alternative="greater")
tt
```

As the p-value $p=`r tt$p.value`$ is less than $\alpha$, we reject the null hypothesis and conclude that the data provide evidence that the mean Hb is higher in male blood donors than in female.

:::
:::

::: {#exr-Hb4}

Based on statistics from blodcentralen we learn that male Hb values (before donation) are normally distributed with mean 188 g/L and standard deviation 16 g/L. Using this new information and the following observed Hb values in five male donors after donation: 154, 140, 147, 162, 172 g/L, is there reason to believe that the mean Hb value for men after blood donation is lower than 188 g/L?

:::{.callout-note collapse="true"}

# Solution
$H_0: \mu = 188$ $H_1: \mu < 188$

Use $\alpha = 0.05$.

Under $H_0$, the Hb value for a single male donor is normally distributed with mean 188 g/L and standard deviation 16 g/L.

$X \sim N(188, 16^2)$

Since the sample size is 5, the sample mean is normally distributed with $\bar X \sim N\left(188, \frac{16^2}{5}\right),$ so the standard error of mean is $SE = \frac{16}{\sqrt{5}}$.


Use the test statistic

$Z = \frac{\bar X - 188}{16/\sqrt{5}} \sim N(0,1)$

Compute $z_{obs}$

```{r}
x <- c(154, 140, 147, 162, 172)
m <- mean(x)
SE <- 16/sqrt(5)
zobs <- (m - 188)/SE
```

Compute the p-value:
$$p=P(\bar X \leq x_{obs}) = P(Z \leq zobs).$$

```{r}
(p <- pnorm(zobs))
```

As $p < \alpha$, we reject $H_0$ and conclude that the data provide evidence that the mean Hb value for men after blood donation is lower than 188 g/L.

:::
:::

::: {#exr-mice12}

In order to study the effect of high-fat diet, 12 mice are fed normal diet (control group) and 12 mice are fed high-fat diet. After a couple of weeks, the mouse weights in grams are recorded:

```{r miceobs, echo=FALSE}
## 12 HF mice
xHF <- c(25, 30, 23, 18, 31, 24, 39, 26, 36, 29, 23, 32)
## 12 control mice
xN <- c(27, 25, 22, 23, 25, 37, 24, 26, 21, 26, 30, 24)
```

High-fat mice (g): `r xHF`

Normal diet mice (g): `r xN`

Does high-fat diet increase body weight in mice?

a)  Assume equal variances.
b)  Do not assume equal variances.

:::{.callout-note collapse="true"}

# Solution

a)

```{r ttesteqvar, echo=TRUE, eval=TRUE}
## Student's t-test with pooled variances
t.test(xHF, xN, var.equal=TRUE, alternative="greater")
```

b)  

```{r ttestuneqvar, echo=TRUE, eval=TRUE}
## Unequal variances with Welch approximation to the degrees of freedom (the default)
t.test(xHF, xN, var.equal=FALSE, alternative="greater")
```

:::
:::
