# Exercises: Hypothesis tests, resampling {.unnumbered}

::: {#exr-pollentest} 
# Pollen 

You believe that the proportion of Swedish students allergic to pollen is greater than 0.3 (the proportion allergic to pollen in Sweden). To test this you observe 20 people in a student group at BMC in Uppsala, 9 or them are allergic to pollen.

Is this reason to believe that the proportion of Swedish students allergic to pollen i greater than 0.3? Perform a hypothesis test to answer the question.

Can you identify any problems with this study setup?

::: {.panel-tabset}
# ​

# Hint

xx

# Solution

$H_0: \pi=0.3$ $H_1: \pi>0.3$

Set the significance level to $\alpha=0.05$.

Test statistic, $X$, the number of allergic people in a sample of size 20.

$x_{obs} = 9$

Simulate null distribution

```{r histpnull, eval=TRUE}
urn <- c(1,1,1,0,0,0,0,0,0,0)
xnull <- replicate(10000, sum(sample(urn, 20, replace=TRUE)))
hist(xnull)
```

Compute p-value, i.e. if null is true what is the probability to observe $x_{obs}$ or higher?

```{r, echo=TRUE}
xobs <- 9
(x <- mean(xnull>=xobs))
```

As $p>\alpha$ we will accept the null hypothesis, i.e. there is no reason to belive that the students are more allergic than the general Swedish population.

Problems with the study: Discuss in your group! Is it reasonable to select 20 students at BMC to answer a question about all students in Sweden?
:::
:::

::: {#exr-diet}

# Diet

A diet study aims to study how the hemoglobin (Hb) levels in blood are affected by an iron-rich diet consisting of tofu, soybeans, broccoli, lentils and peas. To perform the study the dietician has recruited 40 male participants, who are randomly assigned to the iron-rich diet or control group (no change in participants diet), 20 participant in each group.

The observed Hb levels (in g/L);

```{r irondiet, echo=TRUE}
ctrl <- c(197, 186, 157, 170, 193, 188, 175, 186, 177, 191, 168, 193, 191, 189, 188, 192, 179, 186, 197, 203)
iron <- c(187, 218, 196, 210, 206, 178, 181, 193, 172, 202, 169, 221, 183, 222, 185, 174, 192, 192, 162, 211)
```

Perform a hypothesis test to investigate if the Hb level is affected (increased or decreased) by the iron-rich diet.

::: {.panel-tabset}
# ​

# Solution

Define $H_0$ and $H_1$

$H_0: \mu_{diet} = \mu_{ctrl}$ No difference in mean iron level between control group and iron rich group

$H_1: \mu_{diet} \neq \mu_{ctrl}$

Will use the significance level, $\alpha=0.05$

Select test statistic $D = \bar X_d - \bar X_c$, where $\bar X_c$ is the mean Hb level in a control group of 20 people and $\bar X_d$ is the mean Hb level in a diet group of 20 people.

The observed value; $d_{obs}$

```{r, echo=TRUE}
mdiet <- mean(iron)
mctrl <- mean(ctrl)
(dobs <- mdiet - mctrl)
```

Compute null distribution using permutation.

```{r histironnull, echo=TRUE}
## Under null all observations are equivalent
allobs <- c(iron, ctrl)
dnull <- replicate(10000, {
  ##Permute the 40 observations and assign the 20 first to the iron group
  x <- sample(allobs)
  d <- mean(x[1:20]) - mean(x[21:40])
})
hist(dnull)
```

Compute p-value;

```{r, echo=TRUE}
(p <- mean(abs(dnull) >= abs(dobs)))
```

As $p>\alpha$, the null hypothesis is accepted, i.e. there is no reason to believe that the iron-rich diet affects the blood Hb level.

:::
:::
