---
editor: source
---

# Exercises: Hypothesis tests, parametric {.unnumbered}

::: {#exr-Hb1}
The hemoglobin value (Hb) in women is  on average 140 g/L. You observe the following Hb values in a set of five male blood donors: 154, 140, 147, 162, 172. Assume that Hb is normally distributed. Is there a reason to believe that the mean Hb value in men differ from that in women? 

::: {.panel-tabset}
# ​

# Solution
Let $X$ denote the Hb value in g/L for male blood donors.

$$H_0: \bar X = 140$$ $$H_1: \bar X \neq 140$$

Use the significance level $\alpha=0.05$.

Use the test statistic $T = \frac{\bar X - \mu_0}{SE},$ where $\mu_0$ is the population mean if $H_0$ is true, i.e. 140 and $SE$ is the standard error of mean, $SE=\frac{\sigma}{\sqrt{n}} \approx \frac{s}{\sqrt{n}}$.

Compute $t_{obs} = \frac{m-140}{s/\sqrt{n}}$

```{r Hb1, echo=TRUE}
x <- c(154, 140, 147, 162, 172)
## sample mean
m <- mean(x)
## standard error of mean
SE <- sd(x)/sqrt(5)
## Observed value of test statistic
(tobs <- (m-140)/SE)
```

Compute the p-value, $P(|T|>|t_{obs}|) = P(T>|t_{obs}|) + P(T<-|t_{obs}|)$

```{r, echo=TRUE}
## P(T>tobs)
pt(tobs, df=4, lower.tail=FALSE)
## p = P(t>tobs) + P(t<-tobs) = 2 * P(t>tobs)
(p <- 2*pt(tobs, df=4, lower.tail=FALSE))
```

As $p > \alpha$ the null hypothesis is accepted and we conclude that there is **not** reason to believe that the Hb values for men differ from that of women.

Note that the same can be achieved uing the function `t.test`;

```{r}
t.test(x, mu=140)
```

:::
:::

::: {#exr-Hb2}
The hemoglobin value (Hb) in men is on average 188 g/L. The Hb values in @exr-Hb1 were actually measured after the men had donated blood. Is there a reason to believe that the mean Hb level for men after blood donation is less than 188 g/L?

::: {.panel-tabset}
# ​

# Solution

$H_0: \bar X = 188$ $H_1: \bar X < 188$

Use the significance level $\alpha=0.5$.

Use the test statistic $T = \frac{\bar X - \mu_0}{SE},$ where $\mu_0=188$.

Compute $t_{obs} = \frac{m-188}{s/\sqrt{n}}$

```{r, echo=TRUE}
## Observed value of test statistic
(tobs <- (m-188)/SE)
```

Compute the p-value, $P(T<t_{obs})$

```{r, echo=TRUE}
## p=P(T<tobs)
(p <- pt(tobs, df=4, lower.tail=TRUE))
```

As $p < \alpha$ the null hypothesis is rejected and we can conclude that there is reason to believe that the Hb values after blood donation is less than 188 g/L.

Note that the same can be achieved uing the function `t.test`;

```{r echo=TRUE}
t.test(x, mu=188, alternative="less")
```

:::
:::


::: {#exr-Hb3}
By observing the Hb values in 5 male blood donors; 154, 140, 147, 162, 172 g/L, and 5 female blood donors: 123, 140, 137, 132, 127 g/L, is there a reason to believe that the Hb level is higher in men than in women?

::: {.panel-tabset}
# ​

# Solution

Let $X_m$ denote the Hb value in men and $X_w$ the Hb value in women.

$H_0: \bar X_m = \bar X_w$ $H_1: \bar X_m > \bar X_w$

Use the significance level $\alpha=0.5$.

Use the test statistic $T = \frac{\bar X_m - \bar X_w}{SE}$. If we assume equal variances we can use Student's t-test and compute SE as the pooled standard deviation. An alternative is to use Welch t-test (unequal variances t-test) and compute $SE=\sqrt{\frac{s_m^2}{n_m} + \frac{s_w^2}{n_w}}$, the test statistic is t-distributed and the degrees of freedom can be approximated using Welch-Satterthwaite's equation, as implemented in `t.test`.

```{r, echo=TRUE}
x <- c(154, 140, 147, 162, 172)
y <- c(123, 140, 137, 132, 127)
## Perform t-test with unequal variances
t.test(x, y, alternative="greater")
```

As the p-value $p=`r t.test(x, y, alternative="greater")$p.value`$ is $<\alpha$, we reject the null hypothesis and conclude that there is reason to belive that Hb on average is higher in male blood donors than in female.

:::
:::

::: {#exr-Hb4}
Based on statistics from blodcentralen we learn that male Hb values (before donation) is normally distributed with mean 188 g/L and standard deviation 16 g/L. Using this new information and the following observed Hb values in five male donors after donation; 154, 140, 147, 162, 172 g/L, is there reason to believe that the mean Hb value for men after blood donation is lower than 188 g/L?

::: {.panel-tabset}
# ​

# Solution
$H_0: \bar X = 188$ $H_1: \bar X < 188$

Use $\alpha = 0.05$

Under $H_0$;

$X \sim N(188, 16)$

$\bar X \sim N(188, 16/\sqrt{5})$

Use the test statistic

$Z = \frac{\bar X - 188}{16/\sqrt{5}} \sim N(0,1)$

Compute $z_{obs}$

```{r}
x <- c(154, 140, 147, 162, 172)
m <- mean(x)
SE <- 16/sqrt(5)
zobs <- (m - 188)/SE
```

Compute the p-value \$P(\bar X \leq 188) = P(Z \leq zobs) = \$

```{r}
pnorm(zobs)
```

As $p<<\alpha$, reject the $H_0$, there is reason to believe that the mean Hb value for men after blood donation is lower than 188 g/L.

:::
:::

::: {#exr-mice12}
In order to study the effect of high-fat diet 12 mice are fed normal diet (control group) and 12 mice are fed high-fat diet. After a couple of weeks the mouse weights in gram are recorded;

```{r miceobs, echo=FALSE}
## 12 HF mice
xHF <- c(25, 30, 23, 18, 31, 24, 39, 26, 36, 29, 23, 32)
## 12 control mice
xN <- c(27, 25, 22, 23, 25, 37, 24, 26, 21, 26, 30, 24)
```

High-fat mice (g): `r xHF` Normal diet mice (g): `r xN`

Does high fat diet increase body weight in mice?

a)  Assume equal variances.
b)  Don't assume equal variances.

::: {.panel-tabset}
# ​

# Solution

a)

```{r ttesteqvar, echo=TRUE, eval=TRUE}
## Student's t-test with pooled variances
t.test(xHF, xN, var.equal=TRUE, alternative="greater")
```

b)  

```{r ttestuneqvar, echo=TRUE, eval=TRUE}
## Unequal variances with Welch approximation to the degrees of freedom (the default)
t.test(xHF, xN, var.equal=FALSE, alternative="greater")
```

:::
:::

<!-- Exercise, known $\sigma$ -->

<!-- If the means of the populations are unknown, but the standard variations are known -->

<!-- Let's assume that both mouse body weights in control and treatment groups are independent and normally distributed, with unknown mean, but known standard deviations, $\sigma_1=3.4$ and $\sigma_2=5.1$. -->

<!-- ```{r echo=F, eval=FALSE} --> <!-- ## Our observed value --> <!-- dobs --> <!-- ## The p-value --> <!-- 1-pnorm(dobs, mean=0, sd=sqrt(5.1^2/12 + 3.4^2/12)) --> <!-- 1-pnorm((dobs-0)/sqrt(5.1^2/12 + 3.4^2/12)) --> <!-- pnorm(dobs, mean=0, sd=sqrt(5.1^2/12 + 3.4^2/12), lower.tail=FALSE) --> <!-- pnorm((dobs-0)/sqrt(5.1^2/12 + 3.4^2/12), lower.tail=FALSE) --> <!-- ``` -->
