---
title: "Welcome"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path="session-intro-files/figures/")
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(cache.path = "tmp")
knitr::opts_chunk$set(cache = TRUE)
```

### Biostatistics Essentials: a blackboard approach
- 21st - 23rd, May, 2019
- SciLifeLab, Uppsala University, BMC, Husargatan 3, E10:1308

<br/><br/>

### Getting to know each other

##### Teaching team
- [Eva Freyhult][eva], PhD, NBIS, SciLifeLab, Uppsala University
- [Olga Dethlefsen][olga], PhD, NBIS, SciLifeLab, University of Stockholm
- [Payam Emami][payam], PhD, NBIS, SciLifeLab, University of Stockholm
- [Paulo Czarnewski][paulo], PhD, NBIS, SciLifeLab, University of Stockholm
- [Marcin Kierczak][marcin], PhD, NBIS, SciLifeLab, Uppsala University
- [Bengt Sennblad][bengt], PhD, NBIS, SciLifeLab, Uppsala University

<br/><br/>

#### What about you? 

<br/><br/>
```{r, echo=F, message=FALSE, warning=FALSE}

rm(list=ls())
library(dplyr)

data.inp <- read.delim("Background.txt")


```


```{r, echo=F, message=F, warning=F, fig.align="center"}

data.stats <- data.inp[, 4:ncol(data.inp)]
colnames(data.stats) <- gsub("How.do.you.feel.about.using.the.below.methods...", "", colnames(data.stats))

data.stats <- as.matrix(data.stats)
data.statsNumeric <- matrix(data=NA, nrow=nrow(data.stats), ncol=ncol(data.stats))
colnames(data.statsNumeric) <- colnames(data.stats)
for (i in 1:nrow(data.statsNumeric)){
  
  for (j in 1:ncol(data.statsNumeric)){
    
    #print(i,j)

    if (data.stats[i,j]=="Newbie"){
      data.statsNumeric[i,j]=0}

    if (data.stats[i,j]=="Intermediate"){
      data.statsNumeric[i,j]=1}

    if (data.stats[i,j]=="Expert"){
      data.statsNumeric[i,j]=2}
    
  }
}


library(ggplot2)
library(gplots)
library(RColorBrewer)
library(reshape)

colornames <- brewer.pal(n=3,name="Greens")

data.statsLong<- melt(data.stats)
idx.na <- which(data.statsLong$value=="")
data.statsLong <- data.statsLong[-idx.na,]

data.statsLong$value <- factor(data.statsLong$value, levels=c("Newbie", "Intermediate", "Expert"))

g.stats <- ggplot(data.statsLong, aes(data.statsLong$X2, fill=value))+geom_bar() + coord_flip() + xlab("") + scale_fill_manual(values=c("springgreen", "springgreen3", "springgreen4")) +  theme(legend.position="top") + theme(legend.title = element_blank()) + ylab("") + theme(legend.spacing.x = unit(0.2, 'cm')) + theme(panel.background = element_rect(fill = "white"))  + ggtitle("Statistics experience")

g.stats


```

<br/><br/>
```{r, echo=F, fig.width=4, fig.height=4, fig.align="center"}

colornames <- brewer.pal(n=4, "Set1")

data.inp$R.experience <- factor(data.inp$R.experience, levels = c("None", "Beginner", "Intermediate", "Advanced"))
  
g.R <- ggplot(data.inp, aes(data.inp$R.experience, fill=data.inp$R.experience))+geom_bar(width = 0.5) + xlab("") + theme(legend.position="top") + theme(legend.title = element_blank()) + ylab("") + theme(legend.spacing.x = unit(0.2, 'cm')) + theme(panel.background = element_rect(fill = "white")) + scale_fill_manual(values=colornames) + ggtitle("R experience")

g.R

```





[eva]: https://nbis.se/about/staff/eva-freyhult/
[olga]: https://nbis.se/about/staff/olga-dethlefsen/
[payam]: https://nbis.se/about/staff/payam-emami/
[paulo]: https://nbis.se/about/staff/paulo-czarnewski/
[marcin]: https://nbis.se/about/staff/marcin-kierczak/
[bengt]: https://nbis.se/about/staff/bengt-sennblad/




<br/><br/>

### Teaching philosophy aka. "blackboard approach"

_In this course we focus on active learning approach. The course participants are expected to do some pre-course reading and exercises, corresponding up to 16h studying. The education consists of teaching blocks alternating between mini-lectures, group discussions, live coding sessions etc. A special focus is put on understanding basic concepts by solving problems with pen and paper. We believe in less-is-more principle, so be ready to say no to digital exhaustion to truly learn something with our modern “blackboard approach”_


- Focus on active learning
- To develop comptetency and capture attention
- Drawing from education theory (behavourist, neuroscience, cognitivism, social learning, humanistic principles)


<br/><br/>

### Course website:
https://nbisweden.github.io/workshop-biostatistics/


