---
title: "Welcome"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path="session-intro-files/figures/")
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(cache.path = "tmp")
knitr::opts_chunk$set(cache = FALSE)


```


### Introduction to biostatistics and machine learning
- 16th - 20th, November, 2020
- online

<br/><br/>

### Getting to know each other
<br/>

##### Teaching team
- [Olga Dethlefsen][olga] 
- [Eva Freyhult][eva] 
- [Bengt Sennblad][bengt] 
- [Payam Emami][payam] 
- [Mun-Gwan Hong](mungwan) 
- [Nima Rafati][nima] 
- [Lucile Soler][lucile] 

[eva]: https://nbis.se/about/staff/eva-freyhult/
[olga]: https://nbis.se/about/staff/olga-dethlefsen/
[bengt]: https://nbis.se/about/staff/bengt-sennblad/
[payam]: https://nbis.se/about/staff/payam-emami/
[mungwan]: https://nbis.se/about/staff/mungwan-hong/
[nima]: https://nbis.se/about/staff/nima-rafati/
[lucile]: https://nbis.se/about/staff/lucile-soler/


<br/><br/>

#### What about you? 

<br/><br/>
```{r, echo=F, message=FALSE, warning=FALSE}

rm(list=ls())
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
library(scales)

data.inp <- read.delim("background.csv", sep=";")
data.inp <- data.inp[1:22, ]
#head(data.inp)

```


```{r, echo=F, message=F, warning=F, fig.align="center"}

colornames <- brewer.pal(n=4, "Set1")

data.inp$R <- factor(data.inp$R.experience, levels = c("None", "Beginner", "Intermediate", "Advanced"))
data.inp$Position <- factor(data.inp$Position)
data.inp$University <- factor(data.inp$University)

bar.width <- 0.7
g1 <- ggplot(data.inp, aes(y=Position)) + geom_bar(fill=colornames[2], alpha=0.5) + ylab("") + theme_classic2() + scale_x_continuous(breaks= pretty_breaks())
g2 <- ggplot(data.inp, aes(y=University)) + geom_bar(fill=colornames[3], alpha=0.5) + ylab("") + theme_classic2()  + scale_x_continuous(breaks= pretty_breaks())
g3 <- ggplot(data.inp, aes(y=R)) + geom_bar(fill=colornames[4], alpha=0.5) + ylab("") + ylab("")+ theme_classic2() + scale_x_continuous(breaks= pretty_breaks())
g4 <- ggplot(data.inp, aes(y=R, fill=Position)) + geom_bar(alpha = 0.5) + ylab("")

ggarrange(g2, g1, g3, ncol=3)

```


<br/><br/>

### Teaching approach
- we try to focus on active learning with us talking less and you doing more
- we try to focus on fundamentals, thinking that if you get the basics right you will have no problems reading up the more advanced things
- it is the first time we run things online and we believe is self-organization, i.e. will let you choose how to work at least half of the time ([Online Classroom](https://nbisweden.github.io/workshop-mlbiostatistics/classroom))
- **be brave**: help us to build a good leaning environment by asking questions in public on Zoom and Zulip. We all benefit from that. 

<br/><br/>

### Educational research project
- a side project
- not much is really known what teaching approaches really do work best during short-term courses
- one thing that is known is that we, humans, need some time to process and learn new information
- can we somehow learn better when using different methods?
- A big thank you if you've sent your informed consent form. Not too late if you haven't. Either way, the course is the same for everyone! 



<br/><br/>

### Course website:
https://nbisweden.github.io/workshop-biostatistics/




