---
title: "Welcome"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path="session-intro-files/figures/")
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(cache.path = "tmp")
knitr::opts_chunk$set(cache = FALSE)

```


### Introduction to biostatistics and machine learning
- 04th - 08th, October, 2021
- online

<br/><br/>


### About us

- [Olga Dethlefsen][olga] 
- [Eva Freyhult][eva] 
- [Bengt Sennblad][bengt] 
- [Payam Emami][payam] 

<br/>

- [Mun-Gwan Hong][mungwan]
- [Per Unneberg][per]
- [Julie Lorent][julie]
- [Vincent van Hoef][vincent]


[eva]: https://nbis.se/about/staff/eva-freyhult/
[olga]: https://nbis.se/about/staff/olga-dethlefsen/
[bengt]: https://nbis.se/about/staff/bengt-sennblad/
[payam]: https://nbis.se/about/staff/payam-emami/

[mungwan]: https://nbis.se/about/staff/mungwan-hong/
[per]: https://nbis.se/about/staff/per-unneberg/
[julie]: https://nbis.se/about/staff/julie-lorent/
[vincent]: https://nbis.se/about/staff/vincent-vanhoef/

<br/>

- NBIS: National Bioinformatics Infrastructure Sweden https://www.scilifelab.se/units/nbis/
- SciLifeLab Bioinformatics Platform



<br/><br/>

#### What about you? 

<br/><br/>
```{r, echo=F, message=FALSE, warning=FALSE}

rm(list=ls())
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
library(scales)
library(tidyverse)

data.input <- read_csv("background-2021.csv")

data.inp <- data.input %>%
  mutate(R = factor(R, levels=c("Beginner", "Intermediate", "Advanced")))

#head(data.inp)

```


```{r, echo=F, message=F, warning=F, fig.align="center", fig.keep="all", fig.height=4.5}

colornames <- brewer.pal(n=4, "Set1")

data.inp$R <- factor(data.inp$R, levels = c("None", "Beginner", "Intermediate", "Advanced"))
data.inp$Position <- factor(data.inp$Position)
data.inp$University <- factor(data.inp$University)

bar.width <- 0.7
g1 <- ggplot(data.inp, aes(x=Position, group = 1)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  geom_text(aes(label = scales::percent(..prop..), y= ..prop.. ), stat= "count", vjust = -.5) +
  xlab("") + 
  ylab("%") + 
  theme_classic2() + 
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "None") + 
  ggtitle("Position")


g2 <- ggplot(data.inp, aes(x=University, group = 1)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  geom_text(aes(label = scales::percent(..prop..), y= ..prop.. ), stat= "count", vjust = -.5) +
  xlab("") + 
  ylab("%") + 
  theme_classic2() + 
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "None") + 
  ggtitle("University")

g3 <- ggplot(data.inp, aes(x=R, group = 1)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  geom_text(aes(label = scales::percent(..prop..), y= ..prop.. ), stat= "count", vjust = -.5) +
  xlab("") + 
  ylab("%") + 
  theme_classic2() + 
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "None") + 
  ggtitle("R experience")


ggarrange(g1, g3, ncol=2)
g2


```


<br/><br/>

### Teaching approach
- we try to focus on fundamentals, thinking that if you get the basics right you will have no problems reading up more advanced things
- we try to focus on active learning with us talking less and you doing more
- **help us to build a stimulating learning environment** by asking questions on Zoom and Canvas

<br/><br/>

### Online classroom
- we believe that you know how to study best ([Online Classroom](https://nbisweden.github.io/workshop-mlbiostatistics/classroom))

<br/><br/>

### Canvas Demo
- Schedule
- Modules 
- Chat
- Discussion
- Files 
- Announcements

*please change your name under the Profile*

<br/><br/>

### Certificate requirements
- active participation during the week
- online presence in live streams and group sessions
- completing "Daily challenge" quiz: opens daily at 15.00, closes at 09:00 the following day


<br/><br/>

### Course website
- https://uppsala.instructure.com/courses/51998 (current)
- https://nbisweden.github.io/workshop-biostatistics/ (generic)




