---
title: "Introduction to biostatistics and machine learning"
author: "Welcome"
format:
  revealjs: 
    theme: [default, custom.scss]
    incremental: false   
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path="session-intro-files/figures/")
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(cache.path = "tmp")
knitr::opts_chunk$set(cache = FALSE)

```

## Introduction to biostatistics and machine learning

-   12th - 16th, September, 2022
-   BMC, Husargatan 3, Uppsala
-   Room: Trippelrummet: E10:1307/8/9

<br/><br/>

## About us

-   [Olga Dethlefsen](https://nbis.se/about/staff/olga-dethlefsen/)
-   [Eva Freyhult](https://nbis.se/about/staff/eva-freyhult/)
-   [Bengt Sennblad](https://nbis.se/about/staff/bengt-sennblad/)
-   [Payam Emami](https://nbis.se/about/staff/payam-emami/)
-   [Mun-Gwan Hong](https://nbis.se/about/staff/mungwan-hong/)
-   [Julie Lorent](https://nbis.se/about/staff/julie-lorent/)

<br/>

-   NBIS: National Bioinformatics Infrastructure Sweden
-   SciLifeLab Bioinformatics Platform
-   <https://www.scilifelab.se/units/nbis/>

<br/><br/>

<!-- ## What about you? -->

<!-- Names and teams  -->

## What about you?

<br/><br/>

```{r, echo=F, message=FALSE, warning=FALSE}

rm(list=ls())
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
library(scales)
library(tidyverse)

data.input <- read_csv("background-2021.csv")

data.inp <- data.input %>%
  mutate(R = factor(R, levels=c("Beginner", "Intermediate", "Advanced")))

```

```{r, echo=F, message=F, warning=F, fig.align="center", fig.keep="all", fig.height=4.5}

colornames <- brewer.pal(n=4, "Set1")

data.inp$R <- factor(data.inp$R, levels = c("None", "Beginner", "Intermediate", "Advanced"))
data.inp$Position <- factor(data.inp$Position)
data.inp$University <- factor(data.inp$University)

bar.width <- 0.7
g1 <- ggplot(data.inp, aes(x=Position, group = 1)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  geom_text(aes(label = scales::percent(..prop..), y= ..prop.. ), stat= "count", vjust = -.5) +
  xlab("") + 
  ylab("%") + 
  theme_classic2() + 
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "None") + 
  ggtitle("Position")


g2 <- ggplot(data.inp, aes(x=University, group = 1)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  geom_text(aes(label = scales::percent(..prop..), y= ..prop.. ), stat= "count", vjust = -.5) +
  xlab("") + 
  ylab("%") + 
  theme_classic2() + 
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "None") + 
  ggtitle("University")

g3 <- ggplot(data.inp, aes(x=R, group = 1)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  geom_text(aes(label = scales::percent(..prop..), y= ..prop.. ), stat= "count", vjust = -.5) +
  xlab("") + 
  ylab("%") + 
  theme_classic2() + 
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "None") + 
  ggtitle("R experience")

ggarrange(g1, g3, ncol=2)

```

## What about you?

<br/><br/>

```{r, echo=F}
print(g2)
```

## Teaching approach

::: incremental
-   we try to focus on fundamentals, thinking that if you get the basics right you will have no problems reading up more advanced things
-   we try to focus on active learning with us talking less and you doing more
-   **help us to build a stimulating learning environment** by actively participating. We love when you ask questions!
:::

<!-- ## Online classroom -->
<!-- - we believe that you know how to study best (Online Classroom) -->
<!-- <br/><br/> -->

## Canvas Demo
-   Schedule
-   Modules
-   Chat
-   Discussion
-   Files
-   Announcements
-   Quiz

*please change your name under the Profile*

<br/><br/>

## Certificate requirements

::: incremental
-   presence in all sessions during the week
    -   we may allow skipping up to 4h during the week
-   active participation during the week
-   completing "Daily challenge" quiz
    - opens daily at 15.00 
    - closes at 09:00 the following day
:::

. . .

*Note that we are not able to provide any formal university credits (högskolepoäng). Many universities, however, recognize the attendance in our courses, and award 1.5 HPs, corresponding to 40h of studying. It is up to participants to clarify and arrange credit transfer with the relevant university department.*



<br/><br/>

## Course website

-   <https://uppsala.instructure.com/courses/70535> (current)
-   <https://nbisweden.github.io/workshop-mlbiostatistics/> (generic)

## Questions? 
