---
title: "Introduction to biostatistics and machine learning"
author: "Welcome"
format:
  revealjs: 
    theme: [default, custom.scss]
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path="session-intro-files/figures/")
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(cache.path = "tmp")
knitr::opts_chunk$set(cache = FALSE)

```

## Introduction to biostatistics and machine learning

-   12th - 16th, September, 2022
-   online

<br/><br/>

## About us

-   [Olga Dethlefsen](https://nbis.se/about/staff/olga-dethlefsen/)
-   [Eva Freyhult](https://nbis.se/about/staff/eva-freyhult/)
-   [Bengt Sennblad](https://nbis.se/about/staff/bengt-sennblad/)
-   [Payam Emami](https://nbis.se/about/staff/payam-emami/)
-   [Mun-Gwan Hong](https://nbis.se/about/staff/mungwan-hong/)
-   [Julie Lorent](https://nbis.se/about/staff/julie-lorent/)

<br/>

-   NBIS: National Bioinformatics Infrastructure Sweden https://www.scilifelab.se/units/nbis/
-   SciLifeLab Bioinformatics Platform

<br/><br/>

<!-- ## What about you? -->
<!-- Names and teams  -->

## What about you?
<br/><br/>
```{r, echo=F, message=FALSE, warning=FALSE}

rm(list=ls())
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
library(scales)
library(tidyverse)

data.input <- read_csv("background-2021.csv")

data.inp <- data.input %>%
  mutate(R = factor(R, levels=c("Beginner", "Intermediate", "Advanced")))

```

```{r, echo=F, message=F, warning=F, fig.align="center", fig.keep="all", fig.height=4.5}

colornames <- brewer.pal(n=4, "Set1")

data.inp$R <- factor(data.inp$R, levels = c("None", "Beginner", "Intermediate", "Advanced"))
data.inp$Position <- factor(data.inp$Position)
data.inp$University <- factor(data.inp$University)

bar.width <- 0.7
g1 <- ggplot(data.inp, aes(x=Position, group = 1)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  geom_text(aes(label = scales::percent(..prop..), y= ..prop.. ), stat= "count", vjust = -.5) +
  xlab("") + 
  ylab("%") + 
  theme_classic2() + 
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "None") + 
  ggtitle("Position")


g2 <- ggplot(data.inp, aes(x=University, group = 1)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  geom_text(aes(label = scales::percent(..prop..), y= ..prop.. ), stat= "count", vjust = -.5) +
  xlab("") + 
  ylab("%") + 
  theme_classic2() + 
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "None") + 
  ggtitle("University")

g3 <- ggplot(data.inp, aes(x=R, group = 1)) + 
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") + 
  geom_text(aes(label = scales::percent(..prop..), y= ..prop.. ), stat= "count", vjust = -.5) +
  xlab("") + 
  ylab("%") + 
  theme_classic2() + 
  scale_y_continuous(labels = scales::percent) + 
  theme(legend.position = "None") + 
  ggtitle("R experience")

ggarrange(g1, g3, ncol=2)

```

## What about you?
<br/><br/>
```{r, echo=F}
print(g2)
```

## Teaching approach

-   we try to focus on fundamentals, thinking that if you get the basics right you will have no problems reading up more advanced things
-   we try to focus on active learning with us talking less and you doing more
-   **help us to build a stimulating learning environment** by actively participating. We love when you ask questions!

<br/><br/>

<!-- ### Online classroom -->

<!-- - we believe that you know how to study best (Online Classroom) -->

<br/><br/>

## Canvas Demo

-   Schedule
-   Modules
-   Chat
-   Discussion
-   Files
-   Announcements
-   Quiz

*please change your name under the Profile*

<br/><br/>

## Certificate requirements

-   presence in all sessions during the week
-   we may allow skipping up to 4h during the week
-   active participation during the week
-   completing "Daily challenge" quiz: opens daily at 15.00, closes at 09:00 the following day

<br/><br/>

## Course website

-   [https://uppsala.instructure.com/courses/70535](https://uppsala.instructure.com/courses/70535) (current)
-   [https://nbisweden.github.io/workshop-mlbiostatistics/](https://nbisweden.github.io/workshop-mlbiostatistics/) (generic)
