---
title: "Welcome"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path="session-intro-files/figures/")
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(cache.path = "tmp")
knitr::opts_chunk$set(cache = FALSE)


```


### Introduction to biostatistics and machine learning
- 04th - 08th, October, 2021
- online

<br/><br/>

### Getting to know each other
<br/>

##### Teaching team
- [Olga Dethlefsen][olga] 
- [Eva Freyhult][eva] 
- [Bengt Sennblad][bengt] 
- [Payam Emami][payam] 


[eva]: https://nbis.se/about/staff/eva-freyhult/
[olga]: https://nbis.se/about/staff/olga-dethlefsen/
[bengt]: https://nbis.se/about/staff/bengt-sennblad/
[payam]: https://nbis.se/about/staff/payam-emami/


<br/><br/>

#### What about you? 

<br/><br/>
```{r, echo=F, message=FALSE, warning=FALSE}

rm(list=ls())
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
library(scales)

data.inp <- read.delim("background.csv", sep=";")
data.inp <- data.inp[1:22, ]
#head(data.inp)

```


```{r, echo=F, message=F, warning=F, fig.align="center"}

colornames <- brewer.pal(n=4, "Set1")

data.inp$R <- factor(data.inp$R.experience, levels = c("None", "Beginner", "Intermediate", "Advanced"))
data.inp$Position <- factor(data.inp$Position)
data.inp$University <- factor(data.inp$University)

bar.width <- 0.7
g1 <- ggplot(data.inp, aes(y=Position)) + geom_bar(fill=colornames[2], alpha=0.5) + ylab("") + theme_classic2() + scale_x_continuous(breaks= pretty_breaks())
g2 <- ggplot(data.inp, aes(y=University)) + geom_bar(fill=colornames[3], alpha=0.5) + ylab("") + theme_classic2()  + scale_x_continuous(breaks= pretty_breaks())
g3 <- ggplot(data.inp, aes(y=R)) + geom_bar(fill=colornames[4], alpha=0.5) + ylab("") + ylab("")+ theme_classic2() + scale_x_continuous(breaks= pretty_breaks())
g4 <- ggplot(data.inp, aes(y=R, fill=Position)) + geom_bar(alpha = 0.5) + ylab("")

ggarrange(g2, g1, g3, ncol=3)

```


<br/><br/>

### Teaching approach
- we try to focus on active learning with us talking less and you doing more
- we try to focus on fundamentals, thinking that if you get the basics right you will have no problems reading up the more advanced things
- we believe that you know how to study best ([Online Classroom](https://nbisweden.github.io/workshop-mlbiostatistics/classroom))
- **be brave**: help us to build a **stimulating learning environment** by asking questions in public on Zoom and Zulip. 

<br/><br/>

### Certificate requirements
- active participation during the week
- online presence in live streams and group sessions
- completing "Daily challenge" session: opens daily at 16.00
- please email us your address that we can send the certificate to

<br/><br/>

### Course website:
https://uppsala.instructure.com/courses/51998

https://nbisweden.github.io/workshop-biostatistics/




