---
title: "Machine Learning for Life Sciences, 3hp"
subtitle: "Online session 2"
author: "Olga Dethlefsen, PhD <br> Payam Emami, PhD"
format: 
  revealjs:
    slide-number: true
    view-distance: 10
    theme: [default, styles.scss]
    chalkboard: 
      buttons: true
    footer: "Machine Learning for Life Sciences (ML4LS)"
  html:
    code-fold: false
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
#| message: false
#| warning: false

# load libraries
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(rmarkdown)
library(ggbeeswarm)
library(gridExtra)
library(scales)
library(ggthemes)
library(faraway)
library(GGally)
library(glmnet)
library(latex2exp)
library(splitTools)
library(ggeffects)
library(pROC)
library(patchwork)

```


## Session plan
<br>

. . .


1. Introduction to project work and demos
2. A mini self-study: explore the demo projects
3. Team work: discuss and outline your own project


# Introduction to project work and demos

## Data science lifecycle
*From research question to biological insight*

```{dot}
digraph workflow {

  graph [rankdir=LR]
  node [shape=box, style="rounded,filled", fillcolor="#DCEEFF"]

  define [label="Define problem"]
  collect [label="Collect data"]
  clean [label="Clean & preprocess data"]
  explore [label="Explore data"]

  transform [label="Feature transformation\n& engineering"]

  inference [label="Inferential statistics"]
  supervised [label="Supervised learning \n(prediction)"]
  unsupervised [label="Unsupervised learning \n(structure discovery)"]

  evaluate [label="Evaluate & interpret"]
  communicate [label="Communicate results"]

  define -> collect -> clean -> explore -> transform

  transform -> inference
  transform -> supervised
  transform -> unsupervised

  inference -> evaluate
  supervised -> evaluate
  unsupervised -> evaluate

  evaluate -> communicate
}
```

## Data science lifecycle
*More realistic case*

```{dot}
digraph workflow {

  graph [rankdir=LR]
  node [shape=box, style="rounded,filled", fillcolor="#DCEEFF"]

  define [label="Define problem"]
  collect [label="Collect data"]
  clean [label="Clean & preprocess data"]
  explore [label="Explore data"]

  transform [label="Feature transformation\n& engineering"]

  inference [label="Inferential statistics"]
  supervised [label="Supervised learning \n(prediction)"]
  unsupervised [label="Unsupervised learning \n(structure discovery)"]

  evaluate [label="Evaluate & interpret"]
  communicate [label="Communicate results"]

  define -> collect -> clean -> explore -> transform

  transform -> inference
  transform -> supervised
  transform -> unsupervised

  inference -> evaluate
  supervised -> evaluate
  unsupervised -> evaluate

  evaluate -> communicate

  collect -> define
  clean -> collect
  clean -> define
  explore -> clean
  transform -> clean
  inference -> define
  inference -> collect
  inference -> clean
  inference -> explore
  supervised -> collect
  supervised -> define
  unsupervised -> collect
  unsupervised -> explore
  evaluate -> define
  evaluate -> collect
  evaluate -> explore
  communicate -> define

}
```

## Data science lifecycle
*Let's focus on bigger picture today*

```{dot}
digraph workflow {

  graph [rankdir=LR]
  node [shape=box, style="rounded,filled", fillcolor="gold"]

  define [label="Define problem"]
  collect [label="Collect data"]
  clean [label="Clean & preprocess data"]
  explore [label="Explore data"]

  transform [label="Feature transformation\n& engineering"]

  inference [label="Inferential statistics", fillcolor="#DCEEFF"]
  supervised [label="Supervised learning \n(prediction)"]
  unsupervised [label="Unsupervised learning \n(structure discovery)"]

  evaluate [label="Evaluate & interpret"]
  communicate [label="Communicate results"]

  define -> collect -> clean -> explore -> transform

  transform -> inference
  transform -> supervised
  transform -> unsupervised

  inference -> evaluate
  supervised -> evaluate
  unsupervised -> evaluate

  evaluate -> communicate
}
```

## Example data
*Diabetes data*

<br>

::::: columns
::: {.column width="50%"}
-   403 participants were interviewed in a study to understand the prevalence of obesity, diabetes, and other cardiovascular risk factors in central Virginia
-   [Diabetes data](https://rdrr.io/cran/faraway/man/diabetes.html) is available as part of `faraway` R package
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| warning: false
#| message: false
#| include: true

# Diabetes data from faraway() package
df <- diabetes

c1 <- colnames(df)
c2 <- c("Subject ID", "Total Cholesterol [mg/dL]", "Stabilize Glucose [mg/dL]", 
        "High Density Lipoprotein [mg/dL]", "Cholesterol / HDL Ratio", "Glycosolated Hemoglobin [%]", 
        "County: Buckingham or Louisa", "age [years]", "gender", 
        "height [in]", "weight [lb]", "frame: small, medium or large", 
        "First Systolic Blood Pressure", "First Diastolic Blood Pressure", 
        "Second Systolic Blood Pressure", "Second Diastolic Blood Pressure", 
        "waist [in]", "hip [in]", 
        "Postprandial Time [min] when labs were drawn")

tbl <- data.frame(Abbreviation = c1, Description = c2)

kbl_font_size <- 14

tbl %>% 
  kbl(align = "c") %>%
  kable_paper("hover", full_width = F) %>%
  kable_styling(font_size = kbl_font_size)

```
:::
:::::

## Example data
*Diabetes data*

<br>

:::::: columns
::: {.column width="45%"}
-  We can calculate BMI as $$BMI = 703 \times \frac{\text{weight [lb]}}{\text{height [in]}^2} $$ 
-  And create a new binary variable `BMI>30` indicating if BMI is over 30 (1) or no (0)
-  First few observations are shown on the right.
:::

::: {.column width="5%"}
:::

::: {.column width="50%"}
```{r}

# add obesity variables
inch2m <- 2.54/100
pound2kg <- 0.45
data_diabetes <- diabetes %>%
  mutate(height  = height * inch2m, height = round(height, 2)) %>% 
  mutate(waist = waist * inch2m) %>%  
  mutate(weight = weight * pound2kg, weight = round(weight, 2)) %>%
  mutate(BMI = weight / height^2, BMI = round(BMI, 2)) %>% 
  mutate(`BMI>30`= cut(BMI, breaks = c(0, 29.9, 100), labels = c(0, 1))) %>% 
  relocate(BMI, .after = id) %>%
  relocate(`BMI>30`, .after = BMI) 
  
# preview data
glimpse(head(data_diabetes))
```
:::
::::::

## Example data

*Diabetes data + 1000 genes*

<br>

In addition, we simulated expression levels for 1,000 genes, with a subset of known obesity-associated genes simulated as up-regulated.

```{r}
#| fig-align: center

# Load clinical data
data_obesity_genes <- read_csv("data/data-obesity.csv") |>
  mutate(`BMI>30` = ifelse(obese == "No", 0, 1)) |>
  mutate(`BMI>30`= factor(`BMI>30`)) |>
  dplyr::select(-obese) |>
  dplyr::relocate(`BMI>30`, .after = "BMI")

# Load gene expression data
data_exprs <- read_csv("data/data-obesity-genes.csv")

# Join data
data <- data_obesity_genes %>%
  left_join(data_exprs, by = "id") #%>% 
  #mutate(obese = factor(obese, levels = c("No", "Yes")))

# show distribution of random 3 genes by obesity
data %>%
  select("BMI>30", 31:33) %>% 
  na.omit() %>%
  pivot_longer(-`BMI>30`, names_to = "gene", values_to = "expression") %>% 
  mutate(gene = factor(gene)) %>% 
  ggplot(aes(x = expression, fill = `BMI>30`)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) + 
  facet_wrap(~gene, scales = "free") +
  theme_minimal() + 
  scale_fill_tableau(name = "BMI>30",  palette = "Classic Blue-Red 6") + 
  theme(legend.position = "top")
  
```


## Example supervised project

*Building a model to predict whether BMI is ≥ 30.*

## Example unsupervised project

*Finding clusters and exploring their characteristics*

# A mini self-study: explore the demo projects

# Team work: discuss and outline your own project

## Other project ideas
<br>

- **Supervised learning**
  - Predict an outcome from a set of input features
  - Evaluate how well the model generalizes to new data
  - Build an interactive app that allows users to enter or select new data and explore predictions

. . .

::: incremental
- **Examples**
  - Predict disease status or treatment response
  - Predict a continuous outcome, such as BMI or a biomarker level
  - Compare predictions, feature importance, or model performance
:::

## Other project ideas
<br>

- **Unsupervised clustering**
  - Identify groups or patterns in your data
  - Characterize and interpret the resulting clusters
  - Build an interactive app that allows collaborators to explore the clusters, samples, features, or metadata

. . .

::: incremental
- **Examples**
  - Patient subgroups based on clinical or molecular profiles
  - Cell/sample clusters based on omics measurements
  - Project new data onto existing clusters, where applicable
:::


## Choose your stack

<br>

`Python` → `scikit-learn` → `Streamlit`

- Data handling: pandas
- Machine learning: scikit-learn
- Interactive app: Streamlit

<br>

`R` → `tidymodels / other modelling packages` → `Shiny`

- Data handling: tidyverse
- Machine learning: tidymodels, glmnet, etc.
- Interactive app: Shiny

. . .

*These are recommended options, not requirements. If your team already uses another suitable framework, you are welcome to use it.*


## Project requirements 
<br>

. . .

::: incremental
- Work in teams on a **life science ML project**
- Start from a clear **biological or research question**
- Use a **supervised or unsupervised** learning approach
- Clearly define the **samples, features, and outcome or structure of interest**
- Use a dataset that is:
  - large and complex enough for a meaningful ML analysis
  - computationally manageable on a standard laptop
  - suitable for analysis steps that take **minutes rather than hours**
- Evaluate and interpret the results appropriately
- Create a simple **interactive application** for predictions or exploration of results
:::

<br>

## Project reproducibility
<br>

. . .

::: incremental

1. **Code & workflow** — scripts/notebooks and the complete preprocessing and analysis pipeline
2. **Documentation** — `README.md` explaining what the project does and how to run it
3. **Environment** — record software and package versions
4. **Licence** — define how the work may be reused
5. **Citation** — e.g. `CITATION.cff` explaining how to cite the project
6. **Versioning** — Git + tagged releases preserve specific versions
7. **Archiving** — e.g. Zenodo creates a persistent archive and DOI

:::

## Final deliverables
<br>

Each team submits

1. **Link to the interactive application**
   - e.g. Streamlit, Shiny, or another suitable framework

2. **Link to the GitHub repository**
   - complete analysis code and workflow
   - `README.md` with instructions
   - software/package versions
   - licence and citation information
   - final tagged release / archived version
   - `CONTRIBUTIONS.md` with brief documentation of each team member’s contribution


## Team work
<br>

Use this session to get your project started as a team.

- Get to know each other
- Make sure everyone has a shared understanding of the **project goal**
- Discuss and make an initial agreement on:
  - research question / project goal
  - dataset
  - supervised or unsupervised approach
  - programming language / tools
  - how you will communicate and work together
  - initial division of tasks

## Before you leave {.smaller}
<br>

Write down initial project plan

**Research question / project goal:**  
What do you want to investigate, predict, or discover?

**Dataset:**  
What data will you use? Where does it come from?

**Samples:**  
What does one sample/observation represent?

**Features:**  
What variables or measurements will be used?

**Type of task:**  
Supervised or unsupervised?

**Expected output:**  
What would you like the model or analysis to reveal or do?

**Programming language / tools:**  
R / Python / other?

**Team communication:**  
How will you communicate and share work?

**Initial division of tasks:**  
Who will start with what?

. . .

<br>

**Talk to Olga or Payam and briefly go through the main points before disconnecting**