# Exercises (regression coefficients) {.unnumbered}

```{r}
#| messages: false
#| warnings: false
#| include: false

library(tidyverse)
```


:::{#exr-lm}

## Dose, genotype and enzyme activity

An experiment investigates how the activity of an enzyme changes with drug dose. Measurements were obtained from two genotypes, `WT` and `Mutant`, and samples were processed in three experimental batches.

Copy and run the code below to create the data.

```{r}
genotype <- factor(
  rep(c("WT", "Mutant"), each = 24),
  levels = c("WT", "Mutant")
)

dose <- rep(
  rep(c(0, 2, 4, 6, 8, 10), each = 4),
  2
)

batch <- factor(
  rep(c("B1", "B2", "B3"), length.out = 48),
  levels = c("B1", "B2", "B3")
)

activity <- c(
  13.43, 13.44, 10.60, 10.36,
  12.34, 13.12, 12.75, 13.67,
  14.61, 15.53, 13.64, 16.10,
  14.77, 20.82, 17.47, 16.79,
  21.95, 18.53, 20.10, 20.34,
  21.73, 22.38, 20.82, 19.50,
  14.65, 14.93, 10.53, 16.83,
  15.47, 13.89, 15.64, 19.03,
  22.03, 20.09, 20.89, 19.93,
  23.02, 24.61, 21.32, 24.07,
  28.11, 22.73, 26.46, 27.81,
  29.73, 28.64, 28.97, 27.76
)

enzyme <- data.frame(
  activity,
  dose,
  genotype,
  batch
)

head(enzyme)

```


a) Make an exploratory plot of `activity` against `dose`, using different
colours for the two genotypes. What patterns do you see?

b) Fit a linear model in R to model enzyme activity given `dose`, `genotype`
and `batch`.

c) What are the reference categories for `genotype` and `batch`?

d) Given the fitted model, what is the expected enzyme activity for a
`Mutant` sample with a `dose` of 5 from batch `B2`? Calculate using the model and check your calculations using `predict()` function.

e) Which model coefficients show evidence of an association with enzyme
activity at the 5% significance level?

:::

::: {.callout-tip title="Solution" collapse="true"}

a)

```{r}
#| code-fold: false
#| warnings: false
#| messages: false

ggplot(enzyme, aes(x = dose, y = activity, colour = genotype)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)


```

Enzyme activity appears to increase with increasing dose. The Mutant
samples also tend to have higher enzyme activity than the WT samples.

b) 

```{r}
#| code-fold: false
#| warnings: false
#| messages: false

model1 <- lm(
  activity ~ dose + genotype + batch,
  data = enzyme
)

summary(model1)
```

c) 

The reference categories are: WT for genotype and B1 for batch.

The coefficients for genotypeMutant, batchB2 and batchB3 therefore
describe differences relative to these reference categories.

d) 

From the fitted model,

$$ \widehat{activity} = 10.196 + 1.272 \cdot dose + 5.098 \cdot I_{\text{Mutant}} + 0.697 \cdot I_{\text{B2}} - 1.018 \cdot I_{\text{B3}}. $$

For a Mutant sample with dose = 5 from batch B2:

$$ \widehat{activity} = 10.196 + 1.272 \cdot 5 + 5.098 + 0.697 \approx 22.35. $$

We can also obtain the prediction in R:

```{r}
new_sample <- data.frame(
  dose = 5,
  genotype = factor("Mutant", levels = levels(enzyme$genotype)),
  batch = factor("B2", levels = levels(enzyme$batch))
)

predict(model1, newdata = new_sample)
```

e) 

At the 5% significance level:

- dose shows evidence of an association with enzyme activity ($p < 2\times10^{-16}$);
- genotypeMutant shows evidence of an association with enzyme activity ($p = 7.97\times10^{-13}$);
- there is no evidence that batchB2 differs from the reference batch B1 ($p = 0.270$);
- there is no evidence that batchB3 differs from the reference batch B1 ($p = 0.112$).

Thus, dose and genotype are clearly associated with enzyme activity in
this fitted model, while the individual batch coefficients are not
significantly different from the reference batch.

:::


::: {#exr-logistic}

## Treatment response

A clinical study investigates whether a biomarker and treatment are associated
with the probability of responding to treatment.

The outcome `response` is coded as:

- `1`: responded to treatment
- `0`: did not respond

The data also contain:

- `marker`: a continuous biomarker measurement
- `treatment`: either `Control` or `Drug`
- `age`: age in years

Copy and run the code below to create the data.

```{r}

response <- c(
  0,1,0,0,1,1,1,1,0,1,
  0,0,0,1,1,1,1,1,0,1,
  0,0,0,1,0,1,1,1,0,1,
  0,0,0,1,0,1,1,1,1,1,
  1,1,1,1,0,1,0,1,0,1,
  0,0,0,0,0,1,1,1,0,1
)

marker <- rep(
  c(0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4, 1.2, 2.8),
  6
)

treatment <- factor(
  rep(c("Control", "Drug"), 30),
  levels = c("Control", "Drug")
)

age <- rep(
  c(35, 42, 51, 60, 67, 48, 55, 39, 63, 45),
  6
)

trial <- data.frame(
  response,
  marker,
  treatment,
  age
)

head(trial)

```


a) Fit a logistic regression model to model the probability of `response`
given `marker`, `treatment` and `age`.

b) What is the reference category for `treatment`? What comparison does the
coefficient `treatmentDrug` therefore represent?

c) Using the Wald tests reported in the model summary, which coefficients show
evidence of an association with the probability of response at the 5%
significance level?

d) The estimated coefficient for `marker` is approximately $1.75$.

- Calculate the corresponding odds ratio.
- Interpret this odds ratio in the context of the study.

e) The estimated coefficient for `treatmentDrug` is approximately $1.51$.

- Calculate the corresponding odds ratio.
- Interpret this odds ratio in the context of the study.

f) Using the fitted model, estimate the probability of response for a
50-year-old patient with a `marker` value of 2.5 who receives the `Drug`
treatment. Check your answer using `predict()` with `type = "response"`.

::: {.callout-tip title="Solution" collapse="true"}

a) 

Fit the logistic regression model:

```{r}
#| code-fold: false

model2 <- glm(
  response ~ marker + treatment + age,
  family = binomial(link = "logit"),
  data = trial
)

summary(model2)
```

b) 

The reference category for treatment is Control.

Therefore, the coefficient treatmentDrug compares patients receiving the
Drug treatment with patients receiving the Control treatment, while
marker and age are held constant.

c) 

At the 5% significance level:

- marker shows evidence of an association with the probability of response ($p = 0.000662$);
- treatmentDrug shows evidence of an association with the probability of response ($p = 0.0497$);
- age does not show evidence of an association with the probability of response ($p = 0.197$).

d) 
The estimated coefficient for marker is

$$ \hat{\beta}_{marker} = 1.753. $$

The corresponding odds ratio is

$$ \exp(1.753) \approx 5.77. $$

Thus, for a one-unit increase in marker, the odds of treatment response are
estimated to be approximately 5.8 times higher, with treatment and age held
constant.

e) The estimated coefficient for treatmentDrug is

$$ \hat{\beta}_{Drug} = 1.507. $$

The corresponding odds ratio is

$$ \exp(1.507) \approx 4.51. $$

Thus, patients receiving the Drug have approximately 4.5 times the odds of
response compared with patients receiving Control, with marker and age
held constant.

The odds ratios can also be calculated in R:

```{r}
#|code-fold: false
exp(coef(model2))
```


f) 

For a 50-year-old patient with marker = 2.5 receiving the Drug, the fitted
log-odds are

$$ -0.970 + 1.753(2.5) + 1.507 - 0.058(50) \approx 2.03. $$

Converting the log-odds to a probability gives

$$ p = \frac{\exp(2.03)} {1+\exp(2.03)} \approx 0.884. $$

Thus, the fitted model predicts a probability of response of approximately
88%.

We can check this directly in R:

```{r}
#| code-fold: false

new_patient <- data.frame(
  marker = 2.5,
  treatment = factor(
    "Drug",
    levels = levels(trial$treatment)
  ),
  age = 50
)

predict(
  model2,
  newdata = new_patient,
  type = "response"
)

```



:::