---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Generalized linear models

## Why do we need GLM?

In linear regression, we described the **expected value of the outcome** for given values of the predictors.

For example, in simple linear regression,

$$
E(Y\mid X=x)=\beta_0+\beta_1x.
$$

More generally,

$$
E(Y_i\mid \mathbf{x}_i)
=
\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}.
$$

This works well for many continuous outcomes, but it is not appropriate for all types of data. For example, probabilities must lie between 0 and 1, while expected counts must be non-negative.

**Generalized linear models (GLMs)** extend the same framework to such outcomes.

We still construct a **linear predictor**

$$
\eta_i
=
\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}.
$$

However, instead of assuming that the expected outcome itself is equal to the linear predictor, we connect them through a **link function** $g(\cdot)$:

$$
g\left(E(Y_i\mid\mathbf{x}_i)\right)=\eta_i.
$$

Equivalently, if we denote the expected outcome by

$$
\mu_i=E(Y_i\mid\mathbf{x}_i),
$$

then

$$
g(\mu_i)=\eta_i.
$$

A GLM therefore specifies:

1. a probability distribution for the response $Y_i$, and
2. a link function connecting its expected value to the linear predictor.

Different choices give different models:

| Outcome    | Model               | Distribution | Link function |
| ---------- | ------------------- | ------------ | ------------- |
| Continuous | Linear regression   | Normal       | Identity      |
| Binary     | Logistic regression | Binomial     | Logit         |
| Counts     | Poisson regression  | Poisson      | Log           |

The ordinary linear model is a special case. With the **identity link**,

$$
g(\mu_i)=\mu_i,
$$

so that

$$
E(Y_i\mid\mathbf{x}_i)
=
\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}.
$$

For other types of outcomes, the link function allows us to retain the same linear predictor while modelling the expected outcome on an appropriate scale.




```{r}
#| message: false
#| include: false
#| echo: false

library(tidyverse)
library(magrittr)
library(faraway)
library(ggplot2)  
library(gridExtra)
library(kableExtra)
library(ggiraphExtra)

font.size <- 12
col.blue.light <- "#a6cee3"
col.blue.dark <- "#1f78b4"
my.ggtheme <- 
  theme_bw() + 
  theme(axis.title = element_text(size = font.size), 
        axis.text = element_text(size = font.size), 
        legend.text = element_text(size = font.size), 
        legend.title = element_blank(), 
        legend.position = "top") 
        

# add obesity and diabetes status to diabetes faraway data
inch2m <- 2.54/100
pound2kg <- 0.45
data_diabetes <- diabetes %>%
  mutate(height  = height * inch2m, height = round(height, 2)) %>% 
  mutate(waist = waist * inch2m) %>%  
  mutate(weight = weight * pound2kg, weight = round(weight, 2)) %>%
  mutate(BMI = weight / height^2, BMI = round(BMI, 2)) %>% 
  mutate(obese= cut(BMI, breaks = c(0, 29.9, 100), labels = c("No", "Yes"))) %>% 
  mutate(diabetic = ifelse(glyhb > 7, "Yes", "No"), diabetic = factor(diabetic, levels = c("No", "Yes"))) %>%
  na.omit()

```

So briefly, 

- GLMs extend linear model framework to outcome variables that do not follow normal distribution.
- They are most frequently used to model binary, categorical or count data.


For instance, consider a binary outcome, such as obesity status, coded as 0 (No) and 1 (Yes).

- If we fit an ordinary linear model to a binary outcome, the fitted values can fall below 0 or above 1 and therefore cannot always be interpreted as probabilities.
- In addition, the assumptions of the normal linear model are not appropriate for a binary response.

```{r}
#| label: fig-log-example
#| message: false
#| warning: false
#| fig-cap: "Example of fitting linear model to binary data, to model obesity status coded as 1 (Yes) and 0 (No) with waist variable. Linear model does not fit the data well in this case The titted values are not constrained to lie between 0 and 1 and therefore cannot always be interpreted as probabilities."
#| fig-height: 4
#| fig-width: 5

data_diabetes %>%
  mutate(obese = as.numeric(obese) - 1) %>%
  ggplot(aes(y=obese, x=waist)) +
  geom_jitter(width=0, height = 0) +
  geom_smooth(method="lm", se=FALSE, color=col.blue.dark) + 
  my.ggtheme

```


## Logistic regression

Let's look again at the binary `obesity` status data and try to fit logistic regression model using `waist` as explanatory variable instead of fitting inappropriate here simple linear model.


```{r}
#| label: fig-obesity
#| fig-cap: "Obesity status data: jittered plot (left) waist measurements and their obesity status for the 130 study participants."
#| fig-cap-location: margin

p1 <- data_diabetes %>%
  mutate(obese = as.numeric(obese) - 1) %>%
  ggplot(aes(y=obese, x=waist)) +
  geom_jitter(width=0, height = 0.05, alpha = 0.7) +
  my.ggtheme

# p2 <- data_diabetes %>%
#   ggplot(aes(x = obese, y = waist, fill = obese)) + 
#   geom_boxplot() + 
#   scale_fill_brewer(palette = "Set2") + 
#   my.ggtheme

p1
```


Let's look again at the binary **`obesity`** status data. We will model obesity status (the probability of being obese) using **`waist`** as an explanatory variable, this time using logistic rather than simple linear regression.

For a binary response, let

$$
Y_i =
\begin{cases}
1 & \text{if the individual is obese},\\
0 & \text{otherwise}.
\end{cases}
$$

We denote the probability that individual $i$ is obese, given their waist measurement $x_i$, by

$$
p_i = P(Y_i = 1 \mid X_i = x_i).
$$

Since $Y_i$ is coded as 0 or 1, its expected value is equal to this probability:

$$
E(Y_i \mid X_i = x_i) = p_i.
$$

For a binary observation,

$$
Y_i \sim \operatorname{Bernoulli}(p_i),
$$

which is equivalent to a binomial distribution with one trial.

In **logistic regression**, we use the **logit link** to relate the probability $p_i$ to the linear predictor, thus rewriting linear model to:

$$
\log\left(\frac{p_i}{1-p_i}\right)
=
\beta_0+\beta_1x_i.
$$



The quantity

$$
\frac{p_i}{1-p_i}
$$

is called the **odds**, and its logarithm is the **log-odds** or **logit**.

Equivalently, we can express the probability as

$$
p_i
=
\frac{\exp(\beta_0+\beta_1x_i)}
{1+\exp(\beta_0+\beta_1x_i)}.
$$

This transformation ensures that the fitted probabilities always lie between 0 and 1.

And here the **link function** $log(\frac{p_i}{1-p_i})$ provides the link between the binomial distribution of $Y_i$ (probabiilty of suffering from obesity) and the linear predictor (waist) 

<br>

In R we can use `glm()` function to fit GLM models:
```{r}
#| code-fold: false
#| collapse: true
#| message: false
#| warning: false
#| fig-cap: "Fitted logistic model to diabetes data given the 130 study participants and using waist as explantatory variable to model obesity status."

# re-code obese status from Yes/No to 1/0
data_diabetes <- 
  data_diabetes %>%
  mutate(obese = as.numeric(obese) - 1)

# fit logistic regression model
logmodel_1 <- glm(obese ~ waist, family = binomial(link="logit"), data = data_diabetes)

# print model summary
print(summary(logmodel_1))

# plot
ggPredict(logmodel_1) + 
  my.ggtheme
# to get predictions use predict() functions
# if no new observations is specified predictions are returned for the values of exploratory variables used
# we specify response to return prediction on the probability scale
obese_predicted <- predict(logmodel_1, type="response")
print(head(obese_predicted))

```

- The regression equation for the fitted model is:
$$log(\frac{\hat{p_i}}{1-\hat{p_i}})=-17.357  +  17.174\cdot x_i$$
- We see from the output that $\hat{\beta_0} = -17.357$ and $\hat{\beta_1} = 17.174$.

The coefficient estimates are obtained using **maximum likelihood estimation**. The idea is to find the values of the model coefficients that make the observed data most likely under the model. This is done using **numerical optimization**, where the algorithm searches over possible coefficient values until it finds the combination that gives the highest likelihood.

### Hypothesis testing

As in linear regression, we can test whether an explanatory variable is
associated with the outcome by testing whether its regression coefficient is
zero.

For a coefficient $\beta_j$, we test

$$
H_0:\beta_j=0
$$

against

$$
H_a:\beta_j\neq0.
$$

In logistic regression, individual coefficients are commonly tested using a
**Wald test**. The Wald statistic is the estimated coefficient divided by its
estimated standard error:

$$
W =
\frac{\hat{\beta}_j - 0}
{e.s.e.(\hat{\beta}_j)}.
$$

Under the null hypothesis, for sufficiently large samples, the Wald statistic
approximately follows the standard normal distribution:

$$
W \sim N(0,1).
$$

This is analogous to the t-test for a coefficient in linear regression. In
the output from `summary(glm(...))`, this statistic is reported as the
**z value**.

In our example, we want to test whether `waist` is associated with the
probability of being obese:

$$
H_0:\beta_1=0.
$$

From the fitted model,

$$
\hat{\beta}_1 = 17.174
$$

with estimated standard error

$$
e.s.e.(\hat{\beta}_1)=2.974.
$$
Therefore,

$$
W =
\frac{17.174}{2.974}
\approx 5.77.
$$

We can calculate the corresponding two-sided *P*-value from the standard normal distribution:

```{r}
#| code-fold: false
#| collapse: true

2 * pnorm(abs(5.77), lower.tail = FALSE)
```

The resulting P-value is very small, providing sufficient evidence to reject the null hypothesis and conclude that waist is associated with the probability of being obese.


### Deviance

Another way to test whether an explanatory variable contributes to a GLM is
to compare two **nested models**:

- a **reduced model**, without the explanatory variable of interest;
- a **full model**, including that variable.

For example, to test the contribution of `waist`, we compare

$$
\text{reduced model: } \operatorname{logit}(p_i)=\beta_0
$$

with

$$
\text{full model: } \operatorname{logit}(p_i)=\beta_0+\beta_1x_i.
$$

The comparison is based on the **deviance**, which measures lack of fit:
smaller deviance indicates a better-fitting model.

The likelihood-ratio test uses the difference in deviance between the two
models:

$$
D_{\text{reduced}} - D_{\text{full}}.
$$

Under the null hypothesis that the additional term does not improve the model,
this difference approximately follows a chi-squared distribution. The degrees
of freedom correspond to the difference in the number of estimated parameters
between the models.

For our example, the intercept-only model has deviance

$$
D_{\text{reduced}} = 178.71,
$$

while the model including `waist` has deviance

$$
D_{\text{full}} = 102.79.
$$

The reduction in deviance is therefore

$$
178.71 - 102.79 = 75.92.
$$


Adding `waist` introduces one additional parameter. Under the null hypothesis,
the deviance difference approximately follows a chi-squared distribution with
1 degree of freedom:

$$
\Delta D \sim \chi^2_1.
$$

For our data, the observed deviance difference is

$$
\Delta D = 75.92.
$$

The corresponding *P*-value is very small, providing strong evidence that
including `waist` improves the fit of the model.

We can perform this comparison directly in R:

```{r}
#| code-fold: false
#| collapse: true

logmodel_0 <- glm(
  obese ~ 1,
  family = binomial(link = "logit"),
  data = data_diabetes
)

anova(logmodel_0, logmodel_1, test = "Chisq")

```


### Odds ratios

The coefficients in logistic regression describe changes in the **log-odds**
of the outcome. Because log-odds are difficult to interpret directly, we often
exponentiate the coefficients:

$$
OR = \exp(\hat{\beta}).
$$

The resulting quantity is an **odds ratio**.

For a numerical predictor, the odds ratio describes the multiplicative change
in the odds of the outcome associated with a one-unit increase in the predictor,
with other variables in the model held constant.

In our example, `waist` is measured in metres. Therefore, $\exp(\hat{\beta}_1)$
would describe the change in odds associated with a 1 metre increase in waist
circumference, which is not a useful unit for interpretation.

Instead, we can consider an increase of 10 cm, corresponding to 0.1 metres:

$$
\exp(0.1\hat{\beta}_1)
=
\exp(0.1 \times 17.174)
\approx 5.6.
$$

Thus, a 10 cm increase in waist circumference is associated with approximately
**5.6 times higher odds of being obese**.

In general:

- an odds ratio of $1$ indicates no change in the odds of the outcome;
- an odds ratio $>1$ indicates higher odds of the outcome as the predictor increases;
- an odds ratio $<1$ indicates lower odds of the outcome as the predictor increases.

The further an odds ratio is from 1, the larger the estimated association on
the odds scale. However, the size of an odds ratio alone does **not** provide
evidence that the relationship is causal.

### Other covariates

As in multiple linear regression, we can extend a logistic regression model by
including additional explanatory variables. These can be numerical, binary or
categorical.

For example, we can investigate whether `gender` is associated with the
probability of being obese after accounting for `waist`:

```{r}
#| message: false
#| warning: false
#| code-fold: false
#| collapse: true
#| fig-cap: "Fitted probability of obesity as a function of waist circumference, shown separately by gender."
#| fig-cap-location: margin

# fit logistic regression including waist and gender
logmodel_2 <- glm(
  obese ~ waist + gender,
  family = binomial(link = "logit"),
  data = data_diabetes
)

# print model summary
print(summary(logmodel_2))

# plot fitted probabilities
ggPredict(logmodel_2) +
  my.ggtheme +
  scale_color_brewer(palette = "Set2")

```

The model can be written as

$$\operatorname{logit}(p_i) = \beta_0 + \beta_1 \,\text{waist}_i + \beta_2 \,\text{gender}_i$$

The interpretation follows the same principle as in multiple linear regression:
each coefficient describes the association with the outcome while the other
variables in the model are held constant.

Thus, the coefficient for gender describes the association between gender and
the probability of being obese for individuals with the same waist
circumference.

## Poisson regression

GLMs can also be used for **count data**, for example:

- number of hospital admissions due to respiratory disease;
- number of mutations observed in a sample;
- number of bird nests in a habitat.

For count data, a common starting point is to assume that

$$
Y_i \sim \operatorname{Poisson}(\mu_i),
$$

where

$$
\mu_i = E(Y_i \mid \mathbf{x}_i)
$$

is the expected count for observation $i$.

In **Poisson regression**, we use the **log link** to relate the expected count
to the linear predictor:

$$
\log(\mu_i)
=
\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}.
$$

Equivalently,

$$
\mu_i
=
\exp\left(
\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}
\right).
$$

The log link ensures that the fitted expected counts are always positive.

Hypothesis testing and comparison of nested models follow the same general
logic as for logistic regression. Individual coefficients can be tested using
Wald tests, while models can be compared using differences in deviance.

::: {#exm-poisson}

### Number of cancer cases

Suppose we wish to model $Y_i$, the number of cancer cases in geographical
region $i$ in Glasgow.

We have data from 271 small regions, together with information on:

- `Y_all`: observed number of cases of all types of cancer in the region in 2013;
- `E_all`: expected number of cancer cases based on the population size and
  demographic structure of the region;
- `pm10`: air pollution;
- `smoke`: percentage of people who smoke;
- `ethnic`: percentage of people who are non-white;
- `log.price`: natural logarithm of average house price;
- `easting` and `northing`: geographical coordinates.

The regions differ in population size and demographic structure. We therefore
do not want to compare the raw numbers of cancer cases directly.

Instead, we include `E_all` as an **offset**. The model becomes

$$
\log(\mu_i)
=
\log(E_i)
+
\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}.
$$

Here, $\log(E_i)$ is included with its coefficient fixed at 1. It accounts for
the number of cases that would already be expected in each region based on its
population and demographic structure.

Equivalently,

$$
\mu_i
=
E_i
\exp\left(
\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}
\right).
$$

We can fit the model using `glm()` with the Poisson family:

```{r}
#| code-fold: false
#| collapse: true

# Read in and preview data
cancer <- read.csv("data/lm/cancer.csv")

head(cancer)

# Fit Poisson regression
epid1 <- glm(
  Y_all ~ pm10 + smoke + ethnic + log.price +
    easting + northing + offset(log(E_all)),
  family = poisson,
  data = cancer
)

print(summary(epid1))
```

### Rate ratios {-}

As in logistic regression, coefficients in Poisson regression are often easier
to interpret after exponentiation:

$$
RR = \exp(\hat{\beta}).
$$

The resulting quantity is a **rate ratio**.

For a numerical predictor, the rate ratio describes the multiplicative change
in the expected rate associated with a one-unit increase in the predictor,
while the other variables in the model are held constant.

For example, suppose we are interested in the association between air pollution
and cancer incidence. The estimated coefficient for `pm10` is

$$
\hat{\beta}_{pm10} = 0.0500.
$$

Since this coefficient is positive, higher `pm10` values are associated with a
higher expected cancer rate.

Exponentiating the coefficient gives

$$
\exp(0.0500) \approx 1.05.
$$

Thus, a one-unit increase in `pm10` is associated with approximately a
**1.05-fold**, or **5% higher**, expected cancer incidence rate, while the other
variables in the model are held constant.


