---
title: "Quiz questions"
subtitle: "Statistical Methods for Life Sciences"
format:
  html:
    toc: true
    number-sections: true
    code-fold: false
---

## Quiz {.unnumbered}

::: {#exr-multiple-regression}

## Multiple regression

Suppose researchers investigate final plant biomass as a function of:

- nitrogen — soil nitrogen concentration (mg/kg), continuous
- light — average light exposure (hours/day), continuous
- genotype — WT, MutA, or MutB, categorical

A multiple linear regression model was fitted:

```{r}
set.seed(123)

n <- 60

plants <- data.frame(
  nitrogen = round(runif(n, 20, 80), 1),
  light = round(runif(n, 6, 14), 1),
  genotype = factor(
    rep(c("WT", "MutA", "MutB"), each = 20),
    levels = c("WT", "MutA", "MutB")
  )
)

plants$biomass <- round(
  3 +
  0.10 * plants$nitrogen +
  0.45 * plants$light +
  
  # genotype main effects
  ifelse(plants$genotype == "MutA", 1.0,
         ifelse(plants$genotype == "MutB", -1.0, 0)) +
  
  # interaction: MutA responds more strongly to nitrogen
  ifelse(plants$genotype == "MutA",
         0.08 * plants$nitrogen, 0) +
  
  rnorm(n, mean = 0, sd = 1.5),
  1
)


fit1 <- lm(biomass ~ nitrogen + light + genotype, data = plants)
summary(fit1)

```

Which statements are correct?

:::

a) The fitted model can be written as

$$ \hat{Y}=0.920+0.126(\text{nitrogen})+0.501(\text{light}) +6.066I(\text{MutA})-1.126I(\text{MutB}). $$

b) The coefficient for nitrogen means that, for every one-unit increase in nitrogen, expected biomass increases by about 0.126 units, holding light and genotype constant.

c) The p-value for genotypeMutB tests the null hypothesis that the mean biomass of MutB plants is equal to the mean biomass of MutA plants.

d) Since the p-value for genotypeMutB is 0.0199, there is evidence at the 5% significance level that MutB differs from WT in expected biomass, after accounting for nitrogen and light.

e) The significant overall F-test shows that every individual predictor in the model is significantly associated with biomass.

::: {.callout-tip title="Answer" collapse="true"}

Correct: a, b, d.


a) True.
With WT as the reference category, the fitted model includes separate indicator variables for MutA and MutB. WT is represented when both indicators are 0.

b) True.
The nitrogen coefficient is interpreted conditionally: for a one-unit increase in nitrogen, expected biomass increases by about 0.126 units, assuming light exposure and genotype stay the same.

c) False.
The p-value for genotypeMutB tests

$$ H_0:\beta_{\text{MutB}}=0, $$

which corresponds to no difference between MutB and the reference group WT, after accounting for nitrogen and light. It does not directly compare MutB with MutA.

d) True.
Because \(p=0.0199<0.05\), we reject \(H_0:\beta_{\text{MutB}}=0\). There is evidence that MutB differs from WT in expected biomass after adjusting for nitrogen and light. Since the coefficient is negative, MutB has lower expected biomass than WT.

e) False.
The overall F-test evaluates whether the predictors are jointly useful:

$$ H_0:\beta_{\text{nitrogen}} =\beta_{\text{light}} =\beta_{\text{MutA}} =\beta_{\text{MutB}}=0. $$

A significant F-test tells us that at least one of these coefficients is non-zero. It does not imply that every individual predictor is significant.

:::

::: {#exr-poisson}

Researchers are studying how antibiotic concentration and bacterial strain affect bacterial growth. For each sample, they record the number of bacterial colonies observed after incubation.

The response variable colonies is therefore a count, and a Poisson GLM with a log link is fitted:

glm(colonies ~ antibiotic + strain,
    family = poisson(link = "log"),
    data = bacteria)

where:

colonies = number of bacterial colonies, a count response
antibiotic = antibiotic concentration, a continuous predictor
strain = bacterial strain, a categorical predictor with levels WT and Mut

WT is the reference level for strain.

Study the session notes on Poisson regression to understand how modelling counts work. 
Then, given the output, which statements are correct?


```{r}
set.seed(123)

n <- 80

bacteria <- data.frame(
  antibiotic = round(runif(n, 0, 5), 2),
  strain = factor(
    sample(c("WT", "Mut"), n, replace = TRUE),
    levels = c("WT", "Mut")
  )
)

# Expected number of colonies
lambda <- exp(
  3.2 -
  0.35 * bacteria$antibiotic +
  0.42 * (bacteria$strain == "Mut")
)

# Generate colony counts
bacteria$colonies <- rpois(n, lambda = lambda)

head(bacteria)

fit <- glm(
  colonies ~ antibiotic + strain,
  family = poisson(link = "log"),
  data = bacteria
)

summary(fit)

exp(coef(fit))
```


a) The fitted model can be written as


$$\log(\hat{\mu}) =
3.183 - 0.335(\text{antibiotic})
+ 0.416I(\text{Mut})$$

where $\hat{\mu}$ is the expected number of colonies.

b) A one-unit increase in antibiotic concentration is associated with a decrease of approximately 0.335 colonies in the expected colony count, holding strain constant.

c) A one-unit increase in antibiotic concentration is associated with an expected colony count that is approximately 28.5% lower, holding strain constant.

d) At the same antibiotic concentration, the expected colony count for the Mut strain is approximately 52% higher than for the WT strain.

e) The p-value for strainMut tests the null hypothesis that there is no difference in expected colony counts between Mut and WT, after accounting for antibiotic concentration.



:::

::: {.callout-tip title="Answer" collapse="true"}

Correct: a, c, d, e.

a) True.
With a Poisson GLM and a log link, the model is written for the log of the expected count. The categorical variable strain is represented using an indicator for Mut, with WT as the reference.

b) False.
The coefficient \(-0.335\) is on the log scale, so it is not an additive decrease of 0.335 colonies. To interpret it on the original count scale, we exponentiate the coefficient.

c) True.

$$ e^{-0.335}\approx 0.715. $$

So a one-unit increase in antibiotic concentration multiplies the expected colony count by about 0.715. This corresponds to

$$ (0.715-1)\times100\%\approx -28.5\%. $$

So the expected count is about 28.5% lower, holding strain constant.

d) True.
For strainMut,

$$ e^{0.416}\approx1.516. $$

So, at the same antibiotic concentration, the expected colony count for Mut is about 1.52 times that for WT, or approximately 52% higher.

e) True.
The test for strainMut evaluates

$$ H_0:\beta_{\text{Mut}}=0. $$

Because WT is the reference level, this corresponds to testing whether Mut and WT have the same expected colony count after accounting for antibiotic concentration. The very small p-value provides strong evidence against this null hypothesis.

:::