---
title: "Quiz questions"
subtitle: "Statistical Methods for Life Sciences"
format:
  html:
    toc: true
    number-sections: true
    code-fold: false
---

## Quiz {.unnumbered}


::: {#exr-quiz-1}

## Question 1: Simple linear regression

Which statements about **simple linear regression** are correct?

a) It can be used to describe the relationship between a numerical
outcome and one numerical explanatory variable.

b) It finds a best-fitting straight line describing how the expected
outcome changes with the explanatory variable.

c) All observations are expected to lie exactly on the fitted
regression line.

d) Individual observations can differ from the values described by
the regression line.

e) The regression line describes the expected value of the outcome
for a given value of the explanatory variable.

:::

::: {.callout-tip title="Answer" collapse="true"}

**Correct answers: a), b), d) and e).**

a) In simple linear regression, we model a numerical outcome $Y$
using one numerical explanatory variable $X$.

b) The fitted model

$$
E[Y\mid X=x] = \beta_0 + \beta_1x
$$

describes how the expected outcome changes with $X$.

c) is false. Observed values do not generally lie exactly on the
regression line. The difference between an observed and fitted value
is called a **residual**.

d) is correct. The regression line describes the expected outcome,
while individual observations may vary around this value.

e) is correct. For a given value of $X$, the regression line gives
the modelled expected value of $Y$.

:::


::: {#exr-quiz-2}

## Question 2: Fitting a linear model in R

Consider the following data:

```{r}
x <- c(1, 2, 3, 4)
y <- c(2, 4, 4, 6)

fit <- lm(y ~ x)

```

Which statements are correct? 

a) The estimated intercept is 1.

b) The estimated slope is 1.2.

c) The fitted value for $x=3$ is 4.6.

d) The residual for the third observation is $4-4.6=-0.6$.

e) The fitted model predicts that $Y$ increases by 1 unit for every
one-unit increase in $X$.

:::

::: {.callout-tip title="Answer" collapse="true"}

Correct answers: a), b), c) and d).

We can inspect the estimated coefficients using:

```{r}
coef(fit)
```

The fitted model is

$\hat{y}=1+1.2x$

The estimated intercept is therefore 1.

b) The estimated slope is 1.2. This means that the expected value of
$Y$ increases by 1.2 for each one-unit increase in $X$.

c) For $x=3$,

$$ \hat y = 1+1.2\times3 = 4.6. $$

d) The observed value for the third observation is 4, so its residual is

$$ e = y-\hat y = 4-4.6 = -0.6. $$

e) is false. The estimated increase is 1.2 units, not 1 unit,
for each one-unit increase in $X$.


:::

::: {#exr-quiz-3}

## Question 3: Estimation and hypothesis testing

Which statements about fitting and testing a linear regression model
are correct? **Select all that apply.**

a) Least squares estimates the model coefficients by minimizing the
residual sum of squares.

b) The residual sum of squares is

$$
RSS=\sum_{i=1}^{n}(y_i-\hat y_i)^2.
$$

c) To test whether an explanatory variable is associated with the
outcome, a common null hypothesis is $H_0:\beta_1=0$.

d) A small p-value provides evidence that the true slope is exactly
equal to the estimated slope.

e) The standard error of $\hat\beta_1$ describes the uncertainty in
the estimated slope.

:::

::: {.callout-tip title="Answer" collapse="true"}

**Correct answers: a), b), c) and e).**

a) Least squares chooses the coefficient estimates that minimize the
squared differences between the observed and fitted values.

b) These squared differences are summed to give the **residual sum
of squares**:

$$
RSS=\sum_{i=1}^{n}(y_i-\hat y_i)^2.
$$

c) To test whether there is evidence of an association between $X$
and $Y$, we can test

$$
H_0:\beta_1=0
$$

against an appropriate alternative hypothesis.

d) is false. A small p-value provides evidence against the null
hypothesis. It does not tell us that the true slope is exactly equal
to the estimated slope.

e) is correct. The standard error describes the sampling uncertainty
of the estimated coefficient $\hat\beta_1$.

:::

::: {#exr-quiz-4}

## Question 4: Model fit and assumptions

Which statements about assessing a fitted linear model are correct?
**Select all that apply.**

a) $R^2$ can be calculated as

$$
R^2=1-\frac{RSS}{TSS}.
$$

b) A larger $R^2$ means that a larger proportion of the variability
in the outcome is described by the fitted model.

c) A high $R^2$ guarantees that the assumptions of the linear model
are satisfied.

d) Residuals can be used to assess assumptions of the linear model.

e) Linear model assumptions include linearity, independence of errors,
constant error variance and normally distributed errors.

:::

::: {.callout-tip title="Answer" collapse="true"}

**Correct answers: a), b), d) and e).**

a) $R^2$ compares the unexplained variation, measured by $RSS$, with
the total variation in the outcome, measured by $TSS$:

$$
R^2=1-\frac{RSS}{TSS}.
$$

b) $R^2$ describes the proportion of variability in the outcome
accounted for by the fitted model.

c) is false. A model can have a high $R^2$ while still violating
important model assumptions.

d) is correct. Examining residuals can help us assess whether the
assumptions of the fitted model are reasonable.

e) is correct. The standard assumptions include **linearity,
independence of errors, constant error variance and normality of
errors**.

:::