---
title: 'Session regression II: multiple linear regression'
output:
  pdf_document: default
  md_document:
    variant: markdown_github
  html_document:
    df_print: paged
header-includes: \usepackage{float}
urlcolor: blue
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path="session-regression-II-files/figures/")
knitr::opts_chunk$set(fig.pos = 'H')
knitr::opts_chunk$set(cache.path = "tmp")
knitr::opts_chunk$set(cache = TRUE)
```

# Learning outcomes

After this session, a student should be able to:

- visualize bivariate relationships in small datasets
- fit a linear regression model containing main and interaction effects
- assess the quality of the model fit
- determine if at least one predictor can explain the response
- determine which predictors explain the response
- asses the accuracy of predictions from the model

-------

# Visualizing the Advertising dataset

In this session we will use the [Advertising dataset](http://www-bcf.usc.edu/~gareth/ISL/data.html).
This simple dataset consists of sales data for 200 products along with the amount of money spent 
on TV, radio, and newspaper ads. We would like to know how best to spend advertizing money to
maximize sales.

To begin with, let us familiarize ourselves with the dataset. 
The data are stored in the `data` subdirectory of this session.

```{r}
# load the data
ads = read.csv('./data/Advertising.csv')
```

First, we check to see what columns were imported

```{r}
head(ads)
```

It looks like a redundant column, `X`, has made it into the table. However, the other columns look like numbers.
That's good.

Now, let us use the `pairs` function to get a quick overview of linear relationships within the dataset. 

```
# visualize all pairwise relationships
pairs(ads)
```

# Estimating the Regression Coefficients
# Estimating coefficients
# Relationship between the response and predictors 
# Model fit 
# Predictions 
# Qualitative predictors 
# Interaction terms
# Non-linear transformation of the predictors 
# Potential problems: non-linearity, collinearity 
# Logistic regression
