## Learning outcomes {-}

- be aware of data types
- compute measures of location, including mean and median
- compute measures of spread, including quantiles, variance and standard deviation
- understand the concept random sample
- compute population mean and variance
- compute sample mean and variance


```{r setup, include=FALSE}
library(tidyverse)
require(kableExtra)
library(reshape2)
library(ggplot2)
library(ggbeeswarm)
require(knitr)
#require(UsingR)
knitr::opts_chunk$set(fig.width=3.5, fig.height=3.5, echo = FALSE, error=FALSE, warnings=FALSE, dpi=600, fig.path = "figures/")
options(digits=2)
```

# Data types {#desc-01datatypes}

Data can be divided into different types; categorical and quantitative (numeric). How to summarize and analyze your data depends on the type.

Categorical data types are divided into;

* Nominal: named. 
*Ex: dead/alive, healthy/sick, WT/mutant, AA/Aa/aa, male/female, red/green/blue*
* Ordinal: named and ordered. 
*Ex: pain (weak, moderate, severe), AA/Aa/aa, very young/young/middle age/old/very old, grade I, II, III, IV*

Quantitative data types are divided into;

* Discrete: finite or countable infinite values. *Ex. counts, number of cells, number of reads*
* Continuous: infinitely many uncountable values. *Ex. height, weight, concentration*



# Categorical data {#desc-02categorical}

## Data summarization and plots

Categorical data can be summarized by counting the number of observations of each category and summarizing in a frequency table or bar plot. Alternatively, the proportions (or percentages) of each category can be calculated.

```{r simulatemice}
set.seed(101)
n <- 100
week <- 5:24
m <- length(week)

mice0 <- data.frame(id=1:n, gender=sample(c('male', 'female'), n, replace = TRUE)) %>% mutate(male=1*(gender=="male"), b0=rnorm(n, 15.5, 1), b1=rnorm(n, 0.5, 0.1), b2=rnorm(n, 1.5,0.5), b3=rnorm(n, 0.3, 0.1), a0 = rnorm(n, 44, 4))
mice <- mice0 %>% slice(rep(1:n(), each=m)) %>% mutate(week=rep(week, n))
mice <- mice %>% mutate(s=rnorm(n*m, 0, 0.2)) %>% mutate(weight=b0+b1*week+b2*male+b3*male*week+s)

##mice <- data.frame(id=factor(rep(1:n, each=m)), week=rep(week, n), gender=rep(sample(c('male', 'female'), n, replace = TRUE), each=m)) %>% mutate(male=1*(gender=="male"), b0=rep(rnorm(n, 15.5, 1), each=m), b1=rep(rnorm(n, 0.5, 0.1), each=m), b2=rep(rnorm(n, 1.5,0.5), each=m), b3=rep(rnorm(n, 0.3, 0.1), each=m), s=rnorm(n*m, 0, 0.2)) %>% mutate(weight=b0+b1*week+b2*male+b3*male*week+s)
mice0 <- mice 
mice <- mice %>% dplyr::select(id, gender, male, week, weight)
write.csv(mice, file="data/mice.csv", quote=FALSE)
```

```{example, label="10mice", echo=TRUE}

**Ten lab mice**
  
Observe gender and weight of your ten lab mice and summarize.
```

If you want to follow this example, you can download the data here; [mice.csv](data/mice.csv). You get the subset used in this example by the following commands; 

```{r read mice, eval=FALSE, echo=TRUE}
## first read the full data set into R
mice <- read.csv("mice.csv")
```


```{r m10subset, echo=TRUE}
## Then extract the specific subset used in this example
m10 <- subset(mice, subset=week==5 & id %in% 1:10, select = c(id, gender, weight))
```

In this example we have only ten observations (mice) and the full data can actually be shown in a table.

```{r m10}
m10  %>% kable(caption="Gender and weight of 10 mice.") %>% kable_styling(full_width = FALSE)
```

We are interested in the gender distribution in our group of mice. Count the frequency of male/female mice and summarize in a table. Also, the fraction or percentage can be useful.

```{r tabm10, fig.show="hold", out.width="49%", fig.cap="The number of male and female mice"}
kable(m10 %>% group_by(gender) %>% tally() %>% mutate("percent (%)"=n/sum(n)*100) , caption="The number of male and female mice.") %>% kable_styling(full_width = FALSE)
```

The frequencies can also be shown in a barplot.

```{r m10barplot, fig.show="hold", out.width="49%", fig.cap="The number of male and female mice shown in barplots generated using ggplot and basic R graphics.", echo=TRUE}
ggplot(m10, aes(x=gender)) + geom_bar()
barplot(table(m10$gender))
```


```{example, label="lefthandedness", echo=TRUE}
**Left handedness**

You are interested in whether left-handedness is associated to a disease you study and observe left-handedness among 30 patients as well as among 40 healthy controls;
``` 

```{r lefthand, echo=FALSE}
set.seed(1123)
patients <- sample(c("L", "R"), 30, prob=c(0.3, 0.7), replace=TRUE)
controls <- sample(c("L", "R"), 40, prob=c(0.1, 0.9), replace=TRUE)
hand <- rbind(data.frame(group="patient", handedness=patients),
              data.frame(group="control", handedness=controls))
```

patients: {`r patients`}

controls: {`r controls`}

Summarized as 

```{r table1}
hand %>% group_by(group) %>% dplyr::summarize(Total=n(), n = sum(handedness=="L"),  f=100*mean(handedness=="L")) %>% transmute(group, "Total n"=Total, "Left handed (%)"=sprintf("%i (%d%%)", n, f)) %>% kable(caption="Summary of left handedness among patients and controls.")
```

or in a contingency table;

```{r}
#hand %>% group_by(group, handedness) %>% dplyr::summarize(n=n()) %>% spread(handedness, n)
addmargins(table(hand$group, hand$handedness)) %>% kable(caption="Cross table or contingency table of group and left/right handedness.")
```

Data can be summarized in barplots in several ways;

```{r barlefthand, fig.show="hold", out.width="33%", fig.cap="Left-handedness in patient and control groups.", echo=TRUE}
## Using ggplot to create barplots
ggplot(hand, aes(x=group, fill=handedness)) + geom_bar()
ggplot(hand, aes(x=group, fill=handedness)) + geom_bar(position="dodge")
ggplot(hand, aes(x=group, fill=handedness)) + geom_bar(position="fill") + ylab("Fraction")

## Using basic R graphics to create barplots
tab <- table(hand$handedness, hand$group)
barplot(tab)
barplot(tab, beside=TRUE)
tabperc <- tab
tabperc[,1] <- 100*tab[,1]/sum(tab[,1])
tabperc[,2] <- 100*tab[,2]/sum(tab[,2])
barplot(tabperc)
```


# Quantitative data {#desc-03quantitative}

Quantitative data (both discrete and continuous) can be visualized in a histogram;

```{r sum10dice, fig.show="hold", out.width="50%", fig.cap="Throw 10 dice and count the total number of dots. Repeat the experiment 1000 times. This histogram summarize the results, i.e. the total number of dots when throwing 10 dice."}
sum10dice <- replicate(1000, sum(sample(1:6, 10, replace=TRUE)))
hist(sum10dice)
ggplot(data.frame(sum10dice), aes(x=sum10dice)) + geom_histogram(bins=25, color="white")
```

or a density plot;

```{r dicedens, fig.cap="Density plot over the total number of dots when throwing 10 dice.", out.width="50%"}
ggplot(data.frame(sum10dice), aes(x=sum10dice)) + geom_density()
```

```{r hist, fig.cap="Histogram over weight of 2000 5 weeks old mice, colored according to gender.", out.width="50%", fig.show="hold"}
micew5 <- mice %>% filter(week==5)
ggplot(micew5, aes(x=weight, fill=gender)) + geom_histogram(bins=15, color="white")
hist(micew5$weight[micew5$gender=="female"], col="orange", breaks=15, main="Weight of mice at 5 weeks", xlab="Weight (g)")
hist(micew5$weight[micew5$gender=="male"], col="cornflowerblue", breaks=15, add=TRUE)
```

Other useful plots are boxplots and beeswarm plots;

```{r boxplot, fig.cap="Boxplot over weight of 100 5 weeks old mice, divided according to gender.", out.width="50%", fig.show="hold"}
ggplot(micew5, aes(x=gender, y=weight)) + geom_boxplot() 
```
```{r beeswarm, fig.cap="Boxplot over weight of 20 5 weeks old mice, divided according to gender.", out.width="50%", fig.show="hold"}
ggplot(micew5, aes(x=gender, y=weight)) + geom_beeswarm() 
```
To study the relationship between two variables scatterplots are useful;

```{r scatter, fig.cap="Scatter plot over length of mouse vs weight (g).", out.width="50%", fig.show="hold"}
ggplot(subset(mice, week==10), aes(x=7.3+weight/20+rnorm(100,0,0.1), y=weight, group=id)) + geom_point() + xlab("length (cm)") + ylab("weight (g)")
```

For a time series or similar line graphs are useful.

```{r lineplot, fig.cap="Line plot over age of mouse vs weight (g) for three mice.", out.width="50%", fig.show="hold"}
ggplot(subset(mice, id %in% 17:19), aes(x=week, y=weight, group=id)) + geom_point() + geom_line()
```




Summary statistics for numeric data are usually divided into measures of location and spread.

<!--   - Interval: no absolute zero, meaningful to compute interval ratios. -->
<!--     Ex: time, temperature -->
<!--   - Ratio: absolute zero, meaningful to compute ratios. Ex. height, -->
<!--     weight, concentration -->
<!-- Often not necessary to distinguish between interval and ratio scale, can be more useful to divide the quantitative scales into -->


# Measures of location {#desc-04location}

- Mode: the most common value, can be computed also for categorical data
- Median: The value that divide the ordered data values into two equally sized groups. 50% of the values are below the median.

```{r normpdf, fig.width=2, out.width="20%"}
df <- data.frame(x=rnorm(200, 3.5, 2))
ggplot(df, aes(y=x)) + geom_boxplot() + theme_bw()
```

- Mean: the arithmetic mean, also called the average

For $n$ onservations $x_1, x_2, \dots, x_n$, the mean value is calculated as;
  
$$\bar x = \frac{x_1+x_2+\dots+x_n}{n} = \frac{1}{n}\displaystyle\sum_{i=1}^n x_i$$
Note, several very different distributions can still have the same mean value.

```{r mean35, warning=FALSE, message=FALSE, fig.width=7, fig.height=4, fig.cap="All these distributions have the same mean value, 3.50."}
#df <- data.frame(x=rnorm(200, 3.5, 2))
df$x <- df$x - mean(df$x) + 3.5
y <- exp(rnorm(200, 0, 1))
y <- y+3.5 - mean(y)
df$y=y

df$z <- rep(c(0,10), c(200*.65, 200*.35))

df$a <- rnorm(200, 3.5, 8)
df$a <- df$a - mean(df$a) + 3.5
b <- exp(rnorm(200, 2, 1))
df$b <- b+3.5 - mean(b)
c <- c(rnorm(150, 10, 3), rnorm(50,0,1))
c <- c + 3.5 - mean(c)
df$c <- c

plot(ggplot(melt(df), aes(x=value)) + geom_histogram(color="white", bins=30)  + facet_wrap(~variable, scale="free") + theme_bw())#+ geom_text(data=data.frame(variable=colnames(df), label=sprintf("mean: %.2f", colMeans(df))), aes(x=8,y=20,label=label))
``` 

# Measures of spread {#desc-05spread}

- Quartiles - the three values that divide the data values into four equally sized groups.

  - Q1. First quartile. 25% of the values are below Q1. Divides the values below the median into equally sized groups.
  - Q2.´ Second quartile. 50% of the values are below Q2. Q2 is the median.
  - Q3. Third quartile. 75% of the values are below Q3. Divides the values above the median into equally sized groups.

```{r boxplot1, fig.width=2, out.width="20%"}
ggplot(df, aes(y=x)) + geom_boxplot() + theme_bw()
```

- IQR: interquartile range: Q3 - Q1
- Variance, $\sigma^2$.
The variance is the mean squared distance from the mean value.
- Standard deviation, $\sigma = \sqrt{\sigma^2}$.

## Variance and standard deviation

The **variance** of a set of observations is their mean squared distance from the mean value;

$$\sigma^2 = \frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2.$$
the variance is measured in the square of the unit in which $x$ was measured. a commonly used measured on the same unit as $x$ is the  **standard deviation**, defined as the square root of the variance;

$$\sigma = \sqrt{\frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2}$$
The denominator $n$ is commonly replaced by $n-1$ and the **sample standard deviation** is calculated instead;

$$s = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (x_i - \bar x)^2}.$$
The latter formula is used if we regard the collection of observations $x_1, \dots, x_n$ as a **sample** drawn from a large **population** of possible observations.

I we want to describe the variance/standard deviation only of our set of observations, the former formula should be used, calculation of a population standard deviation $\sigma$ (i.e. we consider the set of observations to be the full population).

If instead, we want to estimate the variance of a larger population from which our smaller sample is drawn, we should calculate the sample standard deviation, $s$.

# Exercises: Descriptive statistics {.unnumbered #desc-exercises}

```{exercise, label="baby", echo=TRUE}
**Data summary**

Consider the below data and summarize each of the variables. There is no need to use R here, just use pen and paper, maybe use R as a calculator.

```{r }
#babies %>% transmute(id=id, smoker=c(smoker="yes", "non-smoker"="no")[smoker], "baby weight" = wtkg, "mother weight"=weightmother, "mother age"=age, gender=sample(c("F", "M"), nrow(babies), replace=TRUE), parity=parity) %>% filter((1:nrow(babies)) %in% sample(1:nrow(babies), 13)) %>%  arrange(smoker) %>% kable() %>% kable_styling("striped")

set.seed(1)
baby <- data.frame(id=1:10, 
                   smoker=rep(c("yes", "no"), each=5), 
                   "baby weight (kg)"=c(2.8,3.2,3.5,2.7,3.3, 3.7,3.3,4.3,3.2,3.0),
                   gender = sample(c("F", "M"), 10, replace=TRUE),
                   "mother weight (kg)" = c(64, 65, 60, 73, 59, 62, 52, 59, 65, 81), 
                   "mother age" = c(21, 27, 31, 32, 39, 26, 27, 21, 28, 33),
                   parity = c(2,1,2,0,3,0,2,0,1,4),
                   married = sample(c("yes", "no"), 10, replace=TRUE),
                   check.names=FALSE)
baby %>% kable() %>% kable_styling("striped", full_width = FALSE)
``` 


```{exercise, label="pill2", echo=TRUE}
**Amount of active substance**
  
The amount of active substance in a pill is stated by the manufacturer to be normally distributed with mean 12 mg and standard deviation 0.5 mg.
You take a sample of five pill and measure the amount of active substance to; 13.0, 12.3, 12.6, 12.5, 12.7 mg.

a) Compute the sample mean
b) Compute the sample variance
c) Compute the sample standard deviation
```

```{exercise, label="visualize", echo=TRUE}
**Distribution of body weight of a population of mice**

a) Download the [mice.csv](data/mice.csv) data set and take a first look at the data. How large is the data.frame, how many rows/columns? what are the column names and what is the data type of each column? How many mice are described in the data set?
Useful commands in R include `summary, View, dim, nrow, ncol, colnames`
b) The id column has identifiers for the mice and each mouse is described by many data points. Select a particular week, create a new data.frame with only weights of mice of this particlar number of weeks. Plot the distribution of weights in at least one way.
Useful commands in R include `subset, hist, density`
c) Summarize the entire data set using boxplots.
d) Can you think of another way to visualize the data set?
```

## Solutions: Descriptive statistics {-}  


\@ref(exr:baby)

* Smokers: 5 (50%) yes
* baby weight (kg) mean (sd): 3.3 (0.44)
* gender: 6 (60%) F
* mother weight (kg) mean(sd): 64 (8.5)
* mother age mean(sd): 28.5 (5.8)
* partity mean(sd): 1.5 (1.4) could also be handled as categorical (ordinal) and report frequencies and percentages.
* married: 4 (40%) yes

Did you compute standard deviations that were slightly different? Then you probably computed the sample standard deviation, which could actually be what you want to report. When do you want to compute sample standard deviation?


\@ref(exr:pill2)

a. 12.62
b. 0.067
c. 0.26
d. 0.22
e. 0.0028
