---
format:
  html:
    code-fold: false
execute:
  warning: false
  message: false
editor_options:
  chunk_output_type: console
---
# KNN demo: data splitting into train, validation and test {.unnumbered}

Let's build a KNN classifier to predict whether BMI is **greater than 30** using two variables from the diabetes dataset: `waist` and `hdl`. We will use **overall classification accuracy** throughout.

We will first reserve 20% of the observations as an untouched test set. The remaining 80% will be split equally into training and validation sets. We will select the best value of `k` using validation accuracy, then **refit on the whole 80% development set** and evaluate once on the test data.

Because KNN is distance-based, we standardize waist and HDL using the **training data only**. 

## Setup

```{r}
#| label: knn-setup
#| include: false
library(tidyverse)
library(splitTools)
library(kknn)

# Standardize predictors using ONLY the current training observations.
# Return datasets with the same outcome and feature names.
scale_using_train <- function(train, new) {
  feature_names <- c("waist", "hdl")
  means <- vapply(train[feature_names], mean, numeric(1))
  sds <- vapply(train[feature_names], sd, numeric(1))
  if (any(!is.finite(sds) | sds == 0)) {
    stop("Cannot standardize a feature with zero or undefined SD.")
  }
  for (feature in feature_names) {
    train[[feature]] <- (train[[feature]] - means[[feature]]) / sds[[feature]]
    new[[feature]] <- (new[[feature]] - means[[feature]]) / sds[[feature]]
  }
  list(train = train, new = new)
}
```

## Read and explore the data

```{r}
#| label: knn-read-data
#| fig-width: 7
#| fig-height: 5
#| fig-align: center

# Input data (same diabetes data used in the original demo).
input_diabetes <- read_csv("data/data-diabetes.csv")

# Convert units and calculate BMI.
inch2cm <- 2.54
pound2kg <- 0.45359237

data_diabetes <- input_diabetes %>%
  mutate(
    height = height * inch2cm / 100,
    waist = waist * inch2cm,
    weight = weight * pound2kg,
    BMI = weight / height^2,
    `BMI>30` = factor(if_else(BMI > 30, "1", "0"), levels = c("0", "1")),
    diabetic = factor(if_else(glyhb > 7, "Yes", "No"), levels = c("No", "Yes")),
    location = factor(location),
    frame = factor(frame),
    gender = factor(gender)
  )

data_input <- data_diabetes %>%
  select(`BMI>30`, waist, hdl) %>%
  drop_na()

count(data_input, `BMI>30`)
glimpse(data_input)

ggplot(data_input, aes(x = waist, y = hdl, fill = `BMI>30`)) +
  geom_point(shape = 21, alpha = 0.7, size = 2) +
  scale_fill_manual(values = c("0" = "blue3", "1" = "brown1")) +
  theme_bw() +
  theme(legend.position = "top")
```

## Split the data

```{r}
#| label: knn-data-split

# First isolate the SAME 20% test set used by the CV exercise.
randseed <- 123
splits <- partition(
  data_input$`BMI>30`,
  p = c(other = 0.80, test = 0.20),
  seed = randseed
)
data_other <- data_input[splits$other, ]
data_test <- data_input[splits$test, ]

# For this single-validation-split demo, divide the 80% development
# data equally into training and validation (40/40/20 overall).
development_splits <- partition(
  data_other$`BMI>30`,
  p = c(train = 0.50, valid = 0.50),
  seed = randseed
)
data_train <- data_other[development_splits$train, ]
data_valid <- data_other[development_splits$valid, ]

# Dimensions and class proportions in each subset.
bind_rows(
  mutate(count(data_train, `BMI>30`), split = "Training"),
  mutate(count(data_valid, `BMI>30`), split = "Validation"),
  mutate(count(data_test, `BMI>30`), split = "Test")
) %>%
  group_by(split) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()
```

## Train KNN for different values of k

```{r}
#| label: knn-train

# Learn scaling only from the training observations.
scaled <- scale_using_train(data_train, data_valid)
train_scaled <- scaled$train
valid_scaled <- scaled$new

# Test a manageable set of odd k values, all valid for this training set.
k_values <- seq(1, min(31, nrow(train_scaled)), by = 2)
validation_accuracy <- numeric(length(k_values))

for (i in seq_along(k_values)) {
  model <- kknn(
    `BMI>30` ~ ., train = train_scaled, test = valid_scaled,
    k = k_values[i], kernel = "rectangular", scale = FALSE
  )
  predictions <- fitted(model)  # Predicted class labels.
  validation_accuracy[i] <- mean(predictions == valid_scaled$`BMI>30`)
}
```

## Choose the best k using validation accuracy

```{r}
#| label: fig-knn-best-parameters
#| fig-cap: "Validation accuracy as a function of the number of neighbors (k)."
#| fig-height: 4.8
#| fig-width: 7
#| fig-align: center

plot(
  k_values, validation_accuracy,
  type = "b", pch = 16, col = "#2578A8",
  xlab = "Number of neighbors (k)", ylab = "Validation accuracy"
)
k_best <- k_values[which.max(validation_accuracy)]
abline(v = k_best, lty = 2, col = "grey50")
cat("Best k (single validation split):", k_best, "\n")
cat("Best validation accuracy:", max(validation_accuracy), "\n")
```

## Refit the final model and evaluate on the test set

```{r}
#| label: knn-final-model

# Refit using ALL development observations (training + validation).
# Only now evaluate once on the previously untouched test set.
final_scaled <- scale_using_train(data_other, data_test)
model_final <- kknn(
  `BMI>30` ~ ., train = final_scaled$train, test = final_scaled$new,
  k = k_best, kernel = "rectangular", scale = FALSE
)
test_pred <- fitted(model_final)
test_accuracy <- mean(test_pred == final_scaled$new$`BMI>30`)
cat("Final test accuracy (single validation split):", test_accuracy, "\n")
```
