---
format:
  html:
    code-fold: false
execute:
  warning: false
  message: false
editor_options:
  chunk_output_type: console
---
# Exercise: cross-validation {.unnumbered}

::: {#exr-knn-cross-validation}

Modify the KNN demo to select the best number of neighbors (`k`) using **five-fold cross-validation** rather than one training/validation split.

1. Keep the **same 20% test set** as in the demo. Perform cross-validation only on the other 80% of the data.
2. For each candidate `k`, fit the classifier on four training folds and calculate **overall accuracy** on the remaining fold. Repeat for all five folds.
3. Calculate the **mean accuracy** and **standard deviation** across folds for each `k`, and plot mean CV accuracy against `k`.
4. Select the best `k` based on mean CV accuracy, refit KNN on all 80% of the development data, and evaluate its accuracy on the untouched 20% test set.
5. Compare your result with the single-validation-split demo: did you select a different `k`? How do the test accuracies compare? **Cross-validation does not guarantee a higher test accuracy.**

**Important:** KNN depends on distance. For every CV fold, calculate the scaling parameters from the **training folds only**, then apply that transformation to the validation fold. Do not use test observations when fitting the scaler or selecting `k`.

**Hint:** Use `splitTools::create_folds(..., k = 5, invert = TRUE)` to obtain validation-fold row indices. Training rows are the remaining observations. Keep the same candidate `k` values as in the demo.

:::

## Suggested solution {.unnumbered}

::: {.solution}
@exr-knn-cross-validation
:::

```{r}
#| label: exercise-knn-setup
#| include: false
library(tidyverse)
library(splitTools)
library(kknn)

# Standardize predictors using ONLY the current training observations.
# Return datasets with the same outcome and feature names.
scale_using_train <- function(train, new) {
  feature_names <- c("waist", "hdl")
  means <- vapply(train[feature_names], mean, numeric(1))
  sds <- vapply(train[feature_names], sd, numeric(1))
  if (any(!is.finite(sds) | sds == 0)) {
    stop("Cannot standardize a feature with zero or undefined SD.")
  }
  for (feature in feature_names) {
    train[[feature]] <- (train[[feature]] - means[[feature]]) / sds[[feature]]
    new[[feature]] <- (new[[feature]] - means[[feature]]) / sds[[feature]]
  }
  list(train = train, new = new)
}
```

### Read and prepare data

```{r}
#| label: exercise-knn-read-data
#| fig-width: 7
#| fig-height: 5
#| fig-align: center

# Input data (same diabetes data used in the original demo).
input_diabetes <- read_csv("data/data-diabetes.csv")

# Convert units and calculate BMI.
inch2cm <- 2.54
pound2kg <- 0.45359237

data_diabetes <- input_diabetes %>%
  mutate(
    height = height * inch2cm / 100,
    waist = waist * inch2cm,
    weight = weight * pound2kg,
    BMI = weight / height^2,
    `BMI>30` = factor(if_else(BMI > 30, "1", "0"), levels = c("0", "1")),
    diabetic = factor(if_else(glyhb > 7, "Yes", "No"), levels = c("No", "Yes")),
    location = factor(location),
    frame = factor(frame),
    gender = factor(gender)
  )

data_input <- data_diabetes %>%
  select(`BMI>30`, waist, hdl) %>%
  drop_na()

count(data_input, `BMI>30`)
glimpse(data_input)

ggplot(data_input, aes(x = waist, y = hdl, fill = `BMI>30`)) +
  geom_point(shape = 21, alpha = 0.7, size = 2) +
  scale_fill_manual(values = c("0" = "blue3", "1" = "brown1")) +
  theme_bw() +
  theme(legend.position = "top")
```

### Reserve the same test set

```{r}
#| label: exercise-knn-split

randseed <- 123
splits <- partition(
  data_input$`BMI>30`,
  p = c(other = 0.80, test = 0.20),
  seed = randseed
)
data_other <- data_input[splits$other, ]
data_test <- data_input[splits$test, ]

# Return validation indices. Their complements are training indices.
validation_folds <- create_folds(
  data_other$`BMI>30`, k = 5, invert = TRUE, seed = randseed
)

# Use the same candidate values as the demo.
k_values <- seq(1, min(31, nrow(data_other)), by = 2)

# Ensure each k is possible in every training fold.
smallest_train_size <- min(vapply(validation_folds,
  function(valid_idx) nrow(data_other) - length(valid_idx), integer(1)))
k_values <- k_values[k_values <= smallest_train_size]
```

### Evaluate every k across five folds

```{r}
#| label: knn-cross-validation

n_folds <- length(validation_folds)
cv_accuracy <- matrix(
  NA_real_, nrow = length(k_values), ncol = n_folds,
  dimnames = list(paste0("k=", k_values), paste0("Fold ", seq_len(n_folds)))
)

for (i in seq_along(k_values)) {
  for (f in seq_len(n_folds)) {
    valid_idx <- validation_folds[[f]]
    train_idx <- setdiff(seq_len(nrow(data_other)), valid_idx)
    fold_train <- data_other[train_idx, ]
    fold_valid <- data_other[valid_idx, ]

    # Fit preprocessing only on the four training folds.
    scaled <- scale_using_train(fold_train, fold_valid)
    model <- kknn(
      `BMI>30` ~ ., train = scaled$train, test = scaled$new,
      k = k_values[i], kernel = "rectangular", scale = FALSE
    )
    cv_accuracy[i, f] <- mean(fitted(model) == scaled$new$`BMI>30`)
  }
}

cv_mean <- rowMeans(cv_accuracy)
cv_sd <- apply(cv_accuracy, 1, sd)
cv_results <- data.frame(k = k_values, mean_accuracy = cv_mean, sd_accuracy = cv_sd)
print(cv_results)
```

### Choose k and evaluate the final model

```{r}
#| label: fig-knn-cv-results
#| fig-cap: "Mean five-fold cross-validation accuracy across values of k."
#| fig-width: 7
#| fig-height: 4.8

plot(
  k_values, cv_mean, type = "b", pch = 16, col = "#2578A8",
  xlab = "Number of neighbors (k)", ylab = "Mean CV accuracy"
)
k_best <- k_values[which.max(cv_mean)]
abline(v = k_best, col = "grey50", lty = 2)
cat("Best k (5-fold CV):", k_best, "\n")
cat("Mean CV accuracy:", max(cv_mean), "\n")
cat("SD across folds:", cv_sd[which.max(cv_mean)], "\n")

# Fit on all development observations; test is used only once.
final_scaled <- scale_using_train(data_other, data_test)
model_final <- kknn(
  `BMI>30` ~ ., train = final_scaled$train, test = final_scaled$new,
  k = k_best, kernel = "rectangular", scale = FALSE
)
test_accuracy <- mean(fitted(model_final) == final_scaled$new$`BMI>30`)
cat("Final test accuracy (5-fold CV):", test_accuracy, "\n")
```
