---
format:
  html:
    code-fold: show
execute:
  warning: false
  message: false
---

# Tidymodels demo {.unnumbered}

In the previous demonstration, we implemented KNN with separate training, validation, and test sets. We then extended this approach to cross-validation. Here, we repeat the KNN workflow using **tidymodels**, a collection of R packages designed to make machine learning workflows more consistent and reproducible.

In `tidymodels`, we define different parts of the analysis separately:

- A **recipe** describes how to prepare the data, including steps such as normalization or handling missing values.
- A **model specification** defines the algorithm and any hyperparameters we want to tune.
- A **workflow** combines the recipe and model, so the same preprocessing is applied whenever the model is fitted or used for prediction.
- **Resampling and tuning** allow us to compare hyperparameter values using cross-validation.

We will predict whether **BMI > 30** using `waist` and `hdl`, select the number of neighbors (\(k\)) using **stratified five-fold cross-validation**, and evaluate performance using **accuracy**. As before, we hold out 20% of the data for a final test, and use the remaining 80% for model development.

The key advantage of a workflow is that preprocessing is performed **within each training fold**. This means that the means and standard deviations used to normalize predictors are estimated without looking at the corresponding validation fold, helping prevent data leakage.

## Load libraries and prepare data

```{r}
#| label: tm-data

library(tidymodels)
library(kknn)
library(readr)
library(dplyr)

input_diabetes <- read_csv("data/data-diabetes.csv")

data_input <- input_diabetes %>%
  mutate(
    height = height * 2.54 / 100,
    waist = waist * 2.54,
    weight = weight * 0.45359237,
    BMI = weight / height^2,
    `BMI>30` = factor(if_else(BMI > 30, "1", "0"), levels = c("0", "1"))
  ) %>%
  select(`BMI>30`, waist, hdl) %>%
  tidyr::drop_na()

count(data_input, `BMI>30`)
```

## Split the data and create CV folds

We keep the test set untouched while selecting \(k\). The five folds are created **only from the development data**. Stratification helps preserve the outcome class proportions across splits.


```{r}
#| label: tm-split

set.seed(123)
data_split <- initial_split(data_input, prop = 0.8, strata = `BMI>30`)
data_other <- training(data_split)
data_test <- testing(data_split)

set.seed(123)
folds <- vfold_cv(data_other, v = 5, strata = `BMI>30`)
folds
```

## Define preprocessing with a recipe

KNN relies on distances, so predictors measured on different scales should be standardized. `step_normalize()` specifies this operation; it does not calculate means and standard deviations yet. Those values are learned when the recipe is fitted, separately for each CV training fold.

```{r}
#| label: tm-recipe

knn_recipe <- recipe(`BMI>30` ~ waist + hdl, data = data_other) %>%
  step_normalize(all_numeric_predictors())

knn_recipe
```

## Specify KNN and build a workflow

Here, `tune()` tells `tidymodels` that we want to choose the number of neighbors using cross-validation. We keep uniform neighbor voting (`weight_func = "rectangular"`) to match the manual KNN demonstration. The workflow then bundles the model and recipe together.

```{r}
#| label: tm-workflow

knn_spec <- nearest_neighbor(
  mode = "classification",
  neighbors = tune(),
  weight_func = "rectangular"
) %>%
  set_engine("kknn")

knn_workflow <- workflow() %>%
  add_recipe(knn_recipe) %>%
  add_model(knn_spec)

knn_workflow
```

## Tune k with five-fold cross-validation

We evaluate odd values of \(k\) from 1 to 31 using the same five folds for each candidate. `tune_grid()` fits the workflow on the training portion of each fold and evaluates it on the corresponding validation portion. Each candidate must be smaller than the smallest training-fold size.

```{r}
#| label: tm-tune

k_values <- seq(1, 31, by = 2)
k_grid <- tibble(neighbors = k_values)

set.seed(123)
knn_results <- tune_grid(
  knn_workflow,
  resamples = folds,
  grid = k_grid,
  metrics = metric_set(accuracy)
)

knn_results
```

## Examine cross-validation performance

`collect_metrics()` summarizes accuracy across folds for each value of \(k\). Its `mean` column gives average CV accuracy, while `std_err` is the standard error of that average. For the plot, we calculate the **standard deviation across folds** as `std_err * sqrt(n)`, where `n` is the number of non-missing fold estimates. These error bars illustrate variation between folds; they are **not** confidence intervals for future performance.

```{r}
#| label: tm-summarize

cv_results <- collect_metrics(knn_results) %>%
  mutate(sd = std_err * sqrt(n))

cv_results %>%
  select(neighbors, mean, sd, n)
```

```{r}
#| label: fig-tm-knn-cv
#| fig-cap: "Mean five-fold cross-validation accuracy for different values of k; error bars show ±1 SD across folds."
#| fig-width: 8
#| fig-height: 4.8

best_k <- select_best(knn_results, metric = "accuracy")

ggplot(cv_results, aes(x = neighbors, y = mean)) +
  geom_errorbar(aes(ymin = pmax(0, mean - sd),
                    ymax = pmin(1, mean + sd)),
                width = 0.35, color = "grey55") +
  geom_line(color = "#2578A8", linewidth = 1) +
  geom_point(color = "#2578A8", size = 2) +
  geom_vline(xintercept = best_k$neighbors, linetype = "dashed",
             color = "#CE6547") +
  scale_x_continuous(breaks = k_values) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Number of neighbors (k)", y = "Mean CV accuracy") +
  theme_bw(base_size = 14)
```

## Select k and evaluate on the test set

`select_best()` chooses the value of \(k\) with the highest mean CV accuracy. `finalize_workflow()` inserts that value into our model specification. Finally, `last_fit()` refits the complete workflow (including normalization) using **all development data (80%)** and evaluates it **once** on the held-out test set (20%).

```{r}
#| label: tm-final

best_k

final_workflow <- finalize_workflow(knn_workflow, best_k)
final_results <- last_fit(
  final_workflow,
  split = data_split,
  metrics = metric_set(accuracy)
)

collect_metrics(final_results)
```

## Swapping KNN for a support vector machine

One advantage of `tidymodels` is that we can **reuse the same recipe and cross-validation folds** with a different algorithm. For example, to fit a **linear support vector machine (SVM)**, we change the model specification and replace it in the workflow. Here we use a fixed cost value (`cost = 1`) for illustration, rather than tuning it.

```{r}
#| label: tm-svm

# Requires the kernlab package: install.packages("kernlab")
svm_spec <- svm_linear(mode = "classification", cost = 1) %>%
  set_engine("kernlab")

svm_workflow <- knn_workflow %>%
  update_model(svm_spec)

set.seed(123)
svm_results <- fit_resamples(
  svm_workflow,
  resamples = folds,
  metrics = metric_set(accuracy)
)

collect_metrics(svm_results)
```

The **data split, normalization, folds, and accuracy metric stay the same**. Only the model changes. We use `fit_resamples()` here because we are evaluating one fixed SVM configuration; if we wanted to choose its `cost` using CV, we could use `tune()` and `tune_grid()` as we did for KNN. These CV results are for demonstration, not a final test-set evaluation.

## Take-home message

The modeling logic is the same as in our manual KNN demonstration: **split the data → use CV to tune \(k\) → refit on all development data → evaluate on the test set**. `tidymodels` makes these steps easier to organize and helps ensure that preprocessing is learned from training data only.

> **Comparison with the manual demo:** The two approaches may produce slightly different results because `rsample` and `splitTools` use different splitting procedures. Using the same random seed does **not** guarantee identical observations in each split. For a strict numerical comparison, reuse exactly the same row indices, candidate values of \(k\), and preprocessing steps.
