---
title: "Session regression I"
output: github_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path="session-regression-I-files/figures/")
```


### Learning outcomes
- to be able to fit regression line and interpret regression model
- to be able to use `lm` function in R for model fitting, obtaining confidence interval and predictions

## Linear Regression 
Linear regression is a very simple approach for supervised learning, when numerical outcome and numerical exposure are concerned. The method of linear regression is used to estimate the best-fitting straight line to describe the association. With linear regression we can answer questions incl.:

- is there a relationship between exposure (e.g. body weigth) and outcome (e.g. plasma volume)?
- how strong is the relationship between the two variables
- what will be a predicted value of the outcome given a new set of exposure values
- how accurately can we predict outcome

### Example data

Example data contain the body weight and plasma volume for eight healthy men. 
```{r}

weight <- c(58, 70, 74, 63.5, 62.0, 70.5, 71.0, 66.0) # body weight (kg)
plasma <- c(2.75, 2.86, 3.37, 2.76, 2.62, 3.49, 3.05, 3.12) # plasma volume (liters)
```

Scatter plot of the data shows that high plasma volume tends to be associated with high weight and *vice verca*. Linear regrssion gives the equation of the straight line that best describes how the outcome changes (increase or decreases) with a change of exposure variable (in red)
```{r, fig-reg, echo=F, fig.align="center", fig.height=5, fig.width=5}

plot(weight, plasma, pch=19, las=1, xlab = "body weight [kg]", ylab="plasma volume [l]")
abline(lm(plasma~weight), col="red") # regression line

```

The equation of the regression line is: 

$$y=beta_0 + beta_1x$$



### Estimating the Coefficients
### Assessing the Accuracy of the Coefficient Estimates
### Asesssing the Accuracy of the Model


```{r reg, echo=F}

weight=c(58, 70, 74, 63.5, 62.0, 70.5, 71.0, 66.0) # body weight (kg)
plasma <- c(2.75, 2.86, 3.37, 2.76, 2.62, 3.49, 3.05, 3.12) # plasma volume (liters)

# correlation 
cor(weight, plasma)

# linear regression
x <- weight
y <- plasma
model <- lm(y~x)
summary(model)

# prediction
y.hat <- predict(model)

n=length(x)
RSS <- sum((y.hat-y)^2)
RSE <- sqrt(RSS/(n-2))

s <- sqrt((sum((y-mean(y))^2) - (model$coefficients[2])^2*sum((x-mean(x))^2))/(n-2))
se.1 <-s/sqrt(sum((x-mean(x))^2))
se.1

se.0 <- s*sqrt((1/n + mean(x)^2/(sum((x-mean(x))^2))))
se.0


#             Estimate Std. Error t value Pr(>|t|)  
# (Intercept)  0.08572    1.02400   0.084   0.9360  
# x            0.04362    0.01527   2.857   0.0289 *
```



## Multiple linear regression
### Estimating the Regression Coefficients
### Estimating coefficients
### Relationship between the response and predictors 
### Model fit 
### Predictions 
### Qualitative predictors 
### Interaction terms
### Non-linear transformation of the predictors 
### Potential problems: non-linearity, collinearity 
### Logistic regression
