## Learning outcomes {-}

- be aware of data types
- compute measures of location, including mean and median
- compute measures of spread, including quantiles, variance and standard deviation
- understand the concept random sample
- compute population mean and variance
- compute sample mean and variance


```{r setup, include=FALSE}
library(tidyverse)
require(kableExtra)
library(reshape2)
library(ggplot2)
library(ggbeeswarm)
require(knitr)
#require(UsingR)
knitr::opts_chunk$set(fig.width=3.5, fig.height=3.5, echo = FALSE, error=FALSE, warnings=FALSE, dpi=600, fig.path = "Rfigures/desc_")
options(digits=2)
```

# Data types {#desc-01datatypes}

Data can be divided into different types; categorical and quantitative (numeric). How to summarize and analyze your data depends on the type.

Categorical data types are divided into;

* Nominal: named. 
*Ex: dead/alive, healthy/sick, WT/mutant, AA/Aa/aa, male/female, red/green/blue*
* Ordinal: named and ordered. 
*Ex: pain (weak, moderate, severe), AA/Aa/aa, very young/young/middle age/old/very old, grade I, II, III, IV*

Quantitative data types are divided into;

* Discrete: finite or countable infinite values. *Ex. counts, number of cells, number of reads*
* Continuous: infinitely many uncountable values. *Ex. height, weight, concentration*



# Categorical data {#desc-02categorical}

## Data summarization and plots

Categorical data can be summarized by counting the number of observations of each category and summarizing in a frequency table or bar plot. Alternatively, the proportions (or percentages) of each category can be calculated.

```{r simulatemice}
set.seed(101)
n <- 100
week <- 5:24
m <- length(week)

mice0 <- data.frame(id=1:n, gender=sample(c('male', 'female'), n, replace = TRUE)) %>% mutate(male=1*(gender=="male"), b0=rnorm(n, 15.5, 1), b1=rnorm(n, 0.5, 0.1), b2=rnorm(n, 1.5,0.5), b3=rnorm(n, 0.3, 0.1), a0 = rnorm(n, 44, 4))
mice <- mice0 %>% slice(rep(1:n(), each=m)) %>% mutate(week=rep(week, n))
mice <- mice %>% mutate(s=rnorm(n*m, 0, 0.2)) %>% mutate(weight=b0+b1*week+b2*male+b3*male*week+s)

##mice <- data.frame(id=factor(rep(1:n, each=m)), week=rep(week, n), gender=rep(sample(c('male', 'female'), n, replace = TRUE), each=m)) %>% mutate(male=1*(gender=="male"), b0=rep(rnorm(n, 15.5, 1), each=m), b1=rep(rnorm(n, 0.5, 0.1), each=m), b2=rep(rnorm(n, 1.5,0.5), each=m), b3=rep(rnorm(n, 0.3, 0.1), each=m), s=rnorm(n*m, 0, 0.2)) %>% mutate(weight=b0+b1*week+b2*male+b3*male*week+s)
mice0 <- mice 
mice <- mice %>% dplyr::select(id, gender, male, week, weight)
write.csv(mice, file="data/mice.csv", quote=FALSE)
```

```{example, label="10mice", echo=TRUE}

**Ten lab mice**
  
Observe gender and weight of your ten lab mice and summarize.
```

If you want to follow this example, you can download the data here; [mice.csv](data/mice.csv). You get the subset used in this example by the following commands; 

```{r read mice, eval=FALSE, echo=TRUE}
## first read the full data set into R
mice <- read.csv("mice.csv")
```


```{r m10subset, echo=TRUE}
## Then extract the specific subset used in this example
m10 <- subset(mice, subset=week==5 & id %in% 1:10, select = c(id, gender, weight))
```

In this example we have only ten observations (mice) and the full data can actually be shown in a table.

```{r m10}
m10  %>% kable(caption="Gender and weight of 10 mice.", row.names = FALSE) %>% kable_styling(full_width = FALSE)
```

We are interested in the gender distribution in our group of mice. Count the frequency of male/female mice and summarize in a table. Also, the fraction or percentage can be useful.

```{r tabm10, fig.show="hold", out.width="49%", fig.cap="The number of male and female mice"}
kable(m10 %>% group_by(gender) %>% tally() %>% mutate("percent (%)"=n/sum(n)*100) , caption="The number of male and female mice.") %>% kable_styling(full_width = FALSE)
```

The frequencies can also be shown in a barplot.

```{r m10barplot, fig.show="hold", out.width="49%", fig.cap="The number of male and female mice shown in barplots generated using ggplot and basic R graphics.", echo=TRUE}
ggplot(m10, aes(x=gender)) + geom_bar()
barplot(table(m10$gender))
```


```{example, label="lefthandedness", echo=TRUE}
**Left handedness**

You are interested in whether left-handedness is associated to a disease you study and observe left-handedness among 30 patients as well as among 40 healthy controls;
``` 

```{r lefthand, echo=FALSE}
set.seed(1123)
patients <- sample(c("L", "R"), 30, prob=c(0.3, 0.7), replace=TRUE)
controls <- sample(c("L", "R"), 40, prob=c(0.1, 0.9), replace=TRUE)
hand <- rbind(data.frame(group="patient", handedness=patients),
              data.frame(group="control", handedness=controls))
```

patients: {`r patients`}

controls: {`r controls`}

Summarized as 

```{r table1}
hand %>% group_by(group) %>% dplyr::summarize(Total=n(), n = sum(handedness=="L"),  f=100*mean(handedness=="L")) %>% transmute(group, "Total n"=Total, "Left handed (%)"=sprintf("%i (%d%%)", n, f)) %>% kable(caption="Summary of left handedness among patients and controls.")
```

or in a contingency table;

```{r}
#hand %>% group_by(group, handedness) %>% dplyr::summarize(n=n()) %>% spread(handedness, n)
addmargins(table(hand$group, hand$handedness)) %>% kable(caption="Cross table or contingency table of group and left/right handedness.")
```

Data can be summarized in barplots in several ways;

```{r barlefthand, fig.show="hold", out.width="33%", fig.cap="Left-handedness in patient and control groups.", echo=TRUE}
## Using ggplot to create barplots
ggplot(hand, aes(x=group, fill=handedness)) + geom_bar()
ggplot(hand, aes(x=group, fill=handedness)) + geom_bar(position="dodge")
ggplot(hand, aes(x=group, fill=handedness)) + geom_bar(position="fill") + ylab("Fraction")

## Using basic R graphics to create barplots
tab <- table(hand$handedness, hand$group)
barplot(tab)
barplot(tab, beside=TRUE)
tabperc <- tab
tabperc[,1] <- 100*tab[,1]/sum(tab[,1])
tabperc[,2] <- 100*tab[,2]/sum(tab[,2])
barplot(tabperc)
```

Sometimes a barplot is plotted using polar coordinates, i.e. a pie chart.

```{r pie, fig.cap="A pie chart over he number of controls that are left/right handed."}
ggplot(hand %>% filter(group=="control"), aes(x="", fill=handedness)) + geom_bar(width=1) + coord_polar("y", start=0)
```


# Quantitative data {#desc-03quantitative}

Quantitative data (both discrete and continuous) can be visualized and summarized in many ways. Common plots include histograms, density plots, boxplots and scatter plots. summary statistics include mean, median, quartile, variance, standard deviation and median absolute deviation.

## Histogram

A histogram bins the data and counts the number of observations that fall into each bin.

```{r sum10dice, fig.show="hold", out.width="50%", fig.cap="Throw 10 dice and count the total number of dots. Repeat the experiment 1000 times. This histogram summarize the results, i.e. the total number of dots when throwing 10 dice.", echo=TRUE}
sum10dice <- replicate(1000, sum(sample(1:6, 10, replace=TRUE)))
hist(sum10dice)
ggplot(data.frame(sum10dice), aes(x=sum10dice)) + geom_histogram(binwidth = 5, center=32.5, color="white")
```

```{r hist, fig.cap="Histogram over weight of 2000 5 weeks old mice, colored according to gender.", out.width="50%", fig.show="hold", echo=TRUE}
micew5 <- mice %>% filter(week==5)
ggplot(micew5, aes(x=weight, fill=gender)) + geom_histogram(bins=15, color="white")
hist(micew5$weight[micew5$gender=="female"], col="orange", breaks=15, main="Weight of mice at 5 weeks", xlab="Weight (g)")
hist(micew5$weight[micew5$gender=="male"], col="cornflowerblue", breaks=15, add=TRUE)
```

## Density plot

A density plot is like a smoothed histogram where the total area under the curve is set to 1. A density plot is an approximation of a distribution.

```{r dicedens, fig.cap="Density plot over the total number of dots when throwing 10 dice.", out.width="50%", echo=TRUE}
ggplot(data.frame(sum10dice), aes(x=sum10dice)) + geom_density()
```

## Boxplot

A boxplot, also called a box-and-whisker plot, shows a box covering 50% of the data and the center line is located at the **median**. The median value is a value such that 50% of the measurements are below the median.

The whiskers extend to the most extreme data point or at most 1.5 times the length of the box. (Note that 1.5 is the default in both ggplot and basic R graphics, but it is also a number that can be changed.) Any measurements further out are shown as outliers. A boxplot is based on both measures of location and of spread, more about this in the following chapters.

```{r boxplot, fig.cap="Boxplot over weight of 100 5 weeks old mice, divided according to gender.", out.width="50%", fig.show="hold", echo=TRUE}
boxplot(weight~gender, data=micew5)
ggplot(micew5, aes(x=gender, y=weight)) + geom_boxplot() 
```

## Beeswarm plot

Instead of, or in addition to, a boxplot, it might be useful to actually show all the measurements.

This can be done in a 1D scatter plot, a so called strip plot.

```{r stripplot, fig.cap="Strip plot over weight of 20 5 weeks old mice, divided according to gender.", out.width="50%", fig.show="hold"}
ggplot(micew5, aes(x=gender, y=weight)) + geom_point() 
```
As some measurements are close to each other it can be difficult to interpret such a plot, in a beeswarm plot the data points are scattered a bit along the x-axis. In a beeswarm plot the x-position is not meaningful, it is just there to make more data points visible.

```{r beeswarm, fig.cap="Beeswarm plot over weight of 20 5 weeks old mice, divided according to gender. Plot to the left generated using the function beeswarm in the R-package beeswarm, figure to the right generated using ggplot and the R-package ggbeeswarm.", out.width="50%", fig.show="hold", echo=TRUE}
library(beeswarm)
beeswarm(weight~gender, data=micew5)
ggplot(micew5, aes(x=gender, y=weight)) + geom_beeswarm(cex=2.5) 
```

## Scatter plot

To study the relationship between two variables scatter plots are useful;

```{r scatter, fig.cap="Scatter plot over length of mouse vs weight (g).", out.width="50%", fig.show="hold"}
ggplot(subset(mice, week==10), aes(x=7.3+weight/20+rnorm(100,0,0.1), y=weight, group=id)) + geom_point() + xlab("length (cm)") + ylab("weight (g)")
```

For a time series or similar line graphs are useful.

```{r lineplot, fig.cap="Line plot over age of mouse vs weight (g) for three mice.", out.width="50%", fig.show="hold"}
ggplot(subset(mice, id %in% 17:19), aes(x=week, y=weight, group=id)) + geom_point() + geom_line()
```




Summary statistics for numeric data are usually divided into measures of location and spread.

# Measures of location {#desc-04location}

- Mode: the most common value, can be computed also for categorical data
- Median: The value that divide the ordered data values into two equally sized groups. 50% of the values are below the median.


```{r normpdf, fig.width=2, out.width="20%"}
set.seed(1111)
df <- data.frame(x=rnorm(200, 3.5, 2))
ggplot(df, aes(y=x)) + geom_boxplot() + theme_bw()
```

- Mean: the arithmetic mean, also called the average

For $n$ onservations $x_1, x_2, \dots, x_n$, the mean value is calculated as;
  
$$\bar x = \frac{x_1+x_2+\dots+x_n}{n} = \frac{1}{n}\displaystyle\sum_{i=1}^n x_i$$
Note, several very different distributions can still have the same mean value.

```{r mean35, warning=FALSE, message=FALSE, fig.width=7, fig.height=4, fig.cap="All these distributions have the same mean value, 3.50."}
#df <- data.frame(x=rnorm(200, 3.5, 2))
df$x <- df$x - mean(df$x) + 3.5
y <- exp(rnorm(200, 0, 1))
y <- y+3.5 - mean(y)
df$y=y

df$z <- rep(c(0,10), c(200*.65, 200*.35))

df$a <- rnorm(200, 3.5, 8)
df$a <- df$a - mean(df$a) + 3.5
b <- exp(rnorm(200, 2, 1))
df$b <- b+3.5 - mean(b)
c <- c(rnorm(150, 10, 3), rnorm(50,0,1))
c <- c + 3.5 - mean(c)
df$c <- c

plot(ggplot(melt(df), aes(x=value)) + geom_histogram(color="white", bins=30)  + facet_wrap(~variable, scale="free") + theme_bw())#+ geom_text(data=data.frame(variable=colnames(df), label=sprintf("mean: %.2f", colMeans(df))), aes(x=8,y=20,label=label))
``` 

# Measures of spread {#desc-05spread}

- Quartiles - the three values that divide the data values into four equally sized groups.

  - Q1. First quartile. 25% of the values are below Q1. Divides the values below the median into equally sized groups.
  - Q2.´ Second quartile. 50% of the values are below Q2. Q2 is the median.
  - Q3. Third quartile. 75% of the values are below Q3. Divides the values above the median into equally sized groups.

```{r boxplot1, fig.width=2, out.width="20%"}
ggplot(df, aes(y=x)) + geom_boxplot() + theme_bw()
```

- IQR: interquartile range: Q3 - Q1
- Variance, $\sigma^2$.
The variance is the mean squared distance from the mean value.
- Standard deviation, $\sigma = \sqrt{\sigma^2}$.

## Variance and standard deviation

The **variance** of a set of observations is their mean squared distance from the mean value;

$$\sigma^2 = \frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2.$$
the variance is measured in the square of the unit in which $x$ was measured. a commonly used measured on the same unit as $x$ is the  **standard deviation**, defined as the square root of the variance;

$$\sigma = \sqrt{\frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2}$$
The denominator $n$ is commonly replaced by $n-1$ and the **sample standard deviation** is calculated instead;

$$s = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (x_i - \bar x)^2}.$$


The latter formula is used if we regard the collection of observations $x_1, \dots, x_n$ as a **sample** drawn from a large **population** of possible observations.

If we want to describe the variance/standard deviation only of our set of observations, the former formula should be used, calculation of a population standard deviation $\sigma$ (i.e. we consider the set of observations to be the full population).

If instead, we want to estimate the variance of a larger population from which our smaller sample is drawn, we should calculate the sample standard deviation, $s$.



<!--   - Interval: no absolute zero, meaningful to compute interval ratios. -->
<!--     Ex: time, temperature -->
<!--   - Ratio: absolute zero, meaningful to compute ratios. Ex. height, -->
<!--     weight, concentration -->
<!-- Often not necessary to distinguish between interval and ratio scale, can be more useful to divide the quantitative scales into -->


# Exercises: Descriptive statistics {.unnumbered #desc-exercises}

```{exercise, label="baby", echo=TRUE}
**Data summary**

Consider the below data and summarize each of the variables. There is no need to use R here, just use pen and paper, maybe use R as a calculator.

```{r }
#babies %>% transmute(id=id, smoker=c(smoker="yes", "non-smoker"="no")[smoker], "baby weight" = wtkg, "mother weight"=weightmother, "mother age"=age, gender=sample(c("F", "M"), nrow(babies), replace=TRUE), parity=parity) %>% filter((1:nrow(babies)) %in% sample(1:nrow(babies), 13)) %>%  arrange(smoker) %>% kable() %>% kable_styling("striped")

set.seed(1)
baby <- data.frame(id=1:10, 
                   smoker=rep(c("yes", "no"), each=5), 
                   "baby weight (kg)"=c(2.8,3.2,3.5,2.7,3.3, 3.7,3.3,4.3,3.2,3.0),
                   gender = sample(c("F", "M"), 10, replace=TRUE),
                   "mother weight (kg)" = c(64, 65, 60, 73, 59, 62, 52, 59, 65, 81), 
                   "mother age" = c(21, 27, 31, 32, 39, 26, 27, 21, 28, 33),
                   parity = c(2,1,2,0,3,0,2,0,1,4),
                   married = sample(c("yes", "no"), 10, replace=TRUE),
                   check.names=FALSE)
baby %>% kable() %>% kable_styling("striped", full_width = FALSE)
``` 



```{exercise, label="pill2", echo=TRUE}
**Amount of active substance**
  
The amount of active substance in a pill is stated by the manufacturer to be normally distributed with mean 12 mg and standard deviation 0.5 mg.
You take a sample of five pill and measure the amount of active substance to; 13.0, 12.3, 12.6, 12.5, 12.7 mg.

a) Compute the sample mean
b) Compute the sample variance
c) Compute the sample standard deviation
```

```{exercise, label="visualize", echo=TRUE}
**Distribution of body weight of a population of mice**

a) Download the [mice.csv](data/mice.csv) data set and take a first look at the data. How large is the data.frame, how many rows/columns? what are the column names and what is the data type of each column? How many mice are described in the data set?
Useful commands in R include `summary, View, dim, nrow, ncol, colnames`
b) The id column has identifiers for the mice and each mouse is described by many data points. Select a particular week, create a new data.frame with only weights of mice at this particular week. Plot the distribution of weights in at least one way.
Useful commands in R include `subset, hist, density`
c) Summarize the entire data set using boxplots.
d) Can you think of another way to visualize the data set?
```

```{exercise, label="PBC", echo=TRUE}
**Mayo Clinic Primary Biliary Cholangitis Data**
  
Start by loading the dataset pbc, from the survival package. To read more about the data set you can read the help text.
```

```{r}
library(survival)
data(pbc)
?pbc
```

a) Take a first look at the data, e.g. using `summary`. All variables, excep sex, are coded numerically. Is this correct? If not, which variables are really categorical? Use `factor` to change them into categorical variables. Run `summary` on the dataset again.
b) There are packages available for easily summarizing a data set, one of them is `tableone`. Install the package, load it and apply it to the pbc data set.
Hint: the main function is called `CreateTableOne`
c) Plot the copper and bili values, both separately using an appropriate plot and together in a scatter plot. 


## Solutions: Descriptive statistics {-}  


\@ref(exr:baby)

* Smokers: 5 (50%) yes
* baby weight (kg) mean (sd): 3.3 (0.44)
* gender: 6 (60%) F
* mother weight (kg) mean(sd): 64 (8.5)
* mother age mean(sd): 28.5 (5.8)
* partity mean(sd): 1.5 (1.4) could also be handled as categorical (ordinal) and report frequencies and percentages.
* married: 4 (40%) yes

Did you compute standard deviations that were slightly different? Then you probably computed the sample standard deviation, which could actually be what you want to report. When do you want to compute sample standard deviation?


\@ref(exr:pill2)

a. 12.62
b. 0.067
c. 0.26
d. 0.22
e. 0.0028
