---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Diabetes data

Before we continue with the descriptive statistics, let's introduce an example data set. 
We will be looking at the data collected in a study to understand the prevalence of obesity, diabetes, and other cardiovascular risk factors in central Virginia, USA.

The data is available as part of `faraway` package. 403 African Americans were interviewed in a study to understand the prevalence of obesity, diabetes, and other cardiovascular risk factors in central Virginia. Available variables include: 

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: true

library(tidyverse)
library(kableExtra)
library(faraway)

# Diabetes data from faraway() package
df <- diabetes

c1 <- colnames(df)
c2 <- c("Subject ID", "Total Cholesterol [mg/dL]", "Stabilize Glucose [mg/dL]", 
        "High Density Lipoprotein [mg/dL]", "Cholesterol / HDL Ratio", "Glycosolated Hemoglobin [%]", 
        "County: Buckingham or Louisa", "age [years]", "gender", 
        "height [in]", "weight [lb]", "frame: small, medium or large", 
        "First Systolic Blood Pressure", "First Diastolic Blood Pressure", 
        "Second Systolic Blood Pressure", "Second Diastolic Blood Pressure", 
        "waist [in]", "hip [in]", 
        "Postprandial Time [min] when labs were drawn")

tbl <- data.frame(Abbreviation = c1, Description = c2)

tbl %>%
  kbl() %>% kable_styling()
  
```

And the first few observations are: 

```{r}

#| echo: false
#| warning: false
#| message: false
#| include: true

library(tidyverse)
library(kableExtra)
library(faraway)

# preview
glimpse(diabetes)

```

Further: 

- Glycosolated hemoglobin greater than 7.0 is usually taken as a positive diagnosis of diabetes.
- We can calculate BMI as $BMI = 703 \times (weight \; [lb] \; / (height \;[in])^2)$ and define obesity as $BMI \ge 30$.
- Alternatively, we can first convert pounds (lb) to kilograms (kg) by multiplying by 0.45 and inches (in) to meters (m) by multiplying by 0.0254 and then calculating BMI as $BMI = (weight \; [kg] \; / (height \;[m])^2)$

In `R` we can add diabetes and obesity status (yes/no) and display first few measurements across all the variables as below: 
```{r}

# add obesity and diabetes variables
inch2m <- 2.54/100
pound2kg <- 0.45
data_diabetes <- diabetes %>%
  mutate(height  = height * inch2m, height = round(height, 2)) %>% 
  mutate(waist = waist * inch2m) %>%  
  mutate(weight = weight * pound2kg, weight = round(weight, 2)) %>%
  mutate(BMI = weight / height^2, BMI = round(BMI, 2)) %>% 
  mutate(obese= cut(BMI, breaks = c(0, 29.9, 100), labels = c("No", "Yes"))) %>% 
  mutate(diabetic = ifelse(glyhb > 7, "Yes", "No"), diabetic = factor(diabetic, levels = c("No", "Yes")))
  
# preview data
glimpse(data_diabetes)

```


We can now use descriptive statistics to understand the diabetes data set more. 




