---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Exercises

```{r, include=T, message=FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(ggbeeswarm)
library(gridExtra)
```


::: {#exr-baby}
## Babies

Consider the below data and summarize each of the variables. Note, there is no need to use R here, just use pen and paper, maybe use R as a calculator.

```{r, echo=F}
set.seed(1)
baby <- data.frame(id=1:10, 
                   smoker=rep(c("yes", "no"), each=5), 
                   "baby weight (kg)"=c(2.8,3.2,3.5,2.7,3.3, 3.7,3.3,4.3,3.2,3.0),
                   gender = sample(c("F", "M"), 10, replace=TRUE),
                   "mother weight (kg)" = c(64, 65, 60, 73, 59, 62, 52, 59, 65, 81), 
                   "mother age" = c(21, 27, 31, 32, 39, 26, 27, 21, 28, 33),
                   parity = c(2,1,2,0,3,0,2,0,1,4),
                   married = sample(c("yes", "no"), 10, replace=TRUE),
                   check.names=FALSE)
```

```{r}
baby %>% kable() %>% 
  kable_styling("striped", full_width = FALSE)
```
:::

::: {#exr-pill}
## Active substance

The amount of active substance in a pill is stated by the manufacturer to be normally distributed with mean 12 mg and standard deviation 0.5 mg. You take a sample of five pills and measure the amount of active substance to be: 13.0, 12.3, 12.6, 12.5, 12.7 mg.

a) Compute the sample mean
b) Compute the sample variance
c) Compute the sample standard deviation
:::

::: {#exr-mice}
## Mice
a) Download the [mice.csv](data/mice.csv) data set and take a first look at the data. How large is the data.frame, how many rows/columns? What are the column names and what is the data type of each column? How many mice are described in the data set?
Useful commands in R include `summary, View, dim, nrow, ncol, colnames`
b) The id column has identifiers for the mice and each mouse is described by many data points. Select a particular week, create a new data.frame with only weights of mice at this particular week. Plot the distribution of weights in at least one way.
Useful commands in R include `subset, hist, density`
c) Summarize the entire data set using box plots
d) Can you think of another way to visualize the data set?
:::


::: {#exr-pbc}
## Primary Biliary Cholangitis (PBC)
  
Load the dataset pbc, from the survival() package. To read more about the data set you can read the help text.

  
```{r}
library(survival)
data(pbc)
?pbc
```

a) Take a first look at the data, e.g. using `summary`. All variables, excep sex, are coded numerically. Is this correct? If not, which variables are really categorical? Use `factor` to change them into categorical variables. Run `summary` on the dataset again.
b) There are packages available for easily summarizing a data set, one of them is `tableone`. Install the package, load it and apply it to the pbc data set.
Hint: the main function is called `CreateTableOne`
c) Plot the copper and bili values, both separately using an appropriate plot and together in a scatter plot. 
::: 


## Solutions: Descriptive statistics {-}  

::: {.solution}
@exr-baby

* Smokers: 5 (50%) yes
* baby weight (kg) mean (sd): 3.3 (0.44)
* gender: 6 (60%) F
* mother weight (kg) mean(sd): 64 (8.5)
* mother age mean(sd): 28.5 (5.8)
* partity mean(sd): 1.5 (1.4) could also be handled as categorical (ordinal) and report frequencies and percentages.
* married: 4 (40%) yes

Did you compute standard deviations that were slightly different? Then you probably computed the sample standard deviation, which could actually be what you want to report. When do you want to compute sample standard deviation?
:::


::: {.solution}
@exr-baby

a. 12.62
b. 0.067
c. 0.26
d. 0.22
e. 0.0028

:::





