---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Measures of spread
```{r, include=T, message=FALSE, warning=FALSE}
# load libraries
library(tidyverse)
library(magrittr)
library(kableExtra)
library(ggplot2)
library(gridExtra)
```


## Quartiles 

Quartiles are the three values that divide the data values into four equally sized groups.

  - Q1. First quartile. 25% of values are below Q1. Divides the values below the median into equally sized groups.
  - Q2. Second quartile. 50% of values are below Q2 and 50% are above Q2. Q2 is the median that we have seen before.
  - Q3. Third quartile. 75% of values are below Q3. Divides the values above the median into equally sized groups.


::: {#exm-mice-quartiles}

## Lab mice (cont.)

Going back to the lab mice example. What are the three quartiles of mice weight at week 5? 

:::

```{r, warning=F, message=F}
# read in mice data
data.mice <- read_csv("data/mice.csv")
data.mice <- data.mice %>%
  mutate(weight = round(weight,2))

# narrow data set to week 5 measurements
data.mice.week5 <- data.mice %>%
  filter(week == 5)

# calculate quartiles
data.mice.week5 %>%
  summarise(x = quantile(weight, c(0.25, 0.5, 0.75))) %>%
  print()

```

We can check if the values agree with the box plot. 
```{r}
#| fig-cap: Bar plot of mice weights at week 5
#| label: fig-mice-boxplot
#| fig-cap-location: margin
#| fig-width: 4
#| fig-height: 4

# define generic ggplot theme
font.size <- 12
my.ggtheme <- theme(axis.title = element_text(size = font.size), 
        axis.text = element_text(size = font.size), 
        legend.text = element_text(size = font.size), 
        legend.title = element_blank(), 
        legend.position = "top") 

data.mice.week5 %>%
  ggplot(aes(x = "", y = weight)) + 
  geom_boxplot() + 
  xlab("") + 
  ylab("weight (g)") + 
  theme_bw() + 
  my.ggtheme

```


## Variance and standard deviation

The **variance** of a set of observations is their mean squared distance from the mean value:

$$\sigma^2 = \frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2.$$ {#eq-sigma2}
The variance is measured in the square of the unit in which $x$ was measured. Another common measure using the same unit as $x$ is **standard deviation**, defined as the square root of the variance:

$$\sigma = \sqrt{\frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2}$$ {#eq-sigma}


Typically, we regard the collection of observations $x_1, \dots, x_n$ as a **sample** drawn from a large **population** of possible observations. It has been shown theoretically that we obtain a better sample estimate of the **population standard deviation** if we divide by $(n-1)$. So the denominator $n$ is commonly replaced by $n-1$ and the **sample standard deviation** is calculated instead:

$$s = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (x_i - \bar x)^2}.$$ {#eq-s}

To reiterate: 

- we calculate standard deviation, $\sigma$, when we consider our observations to be entire population (@eq-sigma)
- we calculate sample standard deviation, $s$, when we consider our observations a random sample from a larger population, as a better estimate of the standard deviation of the larger population (@eq-s)

::: {#exm-mice-sd}

## Lab mice (cont.)

Let's calculate variance, standard deviation and sample standard deviation for mice weight at week 5. 

:::

```{r}

# get weights measurements at week 5
weight.week5 <- data.mice.week5 %>%
  pull(weight)

# number of observations
n <- length(weight.week5)
  
# calculate mean (arithmetic)
weight.mean <- mean(weight.week5)

# calculate variance following variance equation
sigma2 <- (sum((weight.week5 - weight.mean)^2))/(n)

# calculate standard deviation following standard deviation equation
sigma <- sqrt((sum((weight.week5 - weight.mean)^2))/(n))

# calculate sample standard deviation following sample standard deviation equation
s <- sqrt((sum((weight.week5 - weight.mean)^2))/(n-1))
# sd(weight.week5) # we can alternatively use sd() function

# collect results
v.name <- c("sigma2", "sigma", "s")
v.values <- c(sigma2, sigma, s)
results <- data.frame(stats = v.name, value = v.values)
print(results)

```




