---
title: "From linear models to GLMs"
# subtitle: "From linear models to GLMs"
# author: Olga Dethlefsen
format: 
  revealjs:
    slide-number: true
    theme: [default, styles.scss]
    chalkboard: 
      buttons: true
    footer: "Statistical Methods for Life Sciences"
  html:
    code-fold: false
editor_options: 
  chunk_output_type: console
bibliography: references.bib

---

```{r}
#| message: false
#| warning: false
#| include: false

# load libraries
library(tidyverse)
library(magrittr)
library(faraway)
library(kableExtra)
library(ggplot2)
library(rmarkdown)
library(gridExtra)
library(ggiraphExtra)
library(latex2exp)

font.size <- 16
col.blue.light <- "#a6cee3"
col.blue.dark <- "#1f78b4"
my.ggtheme <- 
  theme_bw() + 
  theme(axis.title = element_text(size = font.size), 
        axis.text = element_text(size = font.size), 
        legend.text = element_text(size = font.size), 
        legend.title = element_blank(), 
        legend.position = "top") 
        
# input data
input_diabetes <- read_csv("data/data-diabetes.csv")

# clean data
inch2cm <- 2.54
pound2kg <- 0.45
data_diabetes <- input_diabetes %>%
  mutate(height  = height * inch2cm / 100, height = round(height, 2)) %>% 
  mutate(waist = waist * inch2cm) %>% 
  mutate(weight = weight * pound2kg, weight = round(weight, 2)) %>%
  mutate(BMI = weight / height^2, BMI = round(BMI, 2)) %>%
  mutate(obese= cut(BMI, breaks = c(0, 29.9, 100), labels = c("No", "Yes"))) %>%
  mutate(diabetic = ifelse(glyhb > 7, "Yes", "No"), diabetic = factor(diabetic, levels = c("No", "Yes"))) %>%
  mutate(location = factor(location)) %>%
  mutate(frame = factor(frame)) %>%
  mutate(gender = factor(gender))

```


## Plan for today {.central}
<br>


Let's

::: incremental

- see how to expand linear models to model outcomes other than continous
- and practice some common cases together

:::


## Linear models describe expected outcomes
<br>

In the linear model, we describe the expected value of the outcome:

$$
E(Y_i \mid \mathbf{x}_i)
=
\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}
$$

This works well for many continuous outcomes.

. . .

<br>

But not all outcomes are continuous.

## When the outcome is not continuous
<br>

Examples:

- binary: disease yes/no, response yes/no
- counts: number of events, number of cases
- proportions: successes out of trials

. . .

<br>

For these outcomes, the expected value has restrictions:

::: incremental

- probabilities must be between 0 and 1
- expected counts must be positive

:::

## Restrictions
*Example*

<br>


-   Fitting a regression line to binary data yields predicted values that could take any value, including $<0$,


```{r}
#| label: log-example
#| message: false
#| warning: false
#| fig-cap: "Example of fitting linear model to binary data, to model obesity status coded as 1 (Yes) and 0 (No) with waist variable. Linear model does not fit the data well in this case and the predicted values can take any value, not only 0 and 1."
#| fig-cap-location: bottom
#| fig-height: 10

font.size <- 24
col.blue.light <- "#a6cee3"
col.blue.dark <- "#1f78b4"
my.ggtheme <- theme_bw() + 
  theme(axis.title = element_text(size = font.size),
        axis.text = element_text(size = font.size),
        legend.text = element_text(size = font.size),
        legend.title = element_blank(),
        legend.position = "top") 
        

data_diabetes %>%
  mutate(obese = as.numeric(obese) - 1) %>%
  ggplot(aes(y=obese, x=waist)) +
  geom_jitter(width=0, height = 0.1, size = 3, alpha = 0.8) +
  geom_smooth(method="lm", se=FALSE, color=col.blue.dark) + 
  my.ggtheme

```


## The GLM idea

We keep the same **linear predictor**:

$$
\eta_i =
\beta_0+\beta_1x_{1i}+\cdots+\beta_px_{pi}.
$$

But instead of modelling the expected outcome directly, we connect it to
the linear predictor through a **link function**:

$$
g(\mu_i)=\eta_i,
$$

where

$$
\mu_i=E(Y_i\mid\mathbf{x}_i).
$$

## Link functions {.center}

<br>

| Outcome              | Model               | Distribution | Link     |
| -------------------- | ------------------- | ------------ | -------- |
| Continuous           | Linear regression   | Normal       | identity |
| Binary / proportions | Logistic regression | Binomial     | logit    |
| Counts               | Poisson regression  | Poisson      | log      |
| Positive continuous  | Gamma regression    | Gamma        | log      |



# Logistic regression

## Logistic regression {.smaller}

<br>

::: columns
::: {.column width="50%"}
- Since the response variable takes only two values (Yes/No), we use a
  **logistic regression** model to model the **probability of being obese
  (Yes)**.

- Let

  $$
  p_i = P(Y_i = 1)
  $$

  denote the probability that individual $i$ is obese.

- For a binary response, we assume

  $$
  Y_i \sim \operatorname{Binomial}(1,p_i),
  $$

  which is equivalent to a Bernoulli distribution.

- Logistic regression uses the **logit link** to relate this probability to
  the linear predictor:

  $$
  \log\left(\frac{p_i}{1-p_i}\right)
  =
  \beta_0+\beta_1x_i.
  $$

- Equivalently, the probability can be written as

  $$
  p_i =
  \frac{\exp(\beta_0+\beta_1x_i)}
  {1+\exp(\beta_0+\beta_1x_i)}.
  $$

  This ensures that the fitted probabilities always lie between 0 and 1.
:::

::: {.column width="5%"}
:::

::: {.column width="45%"}
```{r}
#| fig-height: 8

p1 <- data_diabetes %>%
  mutate(obese = as.numeric(obese) - 1) %>%
  ggplot(aes(y=obese, x=waist)) +
  geom_jitter(width=0, height = 0.05, alpha = 0.7, size = 3) +
  my.ggtheme

p2 <- data_diabetes %>%
  ggplot(aes(x = obese, y = waist, fill = obese)) + 
  geom_boxplot() + 
  scale_fill_brewer(palette = "Set2") + 
  my.ggtheme

# 
p1
```
:::
:::


## Logistic regression
*R* 
<br>

In R we can use `glm()` function to fit GLM models:

``` r
# fit logistic regression model
logmodel_1 <- glm(obese ~ waist, family = binomial(link="logit"), data = data_diabetes)

# print model summary
print(summary(logmodel_1))
```

```{r}
#| code-fold: false
#| collapse: true
#| message: false
#| warning: false
#| fig-cap: "Fitted logistic model to diabetes data given the 130 study participants and using waist as explantatory variable to model obesity status."

# re-code obese status from Yes/No to 1/0
data_diabetes <- 
  data_diabetes %>%
  mutate(obese = as.numeric(obese) - 1)

# fit logistic regression model
logmodel_1 <- glm(obese ~ waist, family = binomial(link="logit"), data = data_diabetes)

# print model summary
print(summary(logmodel_1))

```


. . .

## Logistic regression

*R* <br>

```{r}
#| code-fold: false
#| collapse: true
#| message: false
#| warning: false
#| fig-cap: "Fitted logistic model to diabetes data given the 130 study participants and using waist as explantatory variable to model obesity status."
#| fig-width: 10
#| fig-height: 6
#| fig-align: center


font.size <- 14
col.blue.light <- "#a6cee3"
col.blue.dark <- "#1f78b4"
my.ggtheme <- theme(axis.title = element_text(size = font.size), 
        axis.text = element_text(size = font.size), 
        legend.text = element_text(size = font.size), 
        legend.title = element_blank(), 
        legend.position = "top") + 
        theme_bw()

# plot
ggPredict(logmodel_1) + 
  my.ggtheme

```

## Logistic regression
*Hypothesis testing* 
<br>

::: columns
::: {.column width="40%"}
- As in linear regression, we can test whether a predictor is associated with the outcome by testing

  $$
  H_0:\beta_j=0.
  $$
-  The Wald test compares how far the estimated coefficient is from the value specified under the null hypothesis, relative to its standard error: $$W = \frac{\hat\beta-\beta}{e.s.e.(\hat\beta)} \sim N(0,1)$$
:::

::: {.column width="5%"}
:::

::: {.column width="55%"}
``` r
# fit logistic regression model
logmodel_1 <- glm(obese ~ waist, family = binomial(link="logit"), data = data_diabetes)
summary(logmodel_1)
```

```{r}
summary(logmodel_1)
```
:::
:::



## Logistic regression
*Deviance (rather than R^2)* 
<br>
<br>

- **Deviance** measures how well a model fits the observed data relative to a
  perfectly fitting (saturated) model. Smaller deviance indicates better fit.

- Deviance can also be used to compare **nested models**, for example to test
  whether adding an explanatory variable improves the model.

- In a **likelihood ratio test**, the test statistic is the difference in
  deviance between the reduced and full models:

  $$
  \Delta D = D_{\text{reduced}} - D_{\text{full}}.
  $$

  Under the null hypothesis,

  $$
  \Delta D \approx \chi^2_{df},
  $$

  where $df$ is the difference in the number of estimated parameters between
  the two models.


## Logistic regression
*Odds ratio* 
<br>

::: columns
::: {.column width="40%"}

- In logistic regression, we often interpret coefficients by exponentiating them:

  $$
  OR = e^{\hat{\beta}}.
  $$

- This gives an **odds ratio**.

- For example, if $\hat{\beta}=0.18$,

  $$
  e^{0.18}\approx1.2.
  $$

  This means that for each one-unit increase in `waist`, the odds of obesity are multiplied by about 1.2.

:::

::: {.column width="5%"}
:::

::: {.column width="55%"}
``` r
# fit logistic regression model
logmodel_1 <- glm(obese ~ waist, family = binomial(link="logit"), data = data_diabetes)

# print model summary
print(summary(logmodel_1))
```

```{r}
summary(logmodel_1)
```
:::
:::


# Common cases

-   Let's go over some common linear models cases together.
