---
title: "Descriptive statistics"
author: "Eva Freyhult"
date: "2019-10-04"
institute: NBIS, SciLifeLab
output:
  xaringan::moon_reader:
    css: ["default", "metropolis"]
    encoding: 'UTF-8'
    self_contained: false
    # lib_dir: libs
    nature:
      ratio: '4:3'
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      # slideNumberFormat: "Biostatistics. %current%/%total%"
---
class: spaced


```{r setup, include=FALSE}
library(tidyverse)
require(kableExtra)
library(reshape2)
library(ggplot2)
require(knitr)
require(UsingR)
knitr::opts_chunk$set(fig.width=3.5, fig.height=3.5, echo = FALSE, cache=TRUE, error=FALSE, warnings=FALSE, dpi=600)
options(digits=2)
```

# Descriptive statistics

.pull-left[
```{r table1_0, results="asis"}
library(tableone)
data(pbc)

## Vector of variables to summarize
myVars <- c("status", "age", "sex", "hepato",
          "edema", "bili", "chol", "albumin", "copper")
## Vector of categorical variables that need transformation
catVars <- c("status", "trt", "ascites", "hepato",
             "spiders", "edema", "stage")
## Create a TableOne object
tab <- CreateTableOne(vars = myVars, data = pbc %>% filter(!is.na(trt)) %>% mutate(treat=paste("Treatment", trt)), strata = "treat", factorVars = catVars, test=FALSE) %>% print(quote=TRUE, printToggle=FALSE)
rownames(tab) = gsub('\\"', '', rownames(tab))
colnames(tab) = gsub('\\"', '', colnames(tab))
tab %>% kable(caption="Table 1. Summary statistics for the Mayo Clinic Primary Biliary Cholangitis Data (data(pbc) from the R-package survival).") %>% kable_styling(font_size=10) %>% gsub("font-size: initial !important;", "font-size: inherited !important;", .)
```
]

.pull-right[
```{r pbc0, out.width="45%", warning=FALSE}
library(ggbeeswarm)
pbc <- pbc %>% mutate(hepato=factor(hepato), trt=factor(trt), status=factor(c("censored", "transplant", "dead")[status+1], levels=c("censored", "transplant", "dead")))
#pbc %>% ggplot(aes(x=copper)) + geom_histogram(bins=20, color="white")
pbc %>% ggplot(aes(x=bili)) + geom_histogram(bins=20, color="white")
pbc %>% mutate(hepato=factor(hepato)) %>% ggplot(aes(x=copper, y=bili, color=hepato)) + geom_point() + scale_x_log10() + scale_y_log10() + theme(legend.position = c(0,1), legend.justification = c(0,1))
pbc %>% ggplot(aes(x=bili, color=status)) + geom_density() + scale_x_log10() + theme(legend.position = c(1,1), legend.justification = c(1,1))
pbc %>% ggplot(aes(x=trt, fill=status)) + geom_bar() + theme(legend.position = c(1,1), legend.justification = c(1,1))
pbc %>% filter(!is.na(trt)) %>% ggplot(aes(x=factor(trt), y=bili, fill=sex)) + geom_boxplot() + scale_y_log10() + xlab("Treatment") + theme(legend.position = c(0,1), legend.justification = c(0,1))
pbc  %>% ggplot(aes(x=status, y=bili, color=sex)) + geom_beeswarm() + scale_y_log10() + theme(legend.position = c(0.25,1), legend.justification = c(0,1))
```
]

---

# Data types

- Categorical
  - Nominal: named.
Ex: dead/alive, healthy/sick, WT/mutant, AA/Aa/aa, male/female, red/green/blue
  - Ordinal: named and ordered. Ex: pain (weak, moderate, severe), AA/Aa/aa, very young/young/middle age/old/very old, grade I, II, III, IV

Reported as frequencies, proportions, summarized using mode

--

- Quantitative (numeric)
<!--   - Interval: no absolute zero, meaningful to compute interval ratios. -->
<!--     Ex: time, temperature -->
<!--   - Ratio: absolute zero, meaningful to compute ratios. Ex. height, -->
<!--     weight, concentration -->
<!-- Often not necessary to distinguish between interval and ratio scale, can be more useful to divide the quantitative scales into -->
  - Discrete: finite or countable infinite values. Ex. counts, number of cells, number of reads
  - Continuous: infinitely many uncountable values.  Ex. height, weight, concentration

Useful summary statistics include mean, median, variance, standard deviation.

---
layout: false
# Data summations and plots

.pull-left[
```{r table1, results="asis"}
library(tableone)
data(pbc)

## Vector of variables to summarize
myVars <- c("status", "age", "sex", "hepato",
          "edema", "bili", "chol", "albumin", "copper")
## Vector of categorical variables that need transformation
catVars <- c("status", "trt", "ascites", "hepato",
             "spiders", "edema", "stage")
## Create a TableOne object
tab <- CreateTableOne(vars = myVars, data = pbc %>% filter(!is.na(trt)) %>% mutate(treat=paste("Treatment", trt)), strata = "treat", factorVars = catVars, test=FALSE) %>% print(quote=TRUE, printToggle=FALSE)
rownames(tab) = gsub('\\"', '', rownames(tab))
colnames(tab) = gsub('\\"', '', colnames(tab))
tab %>% kable(caption="Table 1. Summary statistics for the Mayo Clinic Primary Biliary Cholangitis Data (data(pbc) from the R-package survival).") %>% kable_styling(font_size=10) %>% gsub("font-size: initial !important;", "font-size: inherited !important;", .)
```
]

.pull-right[
```{r pbc, out.width="45%", warning=FALSE}
library(ggbeeswarm)
pbc <- pbc %>% mutate(hepato=factor(hepato), trt=factor(trt), status=factor(c("censored", "transplant", "dead")[status+1], levels=c("censored", "transplant", "dead")))
#pbc %>% ggplot(aes(x=copper)) + geom_histogram(bins=20, color="white")
pbc %>% ggplot(aes(x=bili)) + geom_histogram(bins=20, color="white")
pbc %>% mutate(hepato=factor(hepato)) %>% ggplot(aes(x=copper, y=bili, color=hepato)) + geom_point() + scale_x_log10() + scale_y_log10() + theme(legend.position = c(0,1), legend.justification = c(0,1))
pbc %>% ggplot(aes(x=bili, color=status)) + geom_density() + scale_x_log10() + theme(legend.position = c(1,1), legend.justification = c(1,1))
pbc %>% ggplot(aes(x=trt, fill=status)) + geom_bar() + theme(legend.position = c(1,1), legend.justification = c(1,1))
pbc %>% filter(!is.na(trt)) %>% ggplot(aes(x=factor(trt), y=bili, fill=sex)) + geom_boxplot() + scale_y_log10() + xlab("Treatment") + theme(legend.position = c(0,1), legend.justification = c(0,1))
pbc  %>% ggplot(aes(x=status, y=bili, color=sex)) + geom_beeswarm() + scale_y_log10() + theme(legend.position = c(0.25,1), legend.justification = c(0,1))
```
]

---
layout: true
# Categorical data
---

Categorical data can be summarized by counting the number of observations of each category and summarizing in a frequency table or bar plot.

Alternatively, the proportions (or percentages) of each category can be calculated.

---

## Example: left-handedness

Is left-handedness associated to a disease you study? Observe left-handedness among 30 patients as well as among 40 healthy controls;

```{r lefthand, echo=FALSE}
set.seed(1123)
patients <- sample(c("L", "R"), 30, prob=c(0.3, 0.7), replace=TRUE)
controls <- sample(c("L", "R"), 40, prob=c(0.1, 0.9), replace=TRUE)
hand <- rbind(data.frame(group="patient", handedness=patients),
              data.frame(group="control", handedness=controls))
```

patients: {`r patients`}

controls: {`r controls`}

--

.pull-left[
```{r table1hand}
hand %>% group_by(group) %>% dplyr::summarize(Total=n(), n = sum(handedness=="L"),  f=100*mean(handedness=="L")) %>% transmute(group, "Total n"=Total, "Left handed (%)"=sprintf("%i (%d%%)", n, f)) %>% kable(caption="Summarize using counts or percentages.")
```
]

--

.pull-right[
```{r}
addmargins(table(hand$group, hand$handedness)) %>% kable(caption="Cross table or contingency table.")
```
]

---

Data can be summarized in barplots in several ways;

```{r barlefthand1, out.width="32%"}
## Using ggplot to create barplots
ggplot(hand, aes(x=group, fill=handedness)) + geom_bar()
```
--

```{r barlefthand2, out.width="32%"}
ggplot(hand, aes(x=group, fill=handedness)) + geom_bar(position="dodge")
```
--

```{r barlefthand3, out.width="32%"}
ggplot(hand, aes(x=group, fill=handedness)) + geom_bar(position="fill") + ylab("Fraction")
```

---
layout: true
# Quantitative data

---


Quantitative data (both discrete and continuous) can be visualized in many ways!

---


## Histogram

```{r simulatemice}
set.seed(101)
n <- 100
week <- 5:24
m <- length(week)

mice0 <- data.frame(id=1:n, gender=sample(c('male', 'female'), n, replace = TRUE)) %>% mutate(male=1*(gender=="male"), b0=rnorm(n, 15.5, 1), b1=rnorm(n, 0.5, 0.1), b2=rnorm(n, 1.5,0.5), b3=rnorm(n, 0.3, 0.1), a0 = rnorm(n, 44, 4))
mice <- mice0 %>% slice(rep(1:n(), each=m)) %>% mutate(week=rep(week, n))
mice <- mice %>% mutate(s=rnorm(n*m, 0, 0.2)) %>% mutate(weight=b0+b1*week+b2*male+b3*male*week+s)

##mice <- data.frame(id=factor(rep(1:n, each=m)), week=rep(week, n), gender=rep(sample(c('male', 'female'), n, replace = TRUE), each=m)) %>% mutate(male=1*(gender=="male"), b0=rep(rnorm(n, 15.5, 1), each=m), b1=rep(rnorm(n, 0.5, 0.1), each=m), b2=rep(rnorm(n, 1.5,0.5), each=m), b3=rep(rnorm(n, 0.3, 0.1), each=m), s=rnorm(n*m, 0, 0.2)) %>% mutate(weight=b0+b1*week+b2*male+b3*male*week+s)
mice0 <- mice 
mice <- mice %>% dplyr::select(id, gender, male, week, weight)
```


```{r hist, out.width="50%", fig.show="hold"}
micew5 <- mice %>% filter(week==5)

ggplot(micew5, aes(x=weight)) + geom_histogram(bins=15, color="white")
```
---

## Histogram

```{r histcol, out.width="50%", fig.show="hold"}
ggplot(micew5, aes(x=weight, fill=gender)) + geom_histogram(bins=15, color="white") + theme(legend.position = c(1,1), legend.justification = c(1,1))
```
---

## Density plot

```{r dens, out.width="50%", fig.show="hold"}
ggplot(micew5, aes(x=weight)) + geom_density()
```

---

## Boxplot

```{r boxplot, out.width="50%", fig.show="hold"}
ggplot(micew5, aes(x=gender, y=weight)) + geom_boxplot()
```
---

## Boxplot

```{r boxplotq, out.width="50%", fig.show="hold"}
q <- quantile(micew5 %>% filter(gender=="male") %>% pull("weight"), c(0.25,0.5,0.75))
bp <- ggplot(micew5, aes(x=gender, y=weight)) + geom_boxplot() 
bp + annotate("text", x=2, y=q[3], label="3rd quartile", hjust=0.5, vjust=-0.5) + 
  annotate("text", x=2, y=q[1], label="1st quartile", hjust=0.5, vjust=-0.5) + 
  annotate("text", x=2, y=q[2], label="median", hjust=0.5, vjust=-0.5) + 
  annotate("text", x=1.5, y=q[2], label="IQR", hjust=0.5, vjust=1, angle=90)
```

---

## Strip plot

```{r stripplot, out.width="50%", fig.show="hold"}
ggplot(micew5, aes(x=gender, y=weight)) + geom_point() 
```

---

## Beeswarm plot

```{r beeswarm, out.width="50%", fig.show="hold"}
ggplot(micew5, aes(x=gender, y=weight)) + geom_beeswarm(cex=2.5) 
```

---

## Scatter plot

Relationship between two variables

```{r scatter, out.width="50%", fig.show="hold"}
ggplot(subset(mice, week==10), aes(x=7.3+weight/20+rnorm(100,0,0.1), y=weight, group=id)) + geom_point() + xlab("length (cm)") + ylab("weight (g)")
```

---

## Line graph

```{r lineplot,  out.width="50%", fig.show="hold"}
ggplot(subset(mice, id %in% 17:19), aes(x=week, y=weight, group=id)) + geom_point() + geom_line()
```

---

Summary statistics for numeric data are usually divided into measures of location and spread.

```{r hist2, out.width="50%", fig.show="hold"}
micew5 <- mice %>% filter(week==5)

ggplot(micew5, aes(x=weight)) + geom_histogram(bins=15, color="white")
```

---
layout: true
# Measures of location

---

- **Mode**: the most common value, can be computed also for categorical data
- **Median**: The value that divide the ordered data values into two equally sized groups. 50% of the values are below the median.

```{r normpdf, fig.width=2, out.width="20%"}
df <- data.frame(x=rnorm(200, 3.5, 2))
ggplot(df, aes(y=x)) + geom_boxplot() + theme_bw()
```

- **Mean**: the arithmetic mean, also called the average

For $n$ onservations $x_1, x_2, \dots, x_n$, the mean value is calculated as;
  
$$\bar x = \frac{x_1+x_2+\dots+x_n}{n} = \frac{1}{n}\displaystyle\sum_{i=1}^n x_i$$

Note, several very different distributions can still have the same mean value.


---
layout: false

# Mean value 3.50

Note, several very different distributions can still have the same mean value!

```{r mean35, warning=FALSE, message=FALSE, fig.width=7, fig.height=4, fig.cap="All these distributions have the same mean value, 3.50."}
#df <- data.frame(x=rnorm(200, 3.5, 2))
df$x <- df$x - mean(df$x) + 3.5
y <- exp(rnorm(200, 0, 1))
y <- y+3.5 - mean(y)
df$y=y

df$z <- rep(c(0,10), c(200*.65, 200*.35))

df$a <- rnorm(200, 3.5, 8)
df$a <- df$a - mean(df$a) + 3.5
b <- exp(rnorm(200, 2, 1))
df$b <- b+3.5 - mean(b)
c <- c(rnorm(150, 10, 3), rnorm(50,0,1))
c <- c + 3.5 - mean(c)
df$c <- c

plot(ggplot(melt(df), aes(x=value)) + geom_histogram(color="white", bins=30)  + facet_wrap(~variable, scale="free") + theme_bw())#+ geom_text(data=data.frame(variable=colnames(df), label=sprintf("mean: %.2f", colMeans(df))), aes(x=8,y=20,label=label))
``` 

---
layout: true
class: left

# Measures of spread

---
**Quartiles** - the three values that divide the data values into four equally sized groups.

.left-column[
  ```{r boxplot1, fig.width=2, out.width="80%"}
ggplot(df, aes(y=x)) + geom_boxplot() + theme_bw()
```
]
.right-column[
* Q1. First quartile. 25% of the values are below Q1. Divides the values below the median into equally sized groups.
* Q2. Second quartile. 50% of the values are below Q2. Q2 is the median.
* Q3. Third quartile. 75% of the values are below Q3. Divides the values above the median into equally sized groups.
]

--

**IQR** interquartile range: Q3 - Q1

**Variance** $\sigma^2$.
The variance is the mean squared distance from the mean value.

**Standard deviation** $\sigma = \sqrt{\sigma^2}$.



---

## Variance and standard deviation

The **variance** of a set of observations is their mean squared distance from the mean value;

$$\sigma^2 = \frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2.$$

--
<!-- the variance is measured in the square of the unit in which $x$ was measured. a commonly used measured on the same unit as $x$ is t -->

The  **standard deviation**, defined as the square root of the variance;

$$\sigma = \sqrt{\frac{1}{n} \sum_{i=1}^n (x_i - \bar x)^2}$$
--

The denominator $n$ is commonly replaced by $n-1$ and the **sample standard deviation** is calculated instead;

$$s = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (x_i - \bar x)^2}.$$

The latter formula is used if we regard the collection of observations $x_1, \dots, x_n$ as a **sample** drawn from a large **population** of possible observations.

<!-- If we want to describe the variance/standard deviation only of our set of observations, the former formula should be used, calculation of a population standard deviation $\sigma$ (i.e. we consider the set of observations to be the full population). -->

<!-- If instead, we want to estimate the variance of a larger population from which our smaller sample is drawn, we should calculate the sample standard deviation, $s$. -->
