---
title: "Describing outcomes with probability models"
subtitle: "Statistical Methods for Life Sciences"
format:
  html:
    toc: true
    number-sections: true
---


```{r}
#| label: setup
#| include: false
set.seed(2026)
mu <- 170
sigma <- 7
```


In the previous section’s BMI exercise, knowing that 50% of adults have a BMI above 25 was not enough to calculate the probability that a randomly selected adult has a BMI above 30.

To calculate this probability, we need more information about how BMI values are distributed above 25. A probability model is one way to describe this distribution.

## Probability models and probability distributions

A **probability model** is a mathematical description of a random process. It specifies possible outcomes, their probabilities and the assumptions about how observations arise.

A **random variable** $X$ assigns a numerical value to each outcome. Its **probability distribution** describes how probability is distributed across those values.

A continuous distribution is described by a **probability density function**, $f(x)$. 

While discrete distribution is described by a **probability mass function**, p(x).

## A normal model for individual heights

For instance, our model might assume that we independently select students from a population where heights are normally distributed with mean 170 cm and SD 7 cm. Let's see how to calculate the probability that a student is taller than 180cm.

Given our model, the probability distribution of the height (X) of one randomly selected student is:

$$
X \sim N(\mu,\sigma^2),
\qquad \mu=170,\quad \sigma=7.
$$
where **mean** $\mu$ and **variance** $\sigma^2$ are distribution parameters. 

- **Expected value:** $E[X]=\mu$, the long-run average.
- **Variance:** $\operatorname{Var}(X)=E[(X-\mu)^2]$, the expected squared distance from the mean.
- **Standard deviation (SD):** $\sigma=\sqrt{\operatorname{Var}(X)}$ the square root of variance, expressed in the original units

So here: 


$$
E[X]=170\text{ cm},
$$

$$
\operatorname{Var}(X)
=E[(X-\mu)^2]
=\sigma^2
=7^2
=49\text{ cm}^2,
$$

$$
\operatorname{SD}(X)
=\sqrt{\operatorname{Var}(X)}
=7\text{ cm}.
$$

And the **probability density function**, $f(x)$ describing distribution has a form:

$$
f(x)=
\frac{1}{\sigma\sqrt{2\pi}}
\exp\left[-\frac{(x-\mu)^2}{2\sigma^2}\right].
$$


```{r}
#| label: fig-height-normal
#| echo: false
#| fig-cap: "Probability density function of the height of a student."
#| fig-height: 4

mu <- 170
sigma <- 7
cutoff <- 180

curve(
  dnorm(x, mean = mu, sd = sigma),
  from = 140, to = 200,
  xlab = "Height (cm)", ylab = "Density",
  lwd = 2
)

xx <- seq(cutoff, 200, length.out = 300)


curve(
  dnorm(x, mean = mu, sd = sigma),
  add = TRUE, lwd = 2
)

abline(v = mu, lty = 2)
```


For a continuous distribution, the density function $f(x)$ gives the **height of the density curve** at a particular value $x$. A higher density means that values around $x$ are relatively more common than values around points with lower density.

Note that normality is a modelling assumption, not a universal property of biological measurements.

## Continuous outcomes: probability is area

The density itself is not a probability. For continuous variables, probabilities are represented by **areas under the density curve**.

The probability that $X$ falls between $a$ and $b$ is the area under the density curve between these values:

$$
P(a < X \leq b)=\int_a^b f(x)\,dx.
$$

The integral adds up the density across the interval.

The total area under the density curve is 1:

$$
\int_{-\infty}^{\infty} f(x)\,dx = 1.
$$



For example, for our height model, the probability density function (PDF) is: 

$$
f(x)=
\frac{1}{7\sqrt{2\pi}}
\exp\left[-\frac{(x-170)^2}{98}\right].
$$


The probability $P(X>180)$ is the area under the normal density curve to the right of 180, shown below:

```{r}
#| label: fig-height-tail
#| echo: false
#| fig-cap: "Probability density function of the height of a student. The shaded area represents the probability of height above 180 cm."
#| fig-height: 4

mu <- 170
sigma <- 7
cutoff <- 180

curve(
  dnorm(x, mean = mu, sd = sigma),
  from = 140, to = 200,
  xlab = "Height (cm)", ylab = "Density",
  lwd = 2
)

xx <- seq(cutoff, 200, length.out = 300)

polygon(
  c(cutoff, xx, 200),
  c(0, dnorm(xx, mean = mu, sd = sigma), 0),
  col = "#8FC4CF", border = NA
)

curve(
  dnorm(x, mean = mu, sd = sigma),
  add = TRUE, lwd = 2
)

abline(v = cutoff, lty = 2)
```

In theory, we calculate this area by integration:

$$
P(X>180)=\int_{180}^{\infty}f(x)\,dx.
$$



The **cumulative distribution function (CDF)**, $F(x)$, gives the accumulated area to the left of a cutoff:

$$
F(x)=P(X\leq x)=\int_{-\infty}^{x}f(t)\,dt.
$$

```{r}
#| label: fig-height-cdf
#| fig-cap: "Cumulative distribution function for the height of one randomly selected student."


x <- seq(145, 195, length.out = 500)

plot(
  x,
  pnorm(x, mean = 170, sd = 7),
  type = "l",
  lwd = 2,
  xlab = "Height (cm)",
  ylab = "Cumulative probability",
  main = ""
)
```

Since total area under the density curve is 1, the probability a student being above 180 cm tall is the complement of the probability at or below 180 cm:

$$
P(X>180)=1-F(180).
$$


The normal CDF cannot be calculated using a simple closed-form expression. Historically, probabilities were obtained from tabulated values of the **standard normal distribution** - a normal distribution with mean 0 and SD 1. Today, we can calculate them numerically.

E.g., in R `pnorm()` evaluates the normal CDF directly, and for our model with mean 170 cm and SD 7 cm:

```{r}
#| code-fold: show
pnorm(180, mean = 170, sd = 7)
```

This gives $F(180)\approx0.9234$. Therefore:

$$
P(X>180)=1-F(180)\approx1-0.9234=0.0766.
$$
So, under this model, about **7.7% of students** are taller than 180 cm.

In R, we can also obtain the upper-tail probability directly by setting `lower.tail = FALSE` and arrive at the same results:

```{r}
#| code-fold: show
pnorm(180, mean = 170, sd = 7, lower.tail = FALSE)
```

## Standardisation: how far from the mean?

A **z-score** expresses distance from the mean in SD units:

$$
Z=\frac{X-\mu}{\sigma}.
$$

- Subtracting the mean centres the distribution at zero.
- Dividing by the SD rescales it to have a standard deviation of one.

For a normally distributed variable, standardisation gives a a new variable $Z$:

$$
Z \sim N(0,1),
$$

that follows the **standard normal distribution**.

Given our height example, we can locate where a measurement e.g. 180 cm lies relative to the mean:

$$
z=\frac{180-170}{7}\approx1.43.
$$

This height is about **1.43 SD above the mean**.

Standardisation changes the scale, but not the corresponding probability. Being above 180 cm is equivalent to having a z-score above $10/7 \approx 1.43$.

Writing $\Phi$ for the standard normal cumulative distribution function:

$$
\begin{aligned}
P(X>180)
&=P\left(Z>\frac{180-170}{7}\right)\\
&=1-\Phi\left(\frac{10}{7}\right)\\
&\approx0.0766.
\end{aligned}
$$

We get the same probability in R using the standard normal distribution, which is the default for `pnorm()`:

```{r}
#| code-fold: show

z <- (180 - 170) / 7

pnorm(z, lower.tail = FALSE)
```

## Percentiles: finding a cutoff

So far, we have started with a height and calculated a probability.
Now we reverse the question:

**Below which height do 95% of students fall?**

The **95th percentile**, $x_{0.95}$, satisfies:

$$
P(X\leq x_{0.95})=0.95.
$$

We first find this cutoff on the standard normal scale.

The symbol $\Phi$ denotes the **standard normal cumulative distribution function**:

$$
\Phi(z)=P(Z\leq z).
$$

It gives the probability below a specified z-score.
Its inverse, $\Phi^{-1}$, goes in the opposite direction: it gives the z-score for a specified cumulative probability.

For a probability of 0.95:

$$
z_{0.95}=\Phi^{-1}(0.95)\approx1.645.
$$

This value comes from the standard normal distribution: approximately 95% of its area lies below 1.645.
We obtain it using software or a standard normal table.

In R:

```{r}
#|code-fold: show
qnorm(0.95)
```

To convert this z-score back to centimeters, we multiply by the SD and add the mean:

$$
\begin{aligned}
x_{0.95}
&=\mu+\sigma z_{0.95}\\
&\approx170+7\times1.645\\
&\approx181.5\text{ cm}.
\end{aligned}
$$

We can also obtain the height cutoff directly in R:

```{r}
qnorm(0.95, mean = 170, sd = 7)
```

Under our model, 95% of students are at or below approximately 181.5 cm, and 5% are above it.

A percentile converts a probability into a cutoff. In R:

- `pnorm()` goes from a value to a cumulative probability;
- `qnorm()` goes from a cumulative probability to a value.


## Discrete outcomes: counting students

Not all outcomes are continuous. In life sciences, we may instead be interested in counts, such as the number of bacterial colonies on a plate or the number of sequencing reads assigned to a gene. The same probability concepts apply, but discrete outcomes are described using probabilities at specific values rather than areas under a continuous curve.

Let's imagine we are interested in how many students in a sample are taller than 180 cm. We sample 20 PhD students from a course at UU and record how many of them are taller than 180 cm.

Let $K$ denote this count. Its possible values are:

$$
K\in\{0,1,2,\ldots,20\}.
$$

Unlike height, this is a **discrete random variable**: it can take only specific, whole-number values, from 0 to 20.

## From one student to a count

For one student, the outcome “taller than 180 cm or not” can be coded as 1 or 0, respectively.

This follows a **Bernoulli distribution**, which describes a single random outcome with two possible values:

- $1$: success
- $0$: failure

If the probability of success is $p$, then the probability of failure is $1-p$:

$$
P(Y=y)=
\begin{cases}
p, & y=1,\\
1-p, & y=0.
\end{cases}
$$

The expected value and variance are:

$$
E[Y]=p,
\qquad
\operatorname{Var}(Y)=p(1-p).
$$

In our example, for one randomly selected student:

- $Y=1$ if the student is taller than 180 cm,
- $Y=0$ otherwise.

Since $X$ denotes the student's height, under our height model:

$$
p=P(X>180)\approx0.0766.
$$

Thus:

$$
Y \sim \operatorname{Bernoulli}(0.0766).
$$

So the Bernoulli distribution describes whether **one student** is taller than 180 cm or not.

A **binomial distribution** describes the number of successes in $n$ independent Bernoulli trials, where each trial has the same probability of success, $p$.

If $Y_i$ denotes the Bernoulli outcome for student $i$, then the total number of successes is:

$$
K=\sum_{i=1}^{n}Y_i.
$$

The binomial distribution is written as:

$$
K\sim\operatorname{Bin}(n,p).
$$

The expected value and variance are:

$$
E[K]=np,
$$

$$
\operatorname{Var}(K)=np(1-p).
$$

In our example, we sample 20 students and define a success as being taller than 180 cm. Since

$$
p=P(X>180)\approx0.0766,
$$

the number of students taller than 180 cm follows:

$$
K\sim\operatorname{Bin}(20,0.0766).
$$

The expected count is therefore:

$$
E[K]=20\times0.0766\approx1.53.
$$

This means that across many repeated samples of 20 students, the average number taller than 180 cm would be about **1.5 students**. Each individual observed count, however, must be a whole number.

## Probability mass function: the probability of each count


For a discrete random variable, the **probability mass function (PMF)** gives the probability of each possible value:

$$
p_K(k)=P(K=k).
$$

Unlike a continuous density, where probability is represented by an area, the value of a PMF is itself a probability.

For a binomial random variable,

$$
K\sim\operatorname{Bin}(n,p),
$$

the probability mass function is:

$$
P(K=k)=
\binom{n}{k}
p^k(1-p)^{n-k},
\qquad k=0,1,\ldots,n.
$$

Here, $\binom{n}{k}$ counts the number of ways in which exactly $k$ successes can occur among $n$ trials.

In our example,

$$
K\sim\operatorname{Bin}(20,0.0766),
$$

where $K$ is the number of students taller than 180 cm among 20 independently selected students.

The PMF therefore gives the probability of observing each possible count, from 0 to 20 students.

We can calculate and plot these probabilities in R using `dbinom()`:

```{r}
#| label: fig-height-count
#| echo: false
#| fig-cap: "Probability of each possible number of students taller than 180 cm among 20 selected students."
#| fig-height: 4

p_tall <- pnorm(
  180, mean = 170, sd = 7,
  lower.tail = FALSE
)

k <- 0:20

plot(
  k,
  dbinom(k, size = 20, prob = p_tall),
  type = "h",
  lwd = 8,
  col = "#287C8E",
  xlab = "Students taller than 180 cm (out of 20)",
  ylab = "Probability"
)

```


The height of each bar is the probability of observing that particular count. For example, the bar at $k=2$ represents:

$$
P(K=2).
$$

The probabilities across all possible counts sum to 1:

$$
\sum_{k=0}^{20} P(K=k)=1.
$$
As an example, let's calculate the probability that exactly five students out of 20 are taller than 180cm. 
To do that, we evaluate the PMF at $k=5$:

$$
P(K=5)=\binom{20}{5}p^5(1-p)^{15}.
$$

Or in R:

```{r}
#| code-fold: show
dbinom(5, size = 20, prob = p_tall)
```

**Note** that for a continuous variable, a PDF gives a density and probabilities are obtained from areas. For a discrete variable, a PMF gives probabilities directly at individual values.


### Cumulative distribution function: adding probabilities 

The **cumulative distribution function** (CDF) gives the probability of a value at or below a cutoff (same meaning as for continuous distribution).

For our count variable (K):

$$
F_K(k)=P(K\leq k)=\sum_{j=0}^{k}P(K=j).
$$

For example, the probability of four or fewer students being taller than 180 cm is:

$$
F_K(4)=P(K=0)+P(K=1)+\cdots+P(K=4).
$$

For continuous variables, we integrate the density. For discrete variables, we add the probability masses.

In R, `pbinom()` evaluates the binomial CDF:

```{r}
pbinom(4, size = 20, prob = p_tall)
```


What is the probability that five or more students are of 20 will be taller than 180cm? 

As before, we can use the complement rule and this time write:

$$
P(K\geq5)=1-P(K\leq4)=1-F_K(4).
$$

In R:

```{r}
#| code-fold: show
1 - pbinom(
  4, size = 20, prob = p_tall,
  lower.tail = TRUE
)

#or 

pbinom(
  4, size = 20, prob = p_tall,
  lower.tail = FALSE
)
```


## Continuous vs. discrete

| Concept | Continuous example: height | Discrete example: count |
|---|---|---|
| Random variable | $X$: one student's height | $K$: number taller than 180 cm among 20 students |
| Model | Normal | Binomial |
| Distribution function | PDF: $f(x)$ gives density | PMF: $p_K(k)=P(K=k)$ gives probability |
| Probability calculation | Integrate density over an interval | Sum probabilities of individual values |
| CDF | $F_X(x)=P(X\leq x)$ | $F_K(k)=P(K\leq k)$ |
| R: density or mass | `dnorm()` gives density | `dbinom()` gives probability |
| R: cumulative probability | `pnorm()` | `pbinom()` |
| Upper tail | $P(X>180)=1-F_X(180)$ | $P(K\geq5)=1-F_K(4)$ |

**The CDF has the same meaning in both cases: probability at or
below a cutoff. The difference is how we accumulate probability:
integration for a continuous density, summation for discrete masses.**

## Other common distributions in life sciences

Beyond normal and binomial distributions, several other
models are useful:

- **Poisson:** counts within a fixed time, area or volume,
  such as colony counts under suitable conditions.
  The mean and variance are equal.

- **Negative binomial:** counts with more variability than
  Poisson allows. Commonly used for RNA-seq counts across
  biological replicates.

- **Log-normal:** positive measurements whose logarithms
  are normally distributed, such as some biological concentrations.

- **Exponential:** waiting times between events occurring
  independently at a constant rate.

- **Hypergeometric:** counts when sampling without replacement
  from a finite set, such as overlap between a selected gene
  list and a pathway.

The **t**, **chi-squared** and **F** distributions also appear
in statistical inference, often as distributions of test statistics.

## Simulation: generating outcomes from a probability model

Once we have specified a probability model, we can use it to
**simulate possible outcomes**.

For example, under our height model:

$$
X \sim N(170, 7^2),
$$

we can simulate the height of one randomly selected student:

```{r}
#| echo: true
#| code-fold: false

rnorm(1, mean = 170, sd = 7)

```

Each simulation gives one possible outcome under the model.

We can also simulate many students:

```{r}
#| echo: true
#| code-fold: false
x <- rnorm(1000, mean = 170, sd = 7)
head(x)
```

The individual simulated values vary, but with many simulations
their distribution resembles the probability distribution specified
by the model.

```{r}
#| echo: true
#| code-fold: false
hist(x, n = 100)
```


## Approximating probabilities by simulation

Simulation can also be used to approximate probabilities.

For example, we previously calculated:

$$ P(X>180) \approx 0.077. $$

Using simulated heights, we can estimate the same probability by
calculating the proportion that exceed 180 cm:

```{r}
#| echo: true
#| code-fold: false

mean(x > 180)

```

With a large number of simulations, this value should be close to
the probability calculated from the normal distribution.

**The law of large numbers** states that if the same experiment is performed many times the average of the result will be close to the expected value.

## Key message

::: {.callout-important}
## Key message

A probability model lets us move from describing possible
outcomes to calculating how likely they are, under stated assumptions.

For continuous measurements, probabilities are areas under
a density curve; for discrete counts, they are sums of
probability masses. 

In both cases, the CDF gives the probability
at or below a cutoff, while a percentile answers the reverse
question: which cutoff corresponds to a given probability?

Simulation generates possible outcomes under an assumed probability model.

By repeating the simulation many times, we can explore the variation
and probabilities implied by that model.
:::