---
title: "Repeating the sampling"
subtitle: "Statistical Methods for Life Sciences"
format:
  html:
    toc: true
    number-sections: true
---

## From individual observations to study results

So far, we have used probability models to describe the height
of one randomly selected student.

In a study, however, we usually observe several individuals
and calculate a summary, such as the sample mean.

**If we repeated the study with different students, would we
obtain the same sample mean?**

A **parameter** describes a population. For example, the population
mean height is denoted by $\mu$.

A **statistic** is calculated from a sample. For example, the
sample mean is

$$
\bar{X}=\frac{1}{n}\sum_{i=1}^{n}X_i.
$$

Before collecting the data, $\bar{X}$ is a random variable because
its value depends on which individuals are selected.

After collecting the data, we obtain one observed value,
denoted by $\bar{x}$.

## Imagine repeating the study

Consider our hypothetical population of student heights, with

$$
\mu=170\text{ cm}
\qquad\text{and}\qquad
\sigma=7\text{ cm}.
$$

Imagine repeating the following study many times:

1. Randomly select five students.
2. Measure their heights.
3. Calculate their mean height.
4. Record the mean and select a new sample.

Each sample is likely to contain different students, so the sample
means will not all be identical.

Provided that the observations are sampled independently from the
same population, the expected value of the sample mean is the
population mean:

$$
E[\bar{X}]=\mu=170\text{ cm}.
$$

The distribution of the sample means across repeated studies is
called the **sampling distribution of the sample mean**.

::: {.callout-note}
The population distribution describes variation among individuals.

The sampling distribution describes variation among statistics calculated from different samples.
:::

## Simulating repeated studies

We can explore the sampling distribution by simulating many
independent studies from our normal population model.

**Before looking at the results, predict:**

- Will sample means vary as much as individual heights?
- What will happen if each study includes 100 students rather
  than five?

```{r}
#| label: fig-sampling-means
#| echo: false
#| fig-cap: "Individual heights and sample means are shown on the same horizontal scale. Larger samples produce less variable sample means."
#| fig-height: 4

set.seed(2026)

mu <- 170
sigma <- 7

individuals <- rnorm(
  10000,
  mean = mu,
  sd = sigma
)

means5 <- replicate(
  10000,
  mean(rnorm(5, mean = mu, sd = sigma))
)

means100 <- replicate(
  10000,
  mean(rnorm(100, mean = mu, sd = sigma))
)

local({
  old_par <- par(mfrow = c(1, 3))
  on.exit(par(old_par))

  values <- list(individuals, means5, means100)
  titles <- c(
    "Individual heights",
    "Sample means: n = 5",
    "Sample means: n = 100"
  )

  for (i in seq_along(values)) {
    xlabel <- if (i == 1) "Height (cm)" else "Mean height (cm)"

    hist(
      values[[i]],
      breaks = 30,
      xlim = c(140, 200),
      main = titles[i],
      xlab = xlabel,
      ylab = "Frequency",
      col = "#8FC4CF",
      border = "white"
    )

    abline(v = mu, lty = 2)
  }
})
```

All three distributions are centred near the population mean
of 170 cm.

However:

- Individual heights show substantial variation.
- Means based on five students vary less than individual heights.
- Means based on 100 students vary much less.

Thus, **sample means vary less than individual observations, and
means based on larger samples vary less than means based on smaller
samples**.

This principle is not limited to means. Other statistics, such as
a sample proportion or a difference between two means, also have
sampling distributions.

## Why do the sample means look normal?

In this example, individual heights follow a normal distribution.
The mean of independent observations from a normal population is
therefore also normally distributed.

The **central limit theorem** extends this result beyond normal
populations:

> For many population distributions, the sampling distribution of
> the sample mean becomes approximately normal as the sample size
> increases.

This does not mean that the individual observations become normally
distributed. It is the **distribution of the sample mean across
repeated samples** that becomes approximately normal.

How large the sample must be for a useful approximation depends on
the shape of the population distribution and the presence of
extreme values.

## What if we have only one study? {.optional}

The previous simulation assumes that we know the population model
and can generate many new samples from it.

In practice, we usually have only one observed sample.

For example, if we measure 20 students, we have one sample containing
20 observations—not 20 independent samples.

Can we use this sample to approximate how the sample mean might vary
across repeated studies?

### Bootstrap: resampling the observed data

The **bootstrap** treats the observed sample as an approximation
of the population and repeatedly resamples from it.

Imagine writing the 20 observed heights on cards:

1. Draw one card at random, record its value and return it.
2. Repeat until 20 values have been drawn.
3. Calculate their mean.
4. Repeat the procedure many times.

The cards are returned after every draw, so sampling is performed
**with replacement**. An observed height can therefore appear more
than once in a bootstrap sample, while another may not appear at all.

Sampling without replacement would not work: selecting all 20 cards
would reproduce the original sample and therefore always give the
same mean.

```{r}
#| label: fig-bootstrap-means
#| echo: false
#| fig-cap: "Bootstrap distribution of the mean based on one illustrative sample of 20 students."
#| fig-height: 4

set.seed(2027)

heights <- rnorm(20, mean = 170, sd = 7)

bootstrap_means <- replicate(
  10000,
  mean(
    sample(
      heights,
      size = length(heights),
      replace = TRUE
    )
  )
)

hist(
  bootstrap_means,
  breaks = 30,
  main = "Bootstrap distribution of the mean",
  xlab = "Mean height (cm)",
  ylab = "Frequency",
  col = "#8FC4CF",
  border = "white"
)

abline(
  v = mean(heights),
  lty = 2,
  lwd = 2
)
```

The bootstrap means are centred near the **observed sample mean**,
not necessarily the true population mean.

The spread of the bootstrap distribution approximates how much the
sample mean would vary across repeated samples.

Bootstrapping does not create new biological information. Its
results still depend on whether the original sample adequately
represents the population.

This simple bootstrap also assumes independent observations. It
cannot correct selection bias or compensate for a poorly designed
study.

::: {.callout-important}
## Key message

A statistic, such as a sample mean, varies from sample to sample.

The distribution of its possible values across repeated samples
is called its **sampling distribution**.

Sample means vary less than individual observations, and means
based on larger samples generally vary less.
:::

Next, we need a way to quantify the spread of the sampling
distribution. This leads to the **standard error**.