---
title: "Describing precision"
subtitle: "Statistical Methods for Life Sciences"
format:
  html:
    toc: true
    number-sections: true
---

## How variable is a sample mean?

In the previous section, we saw that sample means vary across
repeated samples.

The **standard error (SE)** quantifies this sampling variation.

It is important to distinguish between the standard deviation
and the standard error:

- **Standard deviation (SD):** describes variation among individual
  observations.
- **Standard error (SE):** describes variation among estimates
  calculated from repeated samples of the same size.

For the sample mean, the SE is the standard deviation of its
sampling distribution.

A smaller SE means that repeated samples tend to produce more
similar estimates. We therefore say that the estimate is
**more precise**.

::: {.callout-note}
## SD and SE answer different questions

**SD:** How different are individuals from one another?

**SE:** How much would our estimate vary if we repeated the study?
:::

## Standard error of the sample mean

For $n$ independent observations from a population with standard
deviation $\sigma$, the standard error of the sample mean is

$$
\operatorname{SE}(\bar{X})
=
\frac{\sigma}{\sqrt{n}}.
$$

The formula shows that the sampling variability of the mean
depends on:

- the variability among individuals, represented by $\sigma$;
- the number of independent observations, represented by $n$.

Greater variation among individuals produces a larger SE.
A larger sample size produces a smaller SE.

For our height model, the population SD is 7 cm:

$$
\begin{aligned}
n=5:
&\quad
\operatorname{SE}(\bar{X})
=
\frac{7}{\sqrt{5}}
\approx 3.13\text{ cm},\\[4pt]
n=100:
&\quad
\operatorname{SE}(\bar{X})
=
\frac{7}{\sqrt{100}}
=
0.70\text{ cm}.
\end{aligned}
$$

This explains the sampling distributions from the previous
section: means based on 100 students were much more tightly
concentrated around 170 cm than means based on five students.

Increasing the sample size improves the **precision of the mean**.
It does not reduce the biological variation among individual
students.

## How does sample size affect precision?

Because

$$
\operatorname{SE}(\bar{X})=\frac{\sigma}{\sqrt{n}},
$$

the SE decreases with the square root of the sample size.

This means that doubling the sample size does not halve the SE.

To halve the SE, we must multiply the sample size by four:

$$
\frac{\sigma}{\sqrt{4n}}
=
\frac{1}{2}\frac{\sigma}{\sqrt{n}}.
$$

Increasing the sample size therefore improves precision, but with
diminishing returns.


## Estimating the standard error from data

The formula

$$
\operatorname{SE}(\bar{X})
=
\frac{\sigma}{\sqrt{n}}
$$

uses the **population standard deviation**, $\sigma$.

In practice, $\sigma$ is usually unknown. We estimate it from the
observed sample using the **sample standard deviation**, $s$:

$$
s
=
\sqrt{
\frac{1}{n-1}
\sum_{i=1}^{n}
(x_i-\bar{x})^2
}.
$$

The sample SD describes the variation among the observed individuals
and provides an estimate of the population SD, $\sigma$.

We can therefore estimate the standard error of the sample mean by

$$
\widehat{\operatorname{SE}}(\bar{X})
=
\frac{s}{\sqrt{n}}.
$$
. . .

<br>

For example, suppose we measure the heights of 20 students:

```{r}
#| echo: true
#| code-fold: false

set.seed(2027)

heights <- rnorm(
  20,
  mean = 170,
  sd = 7
)

# calculate sample SD
s <- sd(heights)
s

# estimate the SE of the sample mean
n <- length(heights)
SE <- s / sqrt(n)
SE
```

This estimated SE describes how much the sample mean would be expected
to vary across repeated samples of 20 independent students.


### Estimating SE using the bootstrap {.optional}

If we generated bootstrap means in the previous section, we can
also estimate the SE by calculating their standard deviation:

```{r}
#| echo: true
#| code-fold: false

bootstrap_means <- replicate(
  10000,
  mean(
    sample(
      heights,
      size = length(heights),
      replace = TRUE
    )
  )
)

sd(bootstrap_means)
```

The formula-based and bootstrap estimates both describe the
estimated sampling variability of the mean. Their values will
usually be similar, although not necessarily identical.

## From the normal to the t-distribution {.smaller}

If the population SD, $\sigma$, is known, standardising the sample mean gives

$$
Z=
\frac{\bar X-\mu}
{\sigma/\sqrt n}
\sim N(0,1).
$$

If $\sigma$ is unknown, which is the usual real-data situation, we replace it with the sample SD, $s$:

$$
T=
\frac{\bar X-\mu}
{s/\sqrt n}.
$$

Now the denominator is itself estimated from the sample, which introduces extra uncertainty.

Therefore, under a normal population model,

$$
T \sim t_{n-1}.
$$

The **t-distribution** is similar to the standard normal distribution but has heavier tails. As the sample size increases, it approaches $N(0,1)$.


Because $s$ is estimated from the sample, there is **extra uncertainty**.

The **t-distribution** accounts for this uncertainty:

- it looks similar to the normal distribution;
- it has heavier tails;
- with larger samples, it approaches the standard normal distribution.

```{r}
#| echo: false


x <- seq(-4, 4, length.out = 500)

par(mar = c(5, 5, 2, 2) + 0.1)

plot(
  x,
  dnorm(x),
  type = "l",
  lwd = 2,
  xlab = "Standardised value",
  ylab = "Density",
  ylim = c(0, 0.42)#,
  #cex.lab = 1.5,
  #cex.axis = 1.3
)

lines(
  x,
  dt(x, df = 4),
  lwd = 2,
  lty = 2, 
  col = "red"
)

lines(
  x,
  dt(x, df = 20),
  lwd = 2,
  lty = 3, 
  col = "blue"
)

legend(
  "topright",
  legend = c("Normal", "t, df = 4", "t, df = 20"),
  lwd = 4,
  lty = c(1, 2, 3),
  bty = "n",
  cex = 1
)

```

## More measurements or more individuals?

Suppose that we want to estimate the mean height in the population.

We could collect 100 measurements in two different ways:

- measure five students 20 times each;
- measure 100 different students once each.

**Would these designs give equally precise estimates of the
population mean?**

::: {.callout-tip title="Solution" collapse="true"}

No. Although both designs produce 100 measurements, they do not
contain the same amount of independent information about variation
between students.

In

$$
\widehat{\operatorname{SE}}(\bar{X})
=
\frac{s}{\sqrt{n}},
$$

$n$ is the number of **independent students**, not simply the total
number of measurements.

If the between-student SD is approximately 7 cm:

$$
\begin{aligned}
\text{Five students:}
&\quad
\operatorname{SE}(\bar{X})
\approx
\frac{7}{\sqrt{5}}
=
3.13\text{ cm},\\[4pt]
\text{One hundred students:}
&\quad
\operatorname{SE}(\bar{X})
\approx
\frac{7}{\sqrt{100}}
=
0.70\text{ cm}.
\end{aligned}
$$

Repeated measurements can reduce measurement error and give a more
precise measurement of each student.

However, repeatedly measuring the same five students does not give
us 100 independent students. Those five students could still happen
to be unusually tall or unusually short.

Therefore, measuring 100 different students generally gives a more
precise estimate of the **population mean**, assuming that each
height is measured reasonably accurately.

In biological studies, this is why technical replicates cannot
replace independent biological replicates.

:::

## Using SE to assess an observed result

Previously, we used a probability distribution to ask whether an
individual height was unusual.

We can now ask a different question:

> Is an observed **sample mean** unusual relative to its expected
> sampling variability?

Assume that student heights follow our normal population model:

$$
X\sim N(170,7^2).
$$

For a sample of 100 independent students, the sample mean follows

$$
\bar{X}
\sim
N\left(170,\frac{7^2}{100}\right).
$$

The sampling distribution therefore has:

$$
E[\bar{X}]=170
$$

and

$$
\operatorname{SE}(\bar{X})
=
\frac{7}{\sqrt{100}}
=
0.70\text{ cm}.
$$

Suppose that we observe a sample mean of 172 cm.

Its distance from the assumed population mean, measured in standard
error units, is

$$
z
=
\frac{\bar{x}-\mu_0}
{\sigma/\sqrt{n}}
=
\frac{172-170}{0.70}
\approx 2.86.
$$

Thus, the observed sample mean is approximately 2.86 SE above the
assumed population mean.

Under the assumed model,

$$
P(\bar{X}\geq172)
=
1-\Phi\left(
\frac{172-170}{0.70}
\right)
\approx 0.0021.
$$

In R:

```{r}
#| echo: true
#| code-fold: false

pnorm(
  172,
  mean = 170,
  sd = 7 / sqrt(100),
  lower.tail = FALSE
)
```

If the population mean is 170 cm and the model assumptions hold,
the probability of obtaining a sample mean of 172 cm or higher
from 100 independent students is approximately 0.2%.

Such a sample mean would therefore be unusual under the assumed
model.

::: {.callout-warning}
This is the probability of obtaining a result at least this large
**under the assumed model**.

It is not the probability that the model or the assumed population
mean is correct.
:::

## An individual value is not a sample mean

An individual height of 172 cm is not unusual under the same
population model.

For an individual student, we use the individual SD of 7 cm:

$$
P(X\geq172)
=
1-\Phi\left(
\frac{172-170}{7}
\right)
\approx0.3875.
$$

```{r}
#| echo: true
#| code-fold: false

pnorm(
  172,
  mean = 170,
  sd = 7,
  lower.tail = FALSE
)
```

Approximately 38.8% of individual students are expected to be at
least 172 cm tall under this model.

The cutoff of 172 cm is the same in both calculations, but the
relevant distributions are different:

- Individual heights have an SD of 7 cm.
- Means of 100 heights have an SE of 0.70 cm.

A value can therefore be common for an individual observation but
unusual for a sample mean.

## Return to the two research groups

In our opening example, two research groups obtained mean heights of
172 cm and 175 cm.

Their observed difference was therefore

$$
\bar{x}_2-\bar{x}_1
=
175-172
=
3\text{ cm}.
$$

We now know that **different random samples can produce different
sample means, even when they come from the same population**.

So an observed difference of 3 cm may simply reflect **sampling
variation**.


The important question is:

> **Is a difference of 3 cm larger than we would expect from sampling
> variation alone?**

To answer this, we would consider the statistic

$$
\bar{X}_2-\bar{X}_1
$$

and its **sampling distribution**: the distribution of differences we
would obtain if we repeatedly drew two independent samples from the
same population under the same conditions.

How much these differences vary depends on, among other things:

- the variation among individuals;
- the number of independent individuals in each sample;
- the sampling design and model assumptions.

We can then compare our **observed difference** with the variation
expected under a specified model.

This is us from **probability** to **statistical inference**.

In the next session, we will use **confidence intervals and hypothesis
tests** to quantify uncertainty and assess evidence about population
parameters.

::: {.callout-important}
## Key message

Different samples produce different results.

The sampling distribution describes how a statistic varies across repeated samples, and the standard error quantifies this variation.

Larger independent samples generally give more precise estimates and smaller standard errors.

Statistical inference uses the observed data and a probability model to assess what the data tell us about the population.
:::