---
title: "Session example"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path="session-example-files/figures/")
```

### Learning outcomes
- outcome 1
- outcome 2
- outcome 3

## Content
- [Including Code](#part1)
- [Including Plots](#part2)
- [Including Links](#part3)


## GitHub Documents 

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code <a name="part1"></a>

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots <a name="part1"></a>

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Including Links <a name="part3"></a>
More [R markdown examples][rmd-examples]



[rmd-examples]: https://rmarkdown.rstudio.com/authoring_basics.html


